### 2024-10-10 paper.qmd
paper

---
title: "Incorporating spatial distribution at age improves predictive ability of SDMs"
format: pdf
editor: visual
---

## Abstract

## Introduction

Why hake and sablefish? 2 species with strong gradients in distribution at age

## Methods

### Data

On the west coast of the USA, the West Coast Groundfish Bottom Trawl Survey (WCGBTS) has been an annual survey conduced from 2003 - present. The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA (Keller et al., 2017). The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data). Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyse. Like many other surveys around the world, no WCGBTS survey occured in 2020 because of the Covid-19 pandemic. Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish. \[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 – 10 years. Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al. 2024) and sablefish is frequently assessed because of its high commercial value. For our analysis, we used hake collected 2007 – 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 – 2023 (mean 1314.4 individuals sampled per year).

### Spatial age models

For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 – 5, and for sablefish ages 0 – 6 (56% and 63% of the total distribution of ages, respectively). For each species - age combination, we constructed a unique spatiotemporal model fit to all years except the last available year for each species. Individual fish were aggregated at the haul level, summarizing the total number of fish sampled and number of fish sampled of the focal age; this allowed counts to be modeled as a binomial family (with logit link). Second, we constructed a spatiotemporal model as an extension of Generalized Linear Mixed Models (GLMMs) such that the prediction in location $s$ and time $t$ can be written as

$$
\begin{aligned}
logit \left( p_{\boldsymbol{s},t} \right) &= \boldsymbol{\beta_{t}} + \omega_{\boldsymbol{s}} + \delta_{\boldsymbol{s},t}
\end{aligned}
$$

where $\boldsymbol{\beta_{t}}$ represent time - varying intercepts modeled as a random walk $\boldsymbol{\beta_{t}} \sim N \left( \boldsymbol{\beta_{t-1}}, \sigma_\beta \right)$, the spatial field $\boldsymbol{\omega_{s}} \sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right)$ and the spatiotemporal fields $\delta_{\boldsymbol{s},t}$ are modeled as an AR(1) process $\boldsymbol{\delta}_{t} = \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}}$, where $\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right)$

Estimation was performed via sdmTMB (Anderson et al) and using TMB / INLA SPDE etc...

### Validating future predictive ability

As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution of ages. For each of the species - age models constructed above, we made predictions to the survey locations in the following year (e.g. age 3 hake in years 2007 -- 2018 used to predict the distribution of age 4 hake in 2008 -- 2019). We related the predicted probabilities of occurrence $\hat{p}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{p}_{\boldsymbol{s},t}$. We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$. The exponentiated slope parameter $exp(\beta_{1})$ represents a the change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival.

As a second validation, we compared aggregated predictions across all ages to a SDM model that used total biomass as a response.

### Links to fisheries data?

## Results

## Discussion

#### Limitations
- not predicting abundance (N) -- just proportions
- We constructed single-age models, and this could be done in a multivariate framework -- but our approach allows for greater flexibility in letting relationships change by age

\break

```{r label="hake-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for Pacific hake ages 1 -- 4.", echo=FALSE}
knitr::include_graphics("plots/hake_spatial_anomalies.png")
```

\break

```{r label="sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for sablefish ages 0 -- 5.", echo=FALSE}
knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
```

\break

```{r label="glm-coefficients", fig.cap="Estimated coefficients relating predicted densities of age a fish in year t to observed numbers the following year. The coefficients are interpreted with a multiplicate effect, such that a coefficient of 0.5 translates to an effect of exp(0.5) = 65% increase on the expected count in the future.", echo=FALSE}
knitr::include_graphics("plots/glm_coefficients.png")
```

### 2024-10-10 06_sablefish_age5.R
Create Pacific hake_1_surveygrid.rds

@@ -1,0 +1,62 @@
+library(tidyverse)
+library(ggplot2)
+library(DHARMa)
+library(viridis)
+
+spp <- 2
+  spp_name <- c("Pacific hake", "sablefish")[spp]
+  
+  if(spp_name == "Pacific hake") {
+    min_age <- 1 # not many age 0s consistently sampled
+    max_age <- 5
+  } 
+  if(spp_name == "sablefish") {
+    min_age <- 0
+    max_age <- 6
+  }
+  
+  for(a in min_age:(max_age - 1)) {
+    d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+    d$species <- spp_name
+    d$age <- a
+    if(a == min_age) {
+      all_output <- d
+    } else {
+      all_output <- rbind(all_output, d)
+    }
+  }
+  
+sablefish <- dplyr::filter(all_output, species == "sablefish", 
+                           year %in% seq(2015,2023,by=2), age == 5) |>
+  dplyr::mutate(p = plogis(est))
+
+
+map_data <- rnaturalearth::ne_countries(
+  scale = "medium",
+  returnclass = "sf", country = "united states of america")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+coast <- suppressWarnings(suppressMessages(
+  sf::st_crop(map_data,
+              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
+
+utm_zone10 <- 3157
+coast_proj <- sf::st_transform(coast, crs = utm_zone10)
+
+# Define the new colors
+land_color <- "#E0CDA9"  # beige for land
+ocean_color <- "#D3EAF2"  # grayish blue for ocean
+
+# Plot coast with custom land and ocean colors
+ggplot(coast_proj) + 
+  geom_sf(fill = land_color) +  # Set land color
+  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p)) + 
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Occurrence") +  # legend title 
+  theme_light() + 
+  labs(x = "Longitude", y = "Latitude") +
+  theme(panel.background = element_rect(fill = ocean_color),  
+        strip.text = element_text(color = "black"),  # Set facet label text color to black
+        strip.background = element_rect(fill = "white")) +  # Set facet label background to white
[…]

### 2024-10-10 02_recursive_model.R
update -- more ages

@@ -5,17 +5,17 @@ library(ggplot2)
 library(DHARMa)
 library(surveyjoin)
 
-spp_name <- c("Pacific hake", "sablefish")[1]
+spp_name <- c("Pacific hake", "sablefish")[2]
 
 if(spp_name == "Pacific hake") {
   min_age <- 1 # not many age 0s consistently sampled
-  max_age <- 5
+  max_age <- 6
   min_year <- 2007
   max_year <- 2019
 } 
 if(spp_name == "sablefish") {
   min_age <- 0
-  max_age <- 6
+  max_age <- 10
   min_year <- 2003
   max_year <- 2023
 }

### 2024-10-10 03_poisson_regression.R

@@ -8,11 +8,11 @@ for(spp in 1:2) {
   
   if(spp_name == "Pacific hake") {
     min_age <- 1 # not many age 0s consistently sampled
-    max_age <- 5
+    max_age <- 6
   } 
   if(spp_name == "sablefish") {
     min_age <- 0
-    max_age <- 6
+    max_age <- 10
   }
   
   output <- data.frame(age = seq(min_age, max_age), mean = NA, lower = NA, upper = NA)
@@ -42,7 +42,7 @@ ggplot(all_output, aes(age, mean, col = species)) +
   ylab("Estimated slope") + 
   coord_flip() + theme_bw()
 
-ggsave("plots/glm_coefficients.png")
+ggsave("plots/glm_coefficients.png", width=7)
 
 
 

### 2024-10-10 04_spatial_anomalty_plots.R

@@ -8,11 +8,11 @@ for(spp in 1:2) {
   
   if(spp_name == "Pacific hake") {
     min_age <- 1 # not many age 0s consistently sampled
-    max_age <- 5
+    max_age <- 6
   } 
   if(spp_name == "sablefish") {
     min_age <- 0
-    max_age <- 6
+    max_age <- 10
   }
   
   for(a in min_age:(max_age - 1)) {
@@ -60,8 +60,8 @@ ocean_color <- "#D3EAF2"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 ggplot(coast_proj) + 
+  geom_point(data = hake, aes(x = X * 1000, y = Y * 1000, col = omega_s), size=0.02) + 
   geom_sf(fill = land_color) +  # Set land color
-  geom_point(data = hake, aes(x = X * 1000, y = Y * 1000, col = omega_s)) + 
   scale_color_gradient2(name = "Anomaly") +  # legend title 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
@@ -69,18 +69,18 @@ ggplot(coast_proj) +
         strip.text = element_text(color = "black"),  # Set facet label text color to black
         strip.background = element_rect(fill = "white")) +  # Set facet label background to white
   facet_wrap(~age, nrow = 1)
-ggsave("plots/hake_spatial_anomalies.png")  
+ggsave("plots/hake_spatial_anomalies.png", width = 7, height = 4)  
   
 
 ggplot(coast_proj) + 
+  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = omega_s), size=0.02) + 
   geom_sf(fill = land_color) +  # Set land color
-  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = omega_s)) + 
   scale_color_gradient2(name = "Anomaly") +  # legend title 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  
         strip.text = element_text(color = "black"),  # Set facet label text color to black
         strip.background = element_rect(fill = "white")) +  # Set facet label background to white
-  facet_wrap(~age, nrow = 1)
-ggsave("plots/sablefish_spatial_anomalies.png")  
+  facet_wrap(~age, nrow = 2)
+ggsave("plots/sablefish_spatial_anomalies.png", width = 7, height = 8)  
 

### 2024-10-10 05_spatial_bycatch_risk_sablefish.R

@@ -35,6 +35,10 @@ sablefish <- dplyr::filter(all_output, species == "sablefish", year %in% seq(201
   dplyr::group_by(year, X, Y) |>
   dplyr::summarise(p = sum(p)) 
 
+# also de-mean the values by year so this is relative
+sablefish <- dplyr::group_by(sablefish, year) |>
+  dplyr::mutate(centered_p = p - mean(p))
+
 map_data <- rnaturalearth::ne_countries(
   scale = "medium",
   returnclass = "sf", country = "united states of america")
@@ -53,14 +57,15 @@ ocean_color <- "#D3EAF2"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 ggplot(coast_proj) + 
+  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = centered_p), size=0.02) + 
   geom_sf(fill = land_color) +  # Set land color
-  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p)) + 
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Occurrence") +  # legend title 
+  #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Occurrence") +  # legend title 
+  scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  
         strip.text = element_text(color = "black"),  # Set facet label text color to black
         strip.background = element_rect(fill = "white")) +  # Set facet label background to white
   facet_wrap(~year, nrow = 1)
-ggsave("plots/sablefish_spatial_risk.png")  
+ggsave("plots/sablefish_spatial_risk.png", width = 7, height = 4)  
 

### 2024-10-10 06_sablefish_age5.R

@@ -8,11 +8,11 @@ spp <- 2
   
   if(spp_name == "Pacific hake") {
     min_age <- 1 # not many age 0s consistently sampled
-    max_age <- 5
+    max_age <- 6
   } 
   if(spp_name == "sablefish") {
     min_age <- 0
-    max_age <- 6
+    max_age <- 10
   }
   
   for(a in min_age:(max_age - 1)) {
@@ -30,6 +30,8 @@ sablefish <- dplyr::filter(all_output, species == "sablefish",
                            year %in% seq(2015,2023,by=2), age == 5) |>
   dplyr::mutate(p = plogis(est))
 
+sablefish <- dplyr::group_by(sablefish, year) |>
+  dplyr::mutate(centered_p = p - mean(p))
 
 map_data <- rnaturalearth::ne_countries(
   scale = "medium",
@@ -49,14 +51,15 @@ ocean_color <- "#D3EAF2"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 ggplot(coast_proj) + 
+  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
+  #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Occurrence") +  # legend title 
+  scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
   geom_sf(fill = land_color) +  # Set land color
-  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p)) + 
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Occurrence") +  # legend title 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  
         strip.text = element_text(color = "black"),  # Set facet label text color to black
         strip.background = element_rect(fill = "white")) +  # Set facet label background to white
   facet_wrap(~year, nrow = 1)
-ggsave("plots/sablefish_spatial_age5.png")  
+ggsave("plots/sablefish_spatial_age5.png", width = 7, height = 4)  
 

### 2024-10-10 paper.qmd
text

@@
 
-For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 – 5, and for sablefish ages 0 – 6 (56% and 63% of the total distribution of ages, respectively).
+For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 – 5, and for sablefish ages 0 – 6 (data from these ages represented 56% and 63% of the total aged fish, respectively).
 For each species - age combination, we constructed a unique spatiotemporal model fit to all years except the last available year for each species.
@@
 
+Figure @fig-hake-spatial-anomaly shows a concentration of age-1 (mostly immature) hake near the coast, and then a clear N/S split for age-2s and 3s with higher probabilities of occurrence in the South (on avg).
+Age-4 hake appear to have a more patchy & northern distribution
+
+Figure @sablefish-spatial-anomaly shows age 0-2 sablefish concentrated in shallow water, and more fish moving offshore by age 3.
+Several interesting patches in coastal OR have very low occurrence rates
+
+Figure @glm-coefficients shows that in general, there's more predictability (or maybe persistence for sablefish than hake).
+Using the distribution of age 1 and 2 fish to predict age 2 and 3 yields similar coefficients across species -- but for older age 4-5 sablefish, these coefficients increase a lot
+
+Figure @bycatch-risk aggregates occurrence probabilities for age 0 - 2 sablefish, potentially useful to identify hotspots
+
+Similarly Figure @age5-sablefish shows occurrence probabilities for age 5 sablefish
+
 ## Discussion
@@
 - not predicting abundance (N) -- just proportions
-- We constructed single-age models, and this could be done in a multivariate framework -- but our approach allows for greater flexibility in letting relationships change by age
+- We constructed single-age models, and this could be done in a multivariate (multinomial) framework -- but our approach allows for greater flexibility in letting relationships change by age.
+Tradeoff is that it doesn't impose sum to 1 constraint, binomials are independent
 
@@
 ```
+
+\break
+
+```{r label="bycatch-risk", fig.cap="Estimated probability of occurrence for age 0 -- 2 sablefish.", echo=FALSE}
+knitr::include_graphics("plots/sablefish_spatial_risk.png")
+```
+
+\break
+
+```{r label="age5-sablefish", fig.cap="Estimated probability of occurrence for age 5 sablefish.", echo=FALSE}
+knitr::include_graphics("plots/sablefish_spatial_age5.png")
+```

### 2024-10-10 paper.qmd
Update paper.pdf

--- 2024-10-10 paper.qmd
@@
 editor: visual
+bibliography: references.bib
 ---
@@
 
-Why hake and sablefish? 2 species with strong gradients in distribution at age
+-   Age structure is important but generally ignored in marine SDMs
+
+-   Lots of reasons why age info might improve predictions – mechanisms include ontogenetic shifts, recruitment pulses, migration, fishing selectivity removing older portions of a population, etc
+
+-   If we can make forward predictions with some skill, there's lots of wins for management. Examples include predicting areas of bycatch risk, where certain cohorts might be in the future, etc
+
+-   Need consistently surveyed populations to examine the forecast skill.
+Use WCBTS as case study, with hake and sablefish targeted as focal species.
+Why hake and sablefish?
+2 species with strong gradients in distribution at age – but in opposite directions and different life histories
+
+-
 
@@
 
-For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 – 5, and for sablefish ages 0 – 6 (56% and 63% of the total distribution of ages, respectively).
+For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 – 5, and for sablefish ages 0 – 9 (data from these ages represented 54% and 75% of the total aged fish, respectively).
 For each species - age combination, we constructed a unique spatiotemporal model fit to all years except the last available year for each species.
@@
 
-Estimation was performed via sdmTMB (Anderson et al) and using TMB / INLA SPDE etc...
+Spatial and spatiotemporal random fields were constructed as Gaussian Markov random fields (GMRFs) using the stochastic partial differential equation approach (SPDE) [@lindgren2011; @lindgren2015].
+The SPDE method models the spatial correlation between points as a Matérn covariance function with smoothness parameter $\nu = 1$.
+Spatial meshes for all ages and species were constructed with a cutoff distance of 50km (this distance controls the spacing of mesh vertices).
+Parameter estimation was done using the sdmTMB software package [@Anderson2024] with R 4.3.1 [@rcore2024].
+The sdmTMB package relies on Template Model Builder (TMB)[@kristensen2016] to quickly and efficiently maximize the marginal log likelihood using auto-differentiation and the Laplace approximation to integrate out random effects.
+Models were evaluated for convergence (positive-definite Hessian matrix, and a maximum absolute log likelihood gradient \< 0.001) and residuals diagnostics were evaluated with the DHARMa package [@dharma2022].
 
@@
 
+Figure @fig-hake-spatial-anomaly shows a concentration of age-1 (mostly immature) hake near the coast, and then a clear N/S split for age-2s and 3s with higher probabilities of occurrence in the South (on avg).
+Age-4 hake appear to have a more patchy & northern distribution
+
+Figure @fig-sablefish-spatial-anomaly shows age 0-2 sablefish concentrated in shallow water, and more fish moving offshore by age 3.
+Several interesting patches in coastal OR have very low occurrence rates
+
+Figure @fig-spatial-parameters shows that for most ages, we see a decrease in the spatial and spatiotemporal variances as well as a decrease in the spatial range.
+This means that the fields for older fish become less peaky -- and the distribution of where fish are becomes patchier ( duh!)
+
+Figure @fig-glm-coefficients shows that in general, there's more predictability (or maybe persistence for sablefish than hake).
+Using the distribution of age 1 and 2 fish to predict age 2 and 3 yields similar coefficients across species -- but for older age 4-5 sablefish, these coefficients increase a lot
+
+Figure @fig-bycatch-risk aggregates occurrence probabilities for age 0 - 2 sablefish, potentially useful to identify hotspots
+
+Similarly Figure @fig-age5-sablefish shows occurrence probabilities for age 5 sablefish
+
 ## Discussion
@@
 #### Limitations
+
 -   not predicting abundance (N) -- just proportions
-- We constructed single-age models, and this could be done in a multivariate framework -- but our approach allows for greater flexibility in letting relationships change by age
+-   We constructed single-age models, and this could be done in a multivariate (multinomial) framework -- but our approach allows for greater flexibility in letting relationships change by age.
+Tradeoff is that it doesn't impose sum to 1 constraint, binomials are independent
 
@@
 
-```{r label="hake-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for Pacific hake ages 1 -- 4.", echo=FALSE}
+```{r label="fig-hake-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for Pacific hake ages 1 -- 4.", echo=FALSE}
 knitr::include_graphics("plots/hake_spatial_anomalies.png")
@@
 
-```{r label="sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for sablefish ages 0 -- 5.", echo=FALSE}
+```{r label="fig-sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for sablefish ages 0 -- 5.", echo=FALSE}
 knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
@@
 
-```{r label="glm-coefficients", fig.cap="Estimated coefficients relating predicted densities of age a fish in year t to observed numbers the following year.
+```{r label="fig-glm-coefficients", fig.cap="Estimated coefficients relating predicted densities of age a fish in year t to observed numbers the following year.
 The coefficients are interpreted with a multiplicate effect, such that a coefficient of 0.5 translates to an effect of exp(0.5) = 65% increase on the expected count in the future.", echo=FALSE}
@@
 ```
+
+\break
+
+```{r label="fig-bycatch-risk", fig.cap="Estimated probability of occurrence for age 0 -- 2 sablefish.", echo=FALSE}
+knitr::include_graphics("plots/sablefish_spatial_risk.png")
+```
+
+\break
+
+```{r label="fig-age5-sablefish", fig.cap="Estimated probability of occurrence for age 5 sablefish.", echo=FALSE}
+knitr::include_graphics("plots/sablefish_spatial_age5.png")
+```
+
+\break
+
+```{r label="fig-spatial-parameters", fig.cap="Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations).
+Points represent mean estimates; lines represent 95% confidence intervals", echo=FALSE}
+knitr::include_graphics("plots/spatial_parameters.png")
+```
+
+\break

### 2024-10-10 02_recursive_model.R

@@ -99,6 +99,15 @@ for(a in min(subset$age):(max(subset$age) - 1)) {
   #res <- residuals(fit, type = "mle-mvn")
   #qqnorm(res);abline(0, 1)
   
+  if(a == min(subset$age)) {
+    pars <- tidy(fit, "ran_pars")
+    pars$age <- a
+    ran_pars <- pars
+  } else {
+    pars <- tidy(fit, "ran_pars")
+    pars$age <- a
+    ran_pars <- rbind(ran_pars, pars)
+  }
   # predict to locations in next age / year
   pred_df <- dplyr::filter(subset, age == (a+1), year > min(subset$year), n_total > 0)
 
@@ -128,3 +137,44 @@ for(a in min(subset$age):(max(subset$age) - 1)) {
   
 }
 
+saveRDS(ran_pars, paste0("predictions/",spp_name,"_ranpars.rds"))
+
+
+
+
+hake_pars <- readRDS(paste0("predictions/","Pacific hake","_ranpars.rds"))
+sablefish_pars <- readRDS(paste0("predictions/","sablefish","_ranpars.rds"))
+hake_pars$species <- "Pacific hake"
+sablefish_pars$species <- "sablefish"
+pars <- rbind(hake_pars, sablefish_pars)
+
+p1 <- dplyr::filter(pars, term == "range") |> 
+  ggplot(aes(age, estimate, col = species)) + 
+  geom_pointrange(aes(ymin=estimate-2*std.error, ymax = estimate+2*std.error), position = position_dodge(0.4)) + 
+  scale_color_viridis_d(option="magma", begin=0.2, end = 0.8) +
+  xlab("Age") + 
+  ylab("Estimated range (km)") + 
+  theme_bw() + 
+  theme(legend.position = "none")
+
+p2 <- dplyr::filter(pars, term == "sigma_O") |> 
+  ggplot(aes(age, estimate, col = species)) + 
+  geom_pointrange(aes(ymin=estimate-2*std.error, ymax = estimate+2*std.error), position = position_dodge(0.4)) + 
+  scale_color_viridis_d(option="magma", begin=0.2, end = 0.8) +
+  xlab("Age") + 
+  ylab(expression(paste("Spatial ", sigma))) + 
+  theme_bw() + 
+  theme(legend.position = "none")
+
+p3 <- dplyr::filter(pars, term == "sigma_E") |> 
+  ggplot(aes(age, estimate, col = species)) + 
[…]

### 2024-10-10 03_poisson_regression.R

@@ -22,7 +22,29 @@ for(spp in 1:2) {
     g <- glm(n ~ log(expected_n), data = d, family = "poisson")
     output[which(output$age==a),"mean"] <- coef(g)[2]
     output[which(output$age==a),c("lower","upper")] <- confint(g)[2,]
+    # 
+    # d$age <- a
+    #d$resid <- residuals(g)
+    # d$resid <- qnorm(simulateResiduals(fittedModel = g)$scaledResiduals)
+    # if(a == min_age) {
+    #   all_d <- d
+    # } else {
+    #   all_d <- rbind(all_d, d)
+    # }
   }
+  # all_d$species <- spp_name
+  # ggplot(data = dplyr::filter(all_d_spp, species=="sablefish"), aes(sample = resid)) +
+  #   stat_qq() +
+  #   stat_qq_line() +  # Add a reference line
+  #   labs(title = "Q-Q Plot of GLM Residuals", x = "Theoretical Quantiles", y = "Deviance Residuals") +
+  #   theme_minimal()
+  # 
+  
+  # if(spp == 1) {
+  #   all_d_spp <- all_d
+  # } else {
+  #   all_d_spp <- rbind(all_d_spp, all_d)
+  # }
   
   output$species <- spp_name
   output <- dplyr::filter(output, !is.na(mean))
@@ -43,6 +65,3 @@ ggplot(all_output, aes(age, mean, col = species)) +
   coord_flip() + theme_bw()
 
 ggsave("plots/glm_coefficients.png", width=7)
-
-
-

### 2024-11-08 paper.qmd
update

@@
 
+Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species’ spatial distributions.
+While many SDMs incorporate external information, such as environmental or habitat features, the majority overlook age-specific information, despite the known influence of age structure on fish distribution and abundance.
+Including age information may be particularly valuable for species exhibiting age-related spatial patterns driven by ontogenetic shifts, recruitment dynamics, and selective fishing pressures.
+Here, we develop an age-structured SDM framework using data from the west coast of the USA, and concentrate on two species with distinct life histories: North Pacific hake (*Merluccius productus*) and sablefish (*Anoplopoma fimbria*).
+Our framework incorporates spatiotemporal models to predict age-specific distributions.
+We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlihght that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes.
+Output from our models can be used to predict the spatial age composition, which may be particularly useful for large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts).
+Our approach is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.
+
 ## Introduction
 
--   Age structure is important but generally ignored in marine SDMs
+Species distribution models (SDMs) are widely used in marine ecology to understand and predict the spatial distribution of species based on environmental and demographic variables [@melo-merino2020].
+These models have informed management decisions for conservation and resource allocation, especially in the face of changing environmental conditions and human impacts, such as fishing pressure [@guisan2005; @elith2009].
+Recent advances in SDM applications in the marine environment have incorporated external drivers such as temperature [@fredston2021], habitat [@phillips2017; @laman2018].
+A second use of these models has been to predict temporal trends in species biomass from spatiotemporal fisheries independent survey data [@thorson2015].
+Trends estimated from SDMs may be included in integrated population models (stock assessments) for commercially fished species, or used to identify risks to populations of conservation concern.
+Recent developments in these models have included depth [@johnson2019] and habitat variables [@cao2017] to improve trend estimation.
 
--   Lots of reasons why age info might improve predictions – mechanisms include ontogenetic shifts, recruitment pulses, migration, fishing selectivity removing older portions of a population, etc
+While many SDMs around the world have included variables such as temperature, depth, or habitat, they have generally overlooked age structure, and instead modeled either species occurrences or densities across all age or size classes.
+Omitting information on age or size may be partially dependent on data availability (ageing fish generally requires sampling otoliths, which increases monetary costs of sampling programs), but may also be driven by the modeling choices of analysts.
+Much of the previous research around the world has demonstrated that age structure is a fundamental characteristic of fish populations, influencing spatial distribution and abundance due to mechanisms like ontogenetic shifts, recruitment variability, and selective fishing pressures [@rijnsdorp2018].
+Ignoring age-specific information in SDMs may lead to incomplete or inaccurate predictions, especially for species with strong age-related spatial gradients.
 
--   If we can make forward predictions with some skill, there's lots of wins for management. Examples include predicting areas of bycatch risk, where certain cohorts might be in the future, etc
+Incorporating age structure into SDMs may provide more accurate predictions by capturing spatial and temporal variability related to life history traits, age-specific habitat preferences, and population dynamics.
+Fish often exhibit ontogenetic shifts, where younger individuals occupy different habitats than adults due to factors like predator avoidance, habitat preferences, and diet changes [@werner1984; @ciannelli2004].
+These shifts can create age-specific distributions across habitats, particularly for species that undergo distinct life history phases.
+Additionally, recruitment pulses and migration patterns vary by age, influencing the spatial distribution of younger versus older fish.
+Fishing selectivity further accentuates these patterns by disproportionately removing older age classes, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013].
+By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relationships.
 
--   Need consistently surveyed populations to examine the forecast skill.
-Use WCBTS as case study, with hake and sablefish targeted as focal species.
-Why hake and sablefish?
-2 species with strong gradients in distribution at age – but in opposite directions and different life histories
+Because of the recursive nature of population dynamics, e.g.
+numbers at age *a* at time *t* can be used to predict numbers at age *a*+1 at time *t*+1, age-structured SDMs with predictive skill also provide a tool for forward prediction, with considerable implications for marine resource management.
+Reliable predictions of species distributions can help managers anticipate and mitigate issues such as bycatch risk by identifying regions with high densities of specific age cohorts [@lewison2004].
+Forecasting the spatial distribution of certain age classes can inform spatially explicit management strategies and help optimize fishing effort to target sustainable cohorts.
+For example, knowing the expected future locations of specific age classes could aid in creating time-area closures that protect vulnerable life stages or ensure harvests focus on more resilient portions of the population.
+However, building robust age-structured SDMs requires high-quality, long-term datasets with consistent sampling methods across both age and space to ensure the validity of these predictions.
 
--
+Around the world, age and life history information is routinely collected by a number of sampling platforms, including from fisheries and fisheries independent surveys.
+While fisheries dependent data may be used to inform age structured models, sampling from fisheries independent surveys is typically more consistent in space and time.
+The primary objective of our paper is to explore the utility of developing age - structured SDMs from fisheries independent data.
+Focusing on two species with contrasting life history characteristics in the North Pacific, the semi-pelagic North Pacific hake (*Merluccius productus*) and longer lived benthic sablefish (*Anoplopoma fimbria*), we first develop a flexible framework for modeling numbers at age.
+Second, we evaluate the ability of this framework to offer predictive skill in future forecasts, and quantify how the predictive relationship changes as a function of age.
+Finally, we demonstrate how model output may be used to examine changing species compositions through time, make maps of potential bycatch hotspots, and identify areas with higher concentrations of mature fish as potential targets of fishing effort.
 
@@
 Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish.
-\[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 – 10 years.
+[more on biological gradients] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 – 10 years.
 Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al.
@@
 For each species - age combination, we constructed a unique spatiotemporal model fit to all years except the last available year for each species.
-Individual fish were aggregated at the haul level, summarizing the total number of fish sampled and number of fish sampled of the focal age; this allowed counts to be modeled as a binomial family (with logit link).
+Individual fish were aggregated at the haul level, allowing us to model counts at age.
+We initially explored modeling counts with a Poisson or Negative Binomial distribution, but as the distributions of numbers at age are dominated by 0s, these models failed to converge.
+Instead, we summarized the total number of fish sampled and number of fish sampled of the focal age; this allowed counts to be modeled as a binomial family (logit link).
 Second, we constructed a spatiotemporal model as an extension of Generalized Linear Mixed Models (GLMMs) such that the prediction in location $s$ and time $t$ can be written as
@@
 
-Figure @fig-hake-spatial-anomaly shows a concentration of age-1 (mostly immature) hake near the coast, and then a clear N/S split for age-2s and 3s with higher probabilities of occurrence in the South (on avg).
-Age-4 hake appear to have a more patchy & northern distribution
+One useful output from our spatiotemporal models is the estimated spatial field, $\omega_s$, representing the average spatial distribution across years.
+Results from our model of North Pacific hake show a concentration of age-1 hake near the coast, and more of latitudinal north - south break for 2 - 3 year old hake (with higher occurrences in the South) @fig-hake-spatial-anomaly.
+As expected, age 4 hake appear to have a northern distribution, though the gradient is less clear than for ages 2 - 3.
+For sablefish, our models estimate a concentration of age 0 - 2 sablefish in shallow water, with a deeper distribution by age 3 (@fig-sablefish-spatial-anomaly).
+Latitudinally, there are several areas in central Oregon that appear to have consistently lower occurrences than average for older age classes (@fig-sablefish-spatial-anomaly).
 
-Figure @fig-sablefish-spatial-anomaly shows age 0-2 sablefish concentrated in shallow water, and more fish moving offshore by age 3.
-Several interesting patches in coastal OR have very low occurrence rates
+Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time.
+For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, @fig-hake-spatial-composition) and strong sablefish cohorts (2013, 2016, 2020, @fig-sablefish-spatial-composition).
+[expand on this]
 
-Figure @fig-spatial-parameters shows that for most ages, we see a decrease in the spatial and spatiotemporal variances as well as a decrease in the spatial range.
-This means that the fields for older fish become less peaky -- and the distribution of where fish are becomes patchier ( duh!)
+When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, we see a decrease in the spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters).
+Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, suggesting that the fields estimated for older fish are less variable.
+Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ ~ 0.13).
+The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).
 
-Figure @fig-glm-coefficients shows that in general, there's more predictability (or maybe persistence for sablefish than hake).
-Using the distribution of age 1 and 2 fish to predict age 2 and 3 yields similar coefficients across species -- but for older age 4-5 sablefish, these coefficients increase a lot
+Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish (@fig-glm-coefficients).
 
@@
 knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
+```
+
+\break
+
+```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial composition for Pacific hake; for plotting purposes only ages 1 - 5 and the last 10 years of available predictions are shown.", echo=FALSE}
+knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
+```
+
+\break
+
+```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial composition for sablefish; for plotting purposes only ages 0 - 4 and the last 10 years of available predictions are shown.", echo=FALSE}
+knitr::include_graphics("plots/spatial_composition_sablefish.png")
 ```
@@
 \break
+\clearpage
+## References

### 2024-11-08 02_recursive_model_nb.R

@@ -1,0 +1,181 @@
+library(nwfscSurvey)
+library(sdmTMB)
+library(tidyverse)
+library(ggplot2)
+library(DHARMa)
+library(surveyjoin)
+
+spp_name <- c("Pacific hake", "sablefish")[1]
+
+if(spp_name == "Pacific hake") {
+  min_age <- 1 # not many age 0s consistently sampled
+  max_age <- 6
+  min_year <- 2007
+  max_year <- 2019
+} 
+if(spp_name == "sablefish") {
+  min_age <- 0
+  max_age <- 10
+  min_year <- 2003
+  max_year <- 2023
+}
+all_years <- min_year:max_year
+
+bio <- pull_bio(common_name = spp_name, survey="NWFSC.Combo")
+names(bio) <- tolower(names(bio))
+haul <- pull_haul(survey="NWFSC.Combo")
+names(haul) <- tolower(names(haul))
+
+# filter ages to be in core range
+d <- bio
+d <- dplyr::filter(d, sex == "F", !is.na(age), 
+                   year >= min_year,
+                   year <= max_year)
+
+# group data by age and year 
+grouped <- dplyr::group_by(d, age, year, trawl_id) |>
+  dplyr::summarise(n = n(), # n is the number of observations of this age in this year
+                   present = ifelse(n > 1, 1, 0), # not used
+                   lat = latitude_dd[1], 
+                   lon = longitude_dd[1]) |>
+  dplyr::ungroup()
+grouped_total <- dplyr::group_by(d, trawl_id) |>
+  dplyr::summarise(n_total = n()) |>
+  dplyr::ungroup()
+grouped <- dplyr::left_join(grouped, grouped_total)
+
+# add trawl data -- after expanding across years and ages
+grid <- expand.grid(trawl_id = unique(grouped$trawl_id), 
+                    age = unique(grouped$age))
+yrs <- dplyr::group_by(grouped, trawl_id) |>
+  dplyr::summarise(year = year[1])
+grid <- dplyr::left_join(grid, yrs)
+locs <- dplyr::group_by(grouped, trawl_id) |>
+  dplyr::summarize(lat = lat[1], lon = lon[1])
+
+d <- dplyr::left_join(grid[,c("trawl_id","age")], grouped, by = c("trawl_id","age")) 
+d$n[which(is.na(d$n))] <- 0
+d$n_total[which(is.na(d$n_total))] <- 0
+
[…]

### 2024-11-08 plot_spatial_compositions.R
update

@@ -1,0 +1,66 @@
+library(tidyverse)
+library(ggplot2)
+
+spp_name <- c("Pacific hake", "sablefish")[2]
+
+if(spp_name == "Pacific hake") {
+  min_age <- 1 # not many age 0s consistently sampled
+  max_age <- 6
+  years <- 2010:2019
+  ages <- seq(min_age, min_age+4)
+} 
+if(spp_name == "sablefish") {
+  min_age <- 0
+  max_age <- 10
+  years <- 2014:2023
+  ages <- seq(min_age, min_age+4)
+}
+
+for(a in min_age:(max_age - 1)) {
+  pred <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+  pred <- dplyr::select(pred, lon, lat, X, Y, year, est) |>
+    dplyr::mutate(age = a) |>
+    dplyr::filter(year %in% years)
+  if(a == min_age) {
+    pred_all <- pred
+  } else {
+    pred_all <- rbind(pred_all, pred)
+  }
+}
+
+# now convert the 'est' to normalized probabilities across ages
+pred_all <- dplyr::mutate(pred_all, p = plogis(est)) |>
+  dplyr::group_by(X,Y,year) |>
+  dplyr::mutate(p_norm = p / sum(p)) |>
+  dplyr::filter(age %in% ages)
+
+map_data <- rnaturalearth::ne_countries(
+  scale = "medium",
+  returnclass = "sf", country = "united states of america")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+coast <- suppressWarnings(suppressMessages(
+  sf::st_crop(map_data,
+              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
+
+utm_zone10 <- 3157
+coast_proj <- sf::st_transform(coast, crs = utm_zone10)
+
+# Define the new colors
+land_color <- "#E0CDA9"  # beige for land
+ocean_color <- "#D3EAF2"  # grayish blue for ocean
+
+# Plot coast with custom land and ocean colors
+g <- ggplot(coast_proj) + 
+  geom_tile(data = pred_all, aes(x = X * 1000, y = Y * 1000, col = p_norm)) + 
+  geom_sf(fill = land_color) +  # Set land color
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Probability") +  # legend title 
[…]

### 2024-12-04 paper.qmd
Update paper.qmd

@@
 
-Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species’ spatial distributions.
+Species distribution models (SDMs) are widely used in fisheries and ecology, providing insights into species' spatial distributions.
 While many SDMs incorporate external information, such as environmental or habitat features, the majority overlook age-specific information, despite the known influence of age structure on fish distribution and abundance.
@@
 Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish.
-[more on biological gradients] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 – 10 years.
+\[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years.
 Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al.
 2024) and sablefish is frequently assessed because of its high commercial value.
-For our analysis, we used hake collected 2007 – 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 – 2023 (mean 1314.4 individuals sampled per year).
+For our analysis, we used hake collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per year).
 
@@
 
-For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 – 5, and for sablefish ages 0 – 9 (data from these ages represented 54% and 75% of the total aged fish, respectively).
+For each species, we filtered by sex to focus on females and truncated ages to focus on those with the highest data availability; for hake this included ages 1 -- 5, and for sablefish ages 0 -- 9 (data from these ages represented 54% and 75% of the total aged fish, respectively).
 For each species - age combination, we constructed a unique spatiotemporal model fit to all years except the last available year for each species.
@@
 For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, @fig-hake-spatial-composition) and strong sablefish cohorts (2013, 2016, 2020, @fig-sablefish-spatial-composition).
-[expand on this]
+\[expand on this\]
 
@@
 Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, suggesting that the fields estimated for older fish are less variable.
-Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ ~ 0.13).
+Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13).
 The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).
@@
 
+```{=tex}
 \break
 \clearpage
+```
 ## References

### 2024-12-14 02_recursive_model.R
update code

@@ -1,180 +1,0 @@
-library(nwfscSurvey)
-library(sdmTMB)
-library(tidyverse)
-library(ggplot2)
-library(DHARMa)
-library(surveyjoin)
-
-spp_name <- c("Pacific hake", "sablefish")[2]
-
-if(spp_name == "Pacific hake") {
-  min_age <- 1 # not many age 0s consistently sampled
-  max_age <- 6
-  min_year <- 2007
-  max_year <- 2019
-} 
-if(spp_name == "sablefish") {
-  min_age <- 0
-  max_age <- 10
-  min_year <- 2003
-  max_year <- 2023
-}
-all_years <- min_year:max_year
-
-bio <- pull_bio(common_name = spp_name, survey="NWFSC.Combo")
-names(bio) <- tolower(names(bio))
-haul <- pull_haul(survey="NWFSC.Combo")
-names(haul) <- tolower(names(haul))
-
-# filter ages to be in core range
-d <- bio
-d <- dplyr::filter(d, sex == "F", !is.na(age), 
-                   year >= min_year,
-                   year <= max_year)
-
-# group data by age and year 
-grouped <- dplyr::group_by(d, age, year, trawl_id) |>
-  dplyr::summarise(n = n(), # n is the number of observations of this age in this year
-                   present = ifelse(n > 1, 1, 0), # not used
-                   lat = latitude_dd[1], 
-                   lon = longitude_dd[1]) |>
-  dplyr::ungroup()
-grouped_total <- dplyr::group_by(d, trawl_id) |>
-  dplyr::summarise(n_total = n()) |>
-  dplyr::ungroup()
-grouped <- dplyr::left_join(grouped, grouped_total)
-
-# add trawl data -- after expanding across years and ages
-grid <- expand.grid(trawl_id = unique(grouped$trawl_id), 
-                    age = unique(grouped$age))
-yrs <- dplyr::group_by(grouped, trawl_id) |>
-  dplyr::summarise(year = year[1])
-grid <- dplyr::left_join(grid, yrs)
-locs <- dplyr::group_by(grouped, trawl_id) |>
-  dplyr::summarize(lat = lat[1], lon = lon[1])
-
-d <- dplyr::left_join(grid[,c("trawl_id","age")], grouped, by = c("trawl_id","age")) 
-d$n[which(is.na(d$n))] <- 0
-d$n_total[which(is.na(d$n_total))] <- 0
-
[…]

### 2024-12-14 02_recursive_model_poisson.R

@@ -79,6 +79,41 @@ glms <- list()
 survey_grid <- surveyjoin::nwfsc_grid
 survey_grid <- add_utm_columns(survey_grid, ll_names = c("lon","lat"))
 
+# Fit the null model
+# subset_age <- dplyr::filter(subset, 
+#                             age %in% seq(min(subset$age), (max(subset$age) - 1)), 
+#                             year < max_year, n_total > 0) |>
+#   dplyr::group_by(trawl_id) |>
+#   dplyr::summarise(n = sum(n), n_total = n_total[1],
+#                    lat = lat[1], lon = lon[1],
+#                    year = year[1],
+#                    X = X[1], Y = Y[1]) |>
+#   as.data.frame()
+# mesh <- make_mesh(subset_age, xy_cols = c("X","Y"), cutoff=50)
+# fit <- sdmTMB(n ~ 1,
+#               time_varying = ~ 1, # time-varying intercept
+#               time_varying_type = "ar1",
+#               spatial = "on", # spatial field on
+#               spatiotemporal="ar1", # random walk in spatiotemporal
+#               time="year",
+#               mesh=mesh,
+#               family = poisson(),
+#               offset = log(subset_age$n_total),
+#               data=subset_age,
+#               priors = sdmTMBpriors(
+#                 ar1_rho = normal(0, 1),
+#                 matern_s = pc_matern(range_gt = 5, sigma_lt = 1)
+#               ),
+#               extra_time = all_years[which(all_years %in% as.numeric(names(table(subset_age$year))) ==FALSE)])
+# # Make predictions to next year
+# pred_df <- dplyr::filter(subset, 
+#                             age %in% seq(min(subset$age)+1, (max(subset$age))), 
+#                             year == max_year, n_total > 0)
+# pred <- predict(object = fit, newdata = pred_df, offset = pred_df$n_total)
+# pred_df$expected_n <- exp(pred$est)
+# saveRDS(pred_df, paste0("predictions/",spp_name,"_null",".rds"))
+
+# Fit the age-structured model
 for(a in min(subset$age):(max(subset$age) - 1)) {
   # Take all age 'a' fish from 2003 - 2022. construct a coarse mesh (small n)
   subset_age <- dplyr::filter(subset, age == a, year < max_year, n_total > 0)
@@ -92,9 +127,12 @@ for(a in min(subset$age):(max(subset$age) - 1)) {
                            spatiotemporal="ar1", # random walk in spatiotemporal
                            time="year",
[…]

### 2024-12-14 03_poisson_regression.R

@@ -2,6 +2,7 @@ library(tidyverse)
 library(ggplot2)
 library(DHARMa)
 library(viridis)
+library(glmmTMB)
 
 for(spp in 1:2) {
   spp_name <- c("Pacific hake", "sablefish")[spp]
@@ -31,6 +32,19 @@ for(spp in 1:2) {
     # } else {
     #   all_d <- rbind(all_d, d)
     # }
+    # Fit a GLMM with year-specific random effects
+    d$x <- log(d$expected_n)
+    g <- glmmTMB(n ~ x|year, data = d, family = "poisson")
+    coefs <- coef(g)[[1]]$year
+    coefs$species <- spp_name
+    coefs$age <- a
+    coefs$year <- as.numeric(rownames(coefs))
+    if(a == min_age & spp == 1) {
+      all_coefs <- coefs
+    } else {
+      all_coefs <- rbind(all_coefs, coefs)
+    }
+    
   }
   # all_d$species <- spp_name
   # ggplot(data = dplyr::filter(all_d_spp, species=="sablefish"), aes(sample = resid)) +
@@ -64,4 +78,36 @@ ggplot(all_output, aes(age, mean, col = species)) +
   ylab("Estimated slope") + 
   coord_flip() + theme_bw()
 
-ggsave("plots/glm_coefficients.png", width=7)
+ggsave("plots/glm_coefficients.png", width=6, height = 5)
+
+
+# look at coefficients from the year model, as a diagnostic
+dplyr::filter(all_coefs) |>
+  ggplot(aes(year, x, group=age, col = as.factor(age))) + 
+  geom_line() + xlab("Year") + 
+  ylab("Estimated random slope") +
+  scale_color_viridis_d(option="magma", begin = 0.2, end = 0.8) + 
+  theme_bw() + 
+  theme(strip.background = element_rect(fill="white")) + 
+  facet_wrap(~species, ncol = 1, scale="free_y")
+ggsave("plots/glm_coefficients.png", width=7, height = 5)
+
+library(tidyr)
+library(ggplot2)
+library(reshape2)
+
+df_wide <- dplyr::filter(all_coefs, species == "Pacific hake") |>
+  dplyr::select(x, species, age, year) |>
+  pivot_wider(names_from = age, values_from = x, names_prefix = "age_")
+correlation_matrix <- cor(df_wide[,-c(1:2)], use = "complete.obs") # Exclude the year column
[…]

### 2024-12-14 05_spatial_bycatch_risk_sablefish.R

@@ -15,57 +15,52 @@ spp <- 2
     max_age <- 6
   }
   
-  for(a in min_age:(max_age - 1)) {
-    d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
-    d$species <- spp_name
-    d$age <- a
-    if(a == min_age) {
-      all_output <- d
-    } else {
-      all_output <- rbind(all_output, d)
-    }
-  }
-  
-sablefish <- dplyr::filter(all_output, species == "sablefish", year %in% seq(2015,2023,by=2)) |>
-  dplyr::mutate(p = plogis(est)) |>
-  dplyr::group_by(year, X, Y) |>
-  dplyr::mutate(p = p / sum(p)) |> # normalize by ages
-  dplyr::ungroup() |>
-  dplyr::filter(age <= 2) |>
-  dplyr::group_by(year, X, Y) |>
-  dplyr::summarise(p = sum(p)) 
+d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
 
-# also de-mean the values by year so this is relative
-sablefish <- dplyr::group_by(sablefish, year) |>
-  dplyr::mutate(centered_p = p - mean(p))
+# read in ports
+ports_rad <- readRDS("~/Documents/Github Projects/spatial-age-models/ports_rad.rds")
+d$dist_AST <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="AST")])^2 + 
+                     (d$Y - ports_rad$Y[which(ports_rad$Pcid=="AST")])^2)
+d$dist_COS <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="COS")])^2 + 
+                     (d$Y - ports_rad$Y[which(ports_rad$Pcid=="COS")])^2)
+d$dist_BRK <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="BRK")])^2 + 
+                     (d$Y - ports_rad$Y[which(ports_rad$Pcid=="BRK")])^2)
+d$dist_CRS <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="CRS")])^2 + 
+                     (d$Y - ports_rad$Y[which(ports_rad$Pcid=="CRS")])^2)
+d$dist_ERK <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="ERK")])^2 + 
[…]

### 2024-12-14 06_plot_spatial_compositions.R

@@ -1,0 +1,96 @@
+library(tidyverse)
+library(ggplot2)
+
+spp_name <- c("Pacific hake", "sablefish")[2]
+
+if(spp_name == "Pacific hake") {
+  min_age <- 1 # not many age 0s consistently sampled
+  max_age <- 6
+  years <- 2010:2019
+  ages <- seq(min_age, min_age+4)
+} 
+if(spp_name == "sablefish") {
+  min_age <- 0
+  max_age <- 10
+  years <- 2014:2023
+  ages <- seq(min_age, min_age+4)
+}
+
+for(a in min_age:(max_age - 1)) {
+  pred <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+  pred <- dplyr::select(pred, lon, lat, X, Y, year, est) |>
+    dplyr::mutate(age = a) |>
+    dplyr::filter(year %in% years)
+  if(a == min_age) {
+    pred_all <- pred
+  } else {
+    pred_all <- rbind(pred_all, pred)
+  }
+}
+
+# now convert the 'est' to normalized probabilities across ages
+pred_all <- dplyr::mutate(pred_all, p = plogis(est)) |>
+  dplyr::group_by(X,Y,year) |>
+  dplyr::mutate(p_norm = p / sum(p)) |>
+  dplyr::filter(age %in% ages)
+
+map_data <- rnaturalearth::ne_countries(
+  scale = "medium",
+  returnclass = "sf", country = "united states of america")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+coast <- suppressWarnings(suppressMessages(
+  sf::st_crop(map_data,
+              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
+
+utm_zone10 <- 3157
+coast_proj <- sf::st_transform(coast, crs = utm_zone10)
+
+# Define the new colors
+land_color <- "#E0CDA9"  # beige for land
+ocean_color <- "#D3EAF2"  # grayish blue for ocean
+
+# Plot coast with custom land and ocean colors
+g <- ggplot(coast_proj) + 
+  geom_tile(data = pred_all, aes(x = X * 1000, y = Y * 1000, col = p_norm)) + 
+  geom_sf(fill = land_color) +  # Set land color
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Expected number") +  # legend title 
[…]

### 2024-12-14 07_sablefish_age5.R

@@ -27,11 +27,13 @@ spp <- 2
   }
   
 sablefish <- dplyr::filter(all_output, species == "sablefish", 
-                           year %in% seq(2015,2023,by=2), age == 5) |>
-  dplyr::mutate(p = plogis(est))
+                           year %in% seq(2015,2023,by=2), age > 5) |>
+  dplyr::mutate(p = exp(est)) |>
+  dplyr::group_by(X, Y, year) |>
+  dplyr::summarize(p = sum(p))
 
-sablefish <- dplyr::group_by(sablefish, year) |>
-  dplyr::mutate(centered_p = p - mean(p))
+# sablefish <- dplyr::group_by(sablefish, year) |>
+#   dplyr::mutate(centered_p = p - mean(p))
 
 map_data <- rnaturalearth::ne_countries(
   scale = "medium",
@@ -50,10 +52,22 @@ land_color <- "#E0CDA9"  # beige for land
 ocean_color <- "#D3EAF2"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
-ggplot(coast_proj) + 
+# g1 <- ggplot(coast_proj) + 
+#   geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
+#   scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Probability") +  # legend title 
+#   #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+#   geom_sf(fill = land_color) +  # Set land color
+#   theme_light() + 
+#   labs(x = "Longitude", y = "Latitude") +
+#   theme(panel.background = element_rect(fill = ocean_color),  
+#         strip.text = element_text(color = "black"),  # Set facet label text color to black
+#         strip.background = element_rect(fill = "white")) +  # Set facet label background to white
+#   facet_wrap(~year, nrow = 1)
+
+g2 <- ggplot(coast_proj) + 
   geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
-  #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Occurrence") +  # legend title 
-  scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE") +  # legend title 
[…]

### 2024-12-14 archive/02_recursive_model.R
plots

@@ -1,0 +1,180 @@
+library(nwfscSurvey)
+library(sdmTMB)
+library(tidyverse)
+library(ggplot2)
+library(DHARMa)
+library(surveyjoin)
+
+spp_name <- c("Pacific hake", "sablefish")[2]
+
+if(spp_name == "Pacific hake") {
+  min_age <- 1 # not many age 0s consistently sampled
+  max_age <- 6
+  min_year <- 2007
+  max_year <- 2019
+} 
+if(spp_name == "sablefish") {
+  min_age <- 0
+  max_age <- 10
+  min_year <- 2003
+  max_year <- 2023
+}
+all_years <- min_year:max_year
+
+bio <- pull_bio(common_name = spp_name, survey="NWFSC.Combo")
+names(bio) <- tolower(names(bio))
+haul <- pull_haul(survey="NWFSC.Combo")
+names(haul) <- tolower(names(haul))
+
+# filter ages to be in core range
+d <- bio
+d <- dplyr::filter(d, sex == "F", !is.na(age), 
+                   year >= min_year,
+                   year <= max_year)
+
+# group data by age and year 
+grouped <- dplyr::group_by(d, age, year, trawl_id) |>
+  dplyr::summarise(n = n(), # n is the number of observations of this age in this year
+                   present = ifelse(n > 1, 1, 0), # not used
+                   lat = latitude_dd[1], 
+                   lon = longitude_dd[1]) |>
+  dplyr::ungroup()
+grouped_total <- dplyr::group_by(d, trawl_id) |>
+  dplyr::summarise(n_total = n()) |>
+  dplyr::ungroup()
+grouped <- dplyr::left_join(grouped, grouped_total)
+
+# add trawl data -- after expanding across years and ages
+grid <- expand.grid(trawl_id = unique(grouped$trawl_id), 
+                    age = unique(grouped$age))
+yrs <- dplyr::group_by(grouped, trawl_id) |>
+  dplyr::summarise(year = year[1])
+grid <- dplyr::left_join(grid, yrs)
+locs <- dplyr::group_by(grouped, trawl_id) |>
+  dplyr::summarize(lat = lat[1], lon = lon[1])
+
+d <- dplyr::left_join(grid[,c("trawl_id","age")], grouped, by = c("trawl_id","age")) 
+d$n[which(is.na(d$n))] <- 0
+d$n_total[which(is.na(d$n_total))] <- 0
+
[…]

### 2024-12-14 paper.qmd
update

@@
 age 3 hake in years 2007 -- 2018 used to predict the distribution of age 4 hake in 2008 -- 2019).
-We related the predicted probabilities of occurrence $\hat{p}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{p}_{\boldsymbol{s},t}$.
+We related the predictions $\hat{p}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{p}_{\boldsymbol{s},t}$.
 We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$.
-The exponentiated slope parameter $exp(\beta_{1})$ represents a the change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival.
+The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival.
+As a diagnostic check and to understand whether there was a potential time component, we fit a second model as a GLMM with R package glmmTMB; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.
 
-As a second validation, we compared aggregated predictions across all ages to a SDM model that used total biomass as a response.
+### Using predictions to quantify bycatch risk
 
-### Links to fisheries data?
+To illustrate potential benefits of this kind of modeling, we used the above approach to generate spatiotemporal predictions of age-1 sablefish (from age-0 sablefish the year prior).
+\[Need some text here on background why sablefish as bycatch is important\].
+Gridded predictions from the Poisson GLM represent the expected number of fish per unit of effort in each location.
+To turn these into a more useful measure of risk, we calculate the total number of age-1 sablefish per unit of effort within a given radius of major fishing ports on the US west coast.
+For consistency with previous work \[TODO: cite ESR\] we used a radius of 232km.
 
@@
 For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, @fig-hake-spatial-composition) and strong sablefish cohorts (2013, 2016, 2020, @fig-sablefish-spatial-composition).
-\[expand on this\]
+These predictions are also useful in revealing subtle differences between strong cohort years.
+For example, the 2009 cohort of Pacific hake (age 1 in 2010) were predominantly distributed in the northern part of the domain, while the 2013 cohort had hotspots along the entire US west coast.
+Similarly for sablefish, the 2016 cohort appeared to be more concentrated in the north, versus the more widely distributed 2013 cohort.
 
-When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, we see a decrease in the spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters).
-Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, suggesting that the fields estimated for older fish are less variable.
+When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters).
+Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, meaning that spatial fields estimated for older fish are less variable.
 Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13).
@@
 Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish (@fig-glm-coefficients).
+When comparing our initial Poisson GLM to a more complicated GLMM with random year effects in the intercept and slope, we found little support for shared correlations across age classes of a given species.
+Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another).
+With the exception of a strong positive correlation between age-0 and age-1 sablefish (\$\$)
 
@@
 
-\break
+\newpage
 
-```{r label="fig-hake-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for Pacific hake ages 1 -- 4.", echo=FALSE}
+```{r label="fig-hake-spatial-anomaly", fig.cap="Estimated spatial anomalies for Pacific hake ages 1 -- 5. These fields represent the average anomalies across all years, 2003 -- 2023.", echo=FALSE}
 knitr::include_graphics("plots/hake_spatial_anomalies.png")
@@
 
-\break
+\newpage
 
-```{r label="fig-sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies (shared across all years) for sablefish ages 0 -- 5.", echo=FALSE}
+```{r label="fig-sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies for sablefish ages 0 -- 9. These fields represent the average anomalies across all years, 2003 -- 2023.", echo=FALSE}
 knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
@@
 
-\break
+\newpage
 
-```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial composition for Pacific hake; for plotting purposes only ages 1 - 5 and the last 10 years of available predictions are shown.", echo=FALSE}
-knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
+```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial composition for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014) and columns represent predictions for ages 1 - 5.
+Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
+knitr::include_graphics("plots/spatial_composition_subset_Pacific hake.png")
 ```
 
-\break
+\newpage
 
-```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial composition for sablefish; for plotting purposes only ages 0 - 4 and the last 10 years of available predictions are shown.", echo=FALSE}
-knitr::include_graphics("plots/spatial_composition_sablefish.png")
+```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial composition for sablefish; rows represent strong cohorts in our dataset (2013, 2016, 2019) and columns represent predictions for ages 0 - 4.
+Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
+knitr::include_graphics("plots/spatial_composition_subset_sablefish.png")
 ```
 
-\break
+\newpage
 
@@
 
-\break
+\clearpage
+
+\newpage
 
@@
 
-\break
+\newpage
 
@@
 knitr::include_graphics("plots/spatial_parameters.png")
+```
+
+\newpage
+
+```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial composition for Pacific hake; rows represent years and columns represent predictions for ages 1 - 5.", echo=FALSE}
+knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
+```
+
+\newpage
+
+```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial composition for sablefish; rows represent years and columns represent predictions for ages 0 - 4.", echo=FALSE}
+
+knitr::include_graphics("plots/spatial_composition_sablefish.png")
 ```

### 2024-12-14 plot_spatial_compositions.R

@@ -1,66 +1,0 @@
-library(tidyverse)
-library(ggplot2)
-
-spp_name <- c("Pacific hake", "sablefish")[2]
-
-if(spp_name == "Pacific hake") {
-  min_age <- 1 # not many age 0s consistently sampled
-  max_age <- 6
-  years <- 2010:2019
-  ages <- seq(min_age, min_age+4)
-} 
-if(spp_name == "sablefish") {
-  min_age <- 0
-  max_age <- 10
-  years <- 2014:2023
-  ages <- seq(min_age, min_age+4)
-}
-
-for(a in min_age:(max_age - 1)) {
-  pred <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
-  pred <- dplyr::select(pred, lon, lat, X, Y, year, est) |>
-    dplyr::mutate(age = a) |>
-    dplyr::filter(year %in% years)
-  if(a == min_age) {
-    pred_all <- pred
-  } else {
-    pred_all <- rbind(pred_all, pred)
-  }
-}
-
-# now convert the 'est' to normalized probabilities across ages
-pred_all <- dplyr::mutate(pred_all, p = plogis(est)) |>
-  dplyr::group_by(X,Y,year) |>
-  dplyr::mutate(p_norm = p / sum(p)) |>
-  dplyr::filter(age %in% ages)
-
-map_data <- rnaturalearth::ne_countries(
-  scale = "medium",
-  returnclass = "sf", country = "united states of america")
-# Crop the polygon for plotting and efficiency:
-# st_bbox(map_data) # find the rough coordinates
-coast <- suppressWarnings(suppressMessages(
-  sf::st_crop(map_data,
-              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
-
-utm_zone10 <- 3157
-coast_proj <- sf::st_transform(coast, crs = utm_zone10)
-
-# Define the new colors
-land_color <- "#E0CDA9"  # beige for land
-ocean_color <- "#D3EAF2"  # grayish blue for ocean
-
-# Plot coast with custom land and ocean colors
-g <- ggplot(coast_proj) + 
-  geom_tile(data = pred_all, aes(x = X * 1000, y = Y * 1000, col = p_norm)) + 
-  geom_sf(fill = land_color) +  # Set land color
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Probability") +  # legend title 
[…]

### 2024-12-16 07_sablefish_age5.R
cleaning

@@ -1,81 +1,0 @@
-library(tidyverse)
-library(ggplot2)
-library(DHARMa)
-library(viridis)
-
-spp <- 2
-  spp_name <- c("Pacific hake", "sablefish")[spp]
-  
-  if(spp_name == "Pacific hake") {
-    min_age <- 1 # not many age 0s consistently sampled
-    max_age <- 6
-  } 
-  if(spp_name == "sablefish") {
-    min_age <- 0
-    max_age <- 10
-  }
-  
-  for(a in min_age:(max_age - 1)) {
-    d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
-    d$species <- spp_name
-    d$age <- a
-    if(a == min_age) {
-      all_output <- d
-    } else {
-      all_output <- rbind(all_output, d)
-    }
-  }
-  
-sablefish <- dplyr::filter(all_output, species == "sablefish", 
-                           year %in% seq(2015,2023,by=2), age > 5) |>
-  dplyr::mutate(p = exp(est)) |>
-  dplyr::group_by(X, Y, year) |>
-  dplyr::summarize(p = sum(p))
-
-# sablefish <- dplyr::group_by(sablefish, year) |>
-#   dplyr::mutate(centered_p = p - mean(p))
-
-map_data <- rnaturalearth::ne_countries(
-  scale = "medium",
-  returnclass = "sf", country = "united states of america")
-# Crop the polygon for plotting and efficiency:
-# st_bbox(map_data) # find the rough coordinates
-coast <- suppressWarnings(suppressMessages(
-  sf::st_crop(map_data,
-              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
-
-utm_zone10 <- 3157
-coast_proj <- sf::st_transform(coast, crs = utm_zone10)
-
-# Define the new colors
-land_color <- "#E0CDA9"  # beige for land
-ocean_color <- "#D3EAF2"  # grayish blue for ocean
-
-# Plot coast with custom land and ocean colors
-# g1 <- ggplot(coast_proj) + 
-#   geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
-#   scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Probability") +  # legend title 
-#   #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
-#   geom_sf(fill = land_color) +  # Set land color
-#   theme_light() + 
[…]

### 2024-12-16 archive/07_sablefish_age5.R
cleaning

@@ -1,0 +1,81 @@
+library(tidyverse)
+library(ggplot2)
+library(DHARMa)
+library(viridis)
+
+spp <- 2
+  spp_name <- c("Pacific hake", "sablefish")[spp]
+  
+  if(spp_name == "Pacific hake") {
+    min_age <- 1 # not many age 0s consistently sampled
+    max_age <- 6
+  } 
+  if(spp_name == "sablefish") {
+    min_age <- 0
+    max_age <- 10
+  }
+  
+  for(a in min_age:(max_age - 1)) {
+    d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+    d$species <- spp_name
+    d$age <- a
+    if(a == min_age) {
+      all_output <- d
+    } else {
+      all_output <- rbind(all_output, d)
+    }
+  }
+  
+sablefish <- dplyr::filter(all_output, species == "sablefish", 
+                           year %in% seq(2015,2023,by=2), age > 5) |>
+  dplyr::mutate(p = exp(est)) |>
+  dplyr::group_by(X, Y, year) |>
+  dplyr::summarize(p = sum(p))
+
+# sablefish <- dplyr::group_by(sablefish, year) |>
+#   dplyr::mutate(centered_p = p - mean(p))
+
+map_data <- rnaturalearth::ne_countries(
+  scale = "medium",
+  returnclass = "sf", country = "united states of america")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+coast <- suppressWarnings(suppressMessages(
+  sf::st_crop(map_data,
+              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
+
+utm_zone10 <- 3157
+coast_proj <- sf::st_transform(coast, crs = utm_zone10)
+
+# Define the new colors
+land_color <- "#E0CDA9"  # beige for land
+ocean_color <- "#D3EAF2"  # grayish blue for ocean
+
+# Plot coast with custom land and ocean colors
+# g1 <- ggplot(coast_proj) + 
+#   geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
+#   scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Probability") +  # legend title 
+#   #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+#   geom_sf(fill = land_color) +  # Set land color
+#   theme_light() + 
[…]

### 2024-12-16 03_poisson_regression.R
plots

@@ -86,11 +86,11 @@ dplyr::filter(all_coefs) |>
   ggplot(aes(year, x, group=age, col = as.factor(age))) + 
   geom_line() + xlab("Year") + 
   ylab("Estimated random slope") +
-  scale_color_viridis_d(option="magma", begin = 0.2, end = 0.8) + 
+  scale_color_viridis_d(option="magma", begin = 0.2, end = 0.8, name = "Age") + 
   theme_bw() + 
   theme(strip.background = element_rect(fill="white")) + 
   facet_wrap(~species, ncol = 1, scale="free_y")
-ggsave("plots/glm_coefficients.png", width=7, height = 5)
+ggsave("plots/glm_coefficients_time.png", width=7, height = 5)
 
 library(tidyr)
 library(ggplot2)

### 2024-12-16 05_spatial_bycatch_risk_sablefish.R

@@ -15,7 +15,39 @@ spp <- 2
     max_age <- 6
   }
   
-d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+d <- readRDS(paste0("predictions/",spp_name,"_",min_age+1,"_surveygrid.rds"))
+
+sablefish <- dplyr::filter(d, 
+                           year %in% seq(2015,2023,by=2)) |>
+  dplyr::mutate(p = exp(est))
+map_data <- rnaturalearth::ne_countries(
+  scale = "medium",
+  returnclass = "sf", country = "united states of america")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+coast <- suppressWarnings(suppressMessages(
+  sf::st_crop(map_data,
+              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
+
+utm_zone10 <- 3157
+coast_proj <- sf::st_transform(coast, crs = utm_zone10)
+
+# Define the new colors
+land_color <- "#E0CDA9"  # beige for land
+ocean_color <- "#D3EAF2"  # grayish blue for ocean
+
+p1 <- ggplot(coast_proj) + 
+  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
+  #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+  geom_sf(fill = land_color) +  # Set land color
+  theme_light() + 
+  labs(x = "Longitude", y = "Latitude") +
+  theme(panel.background = element_rect(fill = ocean_color),  
+        strip.text = element_text(color = "black"),  # Set facet label text color to black
+        strip.background = element_rect(fill = "white"),
+        axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5)) +  # Set facet label background to white
+  facet_wrap(~year, nrow = 1)
 
 # read in ports
 ports_rad <- readRDS("~/Documents/Github Projects/spatial-age-models/ports_rad.rds")
@@ -55,12 +87,14 @@ distances_long <- dplyr::rename(distances_long, Pcid = port)
 distances_long <- dplyr::left_join(distances_long, ports_rad[,c("Pcid","Name")])
[…]

### 2024-12-16 06_plot_spatial_compositions.R

@@ -29,9 +29,9 @@ for(a in min_age:(max_age - 1)) {
 }
 
 # now convert the 'est' to normalized probabilities across ages
-pred_all <- dplyr::mutate(pred_all, p = plogis(est)) |>
-  dplyr::group_by(X,Y,year) |>
-  dplyr::mutate(p_norm = p / sum(p)) |>
+pred_all <- dplyr::mutate(pred_all, p = exp(est)) |>
+  #dplyr::group_by(X,Y,year) |>
+  #dplyr::mutate(p_norm = p / sum(p)) |>
   dplyr::filter(age %in% ages)
 
 map_data <- rnaturalearth::ne_countries(
@@ -52,9 +52,9 @@ ocean_color <- "#D3EAF2"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 g <- ggplot(coast_proj) + 
-  geom_tile(data = pred_all, aes(x = X * 1000, y = Y * 1000, col = p_norm)) + 
+  geom_tile(data = pred_all, aes(x = X * 1000, y = Y * 1000, col = p)) + 
   geom_sf(fill = land_color) +  # Set land color
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Expected number") +  # legend title 
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
   #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
@@ -80,9 +80,9 @@ if(spp_name == "Pacific hake") {
 
 
 g <- ggplot(coast_proj) + 
-  geom_point(data = pred_subset, aes(x = X * 1000, y = Y * 1000, col = p_norm),size=0.05) + 
+  geom_point(data = pred_subset, aes(x = X * 1000, y = Y * 1000, col = p),size=0.05) + 
   geom_sf(fill = land_color) +  # Set land color
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "Expected number") +  # legend title 
+  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
   #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +

### 2024-12-16 paper.qmd
paper

@@
 ---
-title: "Incorporating spatial distribution at age improves predictive ability of SDMs"
+title: "Advancing spatially explicit fisheries management with age-structured species distribution models"
 format: pdf
+date: "`r Sys.Date()`"
+author: 
+  - name: Eric & Nick
 editor: visual
 bibliography: references.bib
+csl: fish-and-fisheries.csl
 ---
@@
 Here, we develop an age-structured SDM framework using data from the west coast of the USA, and concentrate on two species with distinct life histories: North Pacific hake (*Merluccius productus*) and sablefish (*Anoplopoma fimbria*).
-Our framework incorporates spatiotemporal models to predict age-specific distributions.
-We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlihght that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes.
-Output from our models can be used to predict the spatial age composition, which may be particularly useful for large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts).
-Our approach is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.
+Our framework uses spatiotemporal models to predict age-specific distributions.
+We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes.
+Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts).
+Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk, helping increase the efficiency of fisheries.
+Our framework is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.
 
@@
 Much of the previous research around the world has demonstrated that age structure is a fundamental characteristic of fish populations, influencing spatial distribution and abundance due to mechanisms like ontogenetic shifts, recruitment variability, and selective fishing pressures [@rijnsdorp2018].
+Recent work has also highlighted that age-based SDMs can improve understanding of environmental drivers on commercially valuable populations [@ono2024].
 Ignoring age-specific information in SDMs may lead to incomplete or inaccurate predictions, especially for species with strong age-related spatial gradients.
@@
 For each species - age combination, we constructed a unique spatiotemporal model fit to all years except the last available year for each species.
-Individual fish were aggregated at the haul level, allowing us to model counts at age.
-We initially explored modeling counts with a Poisson or Negative Binomial distribution, but as the distributions of numbers at age are dominated by 0s, these models failed to converge.
-Instead, we summarized the total number of fish sampled and number of fish sampled of the focal age; this allowed counts to be modeled as a binomial family (logit link).
-Second, we constructed a spatiotemporal model as an extension of Generalized Linear Mixed Models (GLMMs) such that the prediction in location $s$ and time $t$ can be written as
+We adopted this approach instead of modeling ages simultaneously [@ono2024], because this framework allows greater flexibility (e.g.
+unique autocorrelation and spatial variance parameters by age).
+Individual fish were aggregated at the haul level, allowing us to model counts at age with a Poisson distribution.
+We constructed a spatiotemporal model as an extension of Generalized Linear Mixed Models (GLMMs) such that the prediction in location $s$ and time $t$ can be written as
 
@@
 \begin{aligned}
-logit \left( p_{\boldsymbol{s},t} \right) &= \boldsymbol{\beta_{t}} + \omega_{\boldsymbol{s}} + \delta_{\boldsymbol{s},t}
+log \left( u_{\boldsymbol{s},t} \right) &= \boldsymbol{\beta_{t}} + \omega_{\boldsymbol{s}} + \delta_{\boldsymbol{s},t}
 \end{aligned}
@@
 Models were evaluated for convergence (positive-definite Hessian matrix, and a maximum absolute log likelihood gradient \< 0.001) and residuals diagnostics were evaluated with the DHARMa package [@dharma2022].
+Code and data to replicate our analysis is publicly available on Github, <https://github.com/ericward-noaa/spatial-age-models>.
 
@@
 
-As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution of ages.
+As a first validation, we leveraged the natural recursive element of our data to quantify the ability of our models to predict the future distribution by age.
 For each of the species - age models constructed above, we made predictions to the survey locations in the following year (e.g.
-age 3 hake in years 2007 -- 2018 used to predict the distribution of age 4 hake in 2008 -- 2019).
-We related the predictions $\hat{p}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{p}_{\boldsymbol{s},t}$.
+age 3 hake in years 2007 -- 2018 was used to predict the distribution of age 4 hake in 2008 -- 2019).
+We related the predictions $\hat{u}_{\boldsymbol{s},t}$ to observations by multiplying predictions by the total number of fish sampled for ageing, $\hat{n}_{s,t} = N_{s,t} \cdot \hat{u}_{\boldsymbol{s},t}$.
 We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$.
 The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival.
-As a diagnostic check and to understand whether there was a potential time component, we fit a second model as a GLMM with R package glmmTMB; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.
+As a diagnostic check and to understand whether there was potential temporal patterns in the response, we fit a second model as a GLMM with R package glmmTMB [@brooks2017]; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.
 
@@
 
-To illustrate potential benefits of this kind of modeling, we used the above approach to generate spatiotemporal predictions of age-1 sablefish (from age-0 sablefish the year prior).
-\[Need some text here on background why sablefish as bycatch is important\].
+To illustrate potential benefits to fisheries management, we used output from our models to generate spatiotemporal predictions of age-1 sablefish (from age-0 sablefish the year prior).
+\[Need some text here on background why sablefish as bycatch is important.
+In 2017 for example, the at-sea whiting fishery had 50 mt of sablefish quota set aside in the area north of 36 deg N, but the fleet caught more than 3x as much\].
 Gridded predictions from the Poisson GLM represent the expected number of fish per unit of effort in each location.
-To turn these into a more useful measure of risk, we calculate the total number of age-1 sablefish per unit of effort within a given radius of major fishing ports on the US west coast.
+To turn these into a potentially more useful measure of risk, we calculate the predicted total age-1 sablefish catch per unit of effort (CPUE) within a given radius of major fishing ports on the US west coast.
 For consistency with previous work \[TODO: cite ESR\] we used a radius of 232km.
+These predictions of port-level CPUE were then summarized as annual time series (1-step ahead forecasts).
 
@@
 Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another).
-With the exception of a strong positive correlation between age-0 and age-1 sablefish (\$\$)
+With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude (@fig-glm-coefficients-time).
 
-Figure @fig-bycatch-risk aggregates occurrence probabilities for age 0 - 2 sablefish, potentially useful to identify hotspots
-
-Similarly Figure @fig-age5-sablefish shows occurrence probabilities for age 5 sablefish
+Our forecasts of age-1 sablefish CPUE illustrates the variability of potential bycatch risk in space and time.
+In most years for example, age-1 CPUE is lowest in ports in the state of California (Eureka, Fort Bragg, Morro Bay) and highest in several ports in the state of Oregon (Astoria, Coos Bay, @fig-bycatch-risk).
+These results also illustrate the high temporal variability in age-1 CPUE, with peaks occuring in 2009, 2014, 2017, and 2021, @fig-bycatch-risk).
+During peak years, there is also variability among ports – forecasts of age-1 CPUE were similar for Astoria and Coos Bay in 2021 for instance, but the potential bycatch risk near Astoria was higher in 2017 (@fig-bycatch-risk).
 
@@
 
-#### Limitations
+In this study, we developed an age-structured species distribution modeling (SDM) framework to explore spatial age composition and predict age-specific distributions of two commercially important fish species with different life history characteristics: North Pacific hake and sablefish.
+Our approach integrates spatiotemporal dynamics, age structure, and forward-predictive capabilities, providing a useful tool for fisheries management.
+By examining age-specific patterns in spatial distributions and validating predictive skill, we highlight the potential utility of incorporating age structure into SDMs for advancing spatially explicit fisheries management.
 
--   not predicting abundance (N) -- just proportions
--   We constructed single-age models, and this could be done in a multivariate (multinomial) framework -- but our approach allows for greater flexibility in letting relationships change by age.
-Tradeoff is that it doesn't impose sum to 1 constraint, binomials are independent
+#### *Age-Structured Spatial Patterns*
+
+Results from our age-structured spatial models underscore the importance of accounting for age in SDMs.
+The spatial patterns observed for both hake and sablefish were consistent with their distinct life histories, supporting the idea that ontogenetic shifts significantly influence fish distributions.
+Younger age classes of hake and sablefish exhibited broader distributions generally in shallower areas, while older individuals were more concentrated offshore (sablefish in deeper waters, hake in more northern regions).
+The deeper distribution of older sablefish reflects their preference for benthic habitats, while the observed north-south gradient in older hake aligns with known migratory behaviors influenced by spawning and feeding grounds [@agostini2006].
+
+#### *Predictive skill and cohort tracking*
+
+Our validation analyses revealed that the predictive skill of our framework increases with age, particularly for sablefish.
+This result is intuitive given that older age classes generally exhibit more stable spatial patterns due to reduced recruitment variability and more predictable habitat preferences.
+In contrast, the lower predictive skill for younger age classes reflects their higher spatiotemporal variability, likely driven by recruitment pulses and early-life survival dynamics.
+Tracking cohorts through time using our framework proved effective in identifying strong year classes, such as the 2010 and 2014 hake cohorts and the 2013, 2016, and 2020 sablefish cohorts.
+Identifying these cohorts may be useful to stock assessment and management – for instance, understanding the spatial distribution of strong cohorts could be used to inform spatial harvest allocation.
+Additionally, the ability to project spatial age composition provides a tool for assessing how environmental changes, such as ocean warming or habitat shifts, may impact the distribution of key cohorts over time.
+
+#### *Managing bycatch risk*
+
+Our case study on age-1 sablefish demonstrates the practical utility of our modeling for mitigating bycatch risk.
+Forecasts of CPUE for age-1 sablefish revealed significant temporal and spatial variability, with higher bycatch risk near Oregon ports in certain years.
+Importantly, these forecasts could be provided to industry partners and fisheries managers 6 – 12 months before fishing occurs.
+Such information could help fishers avoid areas associated with high bycatch risk, or be used to adjust the spatial distribution of quota in certain years.
+The observed peaks in age-1 sablefish CPUE during certain years suggest the influence of recruitment pulses, which are thought to be tied to environmental drivers [@tolimieri2018].
+Future efforts could also incorporating environmental covariates to enhance its predictive accuracy and further support climate-resilient fisheries management.
+
+#### Need for reliable age data
+
+Perhaps the biggest challenge in implementing our modeling framework in the future is that these models depend on reliable age data being collected.
+While the WCGBTS provides a robust dataset, gaps in sampling (e.g., the absence of a 2020 survey due to the COVID-19 pandemic) and variability in otolith aging efforts over time could introduce uncertainty.
+The costs of ship-based surveys, such as the WCGBTS survey, have increased around the world, and numerous efforts are being implemented to make these programs more efficient [@ices2020].
+While age samples are more expensive to collect than length information, there is also increased precision associated with age data.
+
+#### Future directions
+
+Our age-structured SDM framework represents a significant advance in spatial modeling, offering novel insights into the dynamics of fish populations and their age-specific distributions.
+From a management perspective, age-structured SDMs offer a powerful tool for addressing critical challenges, including bycatch mitigation, spatial planning, and sustainable harvesting.
+By providing high-resolution predictions of spatial age composition, these models enable managers to align fishing effort with ecological and economic goals, ensuring the long-term sustainability of marine resources.
+Future research could extend this framework by incorporating environmental covariates, such as temperature or oxygen, to account for dynamic habitat preferences, or fisheries removals to account for local depletion [@ono2016].
+Additionally, exploring the potential for multi-species models could provide insights into ecosystem-level interactions and their influence on spatial distributions.
 
@@
 
-```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial composition for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014) and columns represent predictions for ages 1 - 5.
+```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014) and columns represent predictions for ages 1 - 5.
 Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
@@
 
-```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial composition for sablefish; rows represent strong cohorts in our dataset (2013, 2016, 2019) and columns represent predictions for ages 0 - 4.
+```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent strong cohorts in our dataset (2013, 2016, 2019) and columns represent predictions for ages 0 - 4.
 Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
@@
 
-```{r label="fig-bycatch-risk", fig.cap="Estimated probability of occurrence for age 0 -- 2 sablefish.", echo=FALSE}
+```{r label="fig-bycatch-risk", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for age-1 sablefish and port-level forecasts of age-1 CPUE within a radius of 232 km.", echo=FALSE}
 knitr::include_graphics("plots/sablefish_spatial_risk.png")
-```
-
-\newpage
-
-```{r label="fig-age5-sablefish", fig.cap="Estimated probability of occurrence for age 5 sablefish.", echo=FALSE}
-knitr::include_graphics("plots/sablefish_spatial_age5.png")
 ```
@@
 
-```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial composition for Pacific hake; rows represent years and columns represent predictions for ages 1 - 5.", echo=FALSE}
+```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent years and columns represent predictions for ages 1 - 5.", echo=FALSE}
 knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
@@
 
-```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial composition for sablefish; rows represent years and columns represent predictions for ages 0 - 4.", echo=FALSE}
+```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent years and columns represent predictions for ages 0 - 4.", echo=FALSE}
 
 knitr::include_graphics("plots/spatial_composition_sablefish.png")
+```
+
+\newpage
+
+```{r label="fig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year.
+", echo=FALSE}
+knitr::include_graphics("plots/glm_coefficients_time.png")
 ```

### 2025-01-10 02_recursive_model_poisson.R
updated code and figs

@@ -1,219 +1,0 @@
-library(nwfscSurvey)
-library(sdmTMB)
-library(tidyverse)
-library(ggplot2)
-library(DHARMa)
-library(surveyjoin)
-
-spp_name <- c("Pacific hake", "sablefish")[1]
-
-if(spp_name == "Pacific hake") {
-  min_age <- 1 # not many age 0s consistently sampled
-  max_age <- 6
-  min_year <- 2007
-  max_year <- 2019
-} 
-if(spp_name == "sablefish") {
-  min_age <- 0
-  max_age <- 10
-  min_year <- 2003
-  max_year <- 2023
-}
-all_years <- min_year:max_year
-
-bio <- pull_bio(common_name = spp_name, survey="NWFSC.Combo")
-names(bio) <- tolower(names(bio))
-haul <- pull_haul(survey="NWFSC.Combo")
-names(haul) <- tolower(names(haul))
-
-# filter ages to be in core range
-d <- bio
-d <- dplyr::filter(d, sex == "F", !is.na(age), 
-                   year >= min_year,
-                   year <= max_year)
-
-# group data by age and year 
-grouped <- dplyr::group_by(d, age, year, trawl_id) |>
-  dplyr::summarise(n = n(), # n is the number of observations of this age in this year
-                   present = ifelse(n > 1, 1, 0), # not used
-                   lat = latitude_dd[1], 
-                   lon = longitude_dd[1]) |>
-  dplyr::ungroup()
-grouped_total <- dplyr::group_by(d, trawl_id) |>
-  dplyr::summarise(n_total = n()) |>
-  dplyr::ungroup()
-grouped <- dplyr::left_join(grouped, grouped_total)
-
-# add trawl data -- after expanding across years and ages
-grid <- expand.grid(trawl_id = unique(grouped$trawl_id), 
-                    age = unique(grouped$age))
-yrs <- dplyr::group_by(grouped, trawl_id) |>
-  dplyr::summarise(year = year[1])
-grid <- dplyr::left_join(grid, yrs)
-locs <- dplyr::group_by(grouped, trawl_id) |>
-  dplyr::summarize(lat = lat[1], lon = lon[1])
-
-d <- dplyr::left_join(grid[,c("trawl_id","age")], grouped, by = c("trawl_id","age")) 
-d$n[which(is.na(d$n))] <- 0
-d$n_total[which(is.na(d$n_total))] <- 0
-
[…]

### 2025-01-10 03_poisson_regression.R

@@ -69,14 +69,14 @@ for(spp in 1:2) {
   }
 }
 
-ggplot(all_output, aes(age, mean, col = species)) + 
-  geom_hline(aes(yintercept=0), col="red", alpha=0.3) + 
-  geom_pointrange(aes(ymin = lower, ymax = upper), position=position_dodge(0.4)) + 
+ggplot(all_output, aes(age, exp(mean), col = species)) + 
+  geom_hline(aes(yintercept=1), col="red", alpha=0.3) + 
+  geom_pointrange(aes(ymin = exp(lower), ymax = exp(upper)), position=position_dodge(0.4)) + 
   geom_point(position=position_dodge(0.4)) + 
-  scale_color_viridis_d(option="magma", begin=0.2, end = 0.8) +
+  scale_color_viridis_d(option="magma", begin=0.2, end = 0.8, name="Species") +
   xlab("Age") + 
   ylab("Estimated slope") + 
-  coord_flip() + theme_bw()
+  theme_bw()
 
 ggsave("plots/glm_coefficients.png", width=6, height = 5)
 

### 2025-01-10 04_spatial_anomaly_plots.R

@@ -55,32 +55,35 @@ utm_zone10 <- 3157
 coast_proj <- sf::st_transform(coast, crs = utm_zone10)
 
 # Define the new colors
-land_color <- "#E0CDA9"  # beige for land
-ocean_color <- "#D3EAF2"  # grayish blue for ocean
+land_color <- "wheat3"  # beige for land
+ocean_color <- "grey80"  # grayish blue for ocean
 
+library(scales)
 # Plot coast with custom land and ocean colors
 ggplot(coast_proj) + 
   geom_point(data = hake, aes(x = X * 1000, y = Y * 1000, col = omega_s), size=0.02) + 
   geom_sf(fill = land_color) +  # Set land color
-  scale_color_gradient2(name = "Anomaly") +  # legend title 
+  scale_color_gradient2(name = "Anomaly",low = scales::muted("blue"), mid = "white", high = scales::muted("red"), midpoint = 0) +  # legend title 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  
         strip.text = element_text(color = "black"),  # Set facet label text color to black
-        strip.background = element_rect(fill = "white")) +  # Set facet label background to white
+        strip.background = element_rect(fill = "white"),
+        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Set facet label background to white
   facet_wrap(~age, nrow = 1)
-ggsave("plots/hake_spatial_anomalies.png", width = 7, height = 4)  
+ggsave("plots/hake_spatial_anomalies.png", width = 7, height = 5)  
   
 
 ggplot(coast_proj) + 
   geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = omega_s), size=0.02) + 
   geom_sf(fill = land_color) +  # Set land color
-  scale_color_gradient2(name = "Anomaly") +  # legend title 
+  scale_color_gradient2(name = "Anomaly",low = scales::muted("blue"), mid = "white", high = scales::muted("red"), midpoint = 0) +  # legend title 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  
         strip.text = element_text(color = "black"),  # Set facet label text color to black
-        strip.background = element_rect(fill = "white")) +  # Set facet label background to white
+        strip.background = element_rect(fill = "white"),
[…]

### 2025-01-10 05_spatial_bycatch_risk_sablefish.R

@@ -33,13 +33,13 @@ utm_zone10 <- 3157
 coast_proj <- sf::st_transform(coast, crs = utm_zone10)
 
 # Define the new colors
-land_color <- "#E0CDA9"  # beige for land
-ocean_color <- "#D3EAF2"  # grayish blue for ocean
+land_color <- "wheat3"  # beige for land
+ocean_color <- "grey80"  # grayish blue for ocean
 
 p1 <- ggplot(coast_proj) + 
-  geom_point(data = sablefish, aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
-  #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+  geom_point(data = dplyr::filter(sablefish, p <= quantile(sablefish$p,0.999)), aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
+  #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
+  scale_color_gradient2(name = "CPUE", low="white", high=scales::muted("red"), trans="sqrt") + 
   geom_sf(fill = land_color) +  # Set land color
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +

### 2025-01-10 06_plot_spatial_compositions.R

@@ -47,15 +47,15 @@ utm_zone10 <- 3157
 coast_proj <- sf::st_transform(coast, crs = utm_zone10)
 
 # Define the new colors
-land_color <- "#E0CDA9"  # beige for land
-ocean_color <- "#D3EAF2"  # grayish blue for ocean
+land_color <- "wheat3"  # beige for land
+ocean_color <- "grey80"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 g <- ggplot(coast_proj) + 
-  geom_tile(data = pred_all, aes(x = X * 1000, y = Y * 1000, col = p)) + 
+  geom_tile(data = dplyr::filter(pred_all, p <= quantile(pred_all$p,0.999)), aes(x = X * 1000, y = Y * 1000, col = p)) + 
   geom_sf(fill = land_color) +  # Set land color
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
-  #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+  #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
+  scale_color_gradient2(name = "CPUE", trans = "sqrt", low = "white", high = scales::muted("red")) + 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  
@@ -80,10 +80,9 @@ if(spp_name == "Pacific hake") {
 
 
 g <- ggplot(coast_proj) + 
-  geom_point(data = pred_subset, aes(x = X * 1000, y = Y * 1000, col = p),size=0.05) + 
+  geom_point(data = dplyr::filter(pred_subset, p <= quantile(pred_subset$p,0.999)), aes(x = X * 1000, y = Y * 1000, col = p),size=0.05) + 
   geom_sf(fill = land_color) +  # Set land color
-  scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
-  #scale_color_gradient2(name = "Centered \n Pr(occurrence)") + 
+  scale_color_gradient2(name = "CPUE", high = scales::muted("red")) +  # legend title 
   theme_light() + 
   labs(x = "Longitude", y = "Latitude") +
   theme(panel.background = element_rect(fill = ocean_color),  

### 2025-01-11 06_plot_spatial_compositions.R
Update 06_plot_spatial_compositions.R

@@ -1,7 +1,7 @@
 library(tidyverse)
 library(ggplot2)
 
-spp_name <- c("Pacific hake", "sablefish")[2]
+spp_name <- c("Pacific hake", "sablefish")[1]
 
 if(spp_name == "Pacific hake") {
   min_age <- 1 # not many age 0s consistently sampled
@@ -52,7 +52,7 @@ ocean_color <- "grey80"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 g <- ggplot(coast_proj) + 
-  geom_tile(data = dplyr::filter(pred_all, p <= quantile(pred_all$p,0.999)), aes(x = X * 1000, y = Y * 1000, col = p)) + 
+  geom_tile(data = dplyr::filter(pred_all, p <= quantile(pred_all$p,0.99)), aes(x = X * 1000, y = Y * 1000, col = p)) + 
   geom_sf(fill = land_color) +  # Set land color
   #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
   scale_color_gradient2(name = "CPUE", trans = "sqrt", low = "white", high = scales::muted("red")) + 
@@ -80,7 +80,7 @@ if(spp_name == "Pacific hake") {
 
 
 g <- ggplot(coast_proj) + 
-  geom_point(data = dplyr::filter(pred_subset, p <= quantile(pred_subset$p,0.999)), aes(x = X * 1000, y = Y * 1000, col = p),size=0.05) + 
+  geom_point(data = dplyr::filter(pred_subset, p <= quantile(pred_subset$p,0.99)), aes(x = X * 1000, y = Y * 1000, col = p),size=0.05) + 
   geom_sf(fill = land_color) +  # Set land color
   scale_color_gradient2(name = "CPUE", high = scales::muted("red")) +  # legend title 
   theme_light() + 

### 2025-01-11 05_spatial_bycatch_risk_sablefish.R
Update 05_spatial_bycatch_risk_sablefish.R

@@ -37,7 +37,7 @@ land_color <- "wheat3"  # beige for land
 ocean_color <- "grey80"  # grayish blue for ocean
 
 p1 <- ggplot(coast_proj) + 
-  geom_point(data = dplyr::filter(sablefish, p <= quantile(sablefish$p,0.999)), aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
+  geom_point(data = dplyr::filter(sablefish, p <= quantile(sablefish$p,0.99)), aes(x = X * 1000, y = Y * 1000, col = p), size = 0.02) + 
   #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
   scale_color_gradient2(name = "CPUE", low="white", high=scales::muted("red"), trans="sqrt") + 
   geom_sf(fill = land_color) +  # Set land color

### 2025-01-11 paper.qmd
Update paper.qmd

@@
 Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts).
-Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk, helping increase the efficiency of fisheries.
+Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk in space, helping increase the efficiency of fisheries.
 Our framework is applicable to a wide range of survey types and platforms around the world, and supports the development of spatially explicit management strategies and optimal allocation of fishing effort.
@@
 These models have informed management decisions for conservation and resource allocation, especially in the face of changing environmental conditions and human impacts, such as fishing pressure [@guisan2005; @elith2009].
-Recent advances in SDM applications in the marine environment have incorporated external drivers such as temperature [@fredston2021], habitat [@phillips2017; @laman2018].
+Recent advances in SDM applications in the marine environment have incorporated external drivers such as temperature [@fredston2021] and habitat features [@phillips2017; @laman2018].
 A second use of these models has been to predict temporal trends in species biomass from spatiotemporal fisheries independent survey data [@thorson2015].
@@
 Additionally, recruitment pulses and migration patterns vary by age, influencing the spatial distribution of younger versus older fish.
-Fishing selectivity further accentuates these patterns by disproportionately removing older age classes, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013].
-By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relationships.
+Fishing selectivity often further accentuates these patterns by disproportionately removing larger, older fish, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013].
+By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relati
 
@@
 The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data).
-Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyse.
-Like many other surveys around the world, no WCGBTS survey occured in 2020 because of the Covid-19 pandemic.
+Importantly, the random stratified sampling design, spatial and seasonal coverage, effort, and gears have remained relatively constant within the period we analyze.
+Like many other surveys around the world, no WCGBTS survey occurred in 2020 because of the Covid-19 pandemic.
 Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish.
@@
 We quantified the relationship between predictions and observations by fitting a simple Poisson GLM, where counts of fish of age $a+1$ in year $t+1$ were treated as the response and $log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)$.
-The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$, and can roughly be interpreted as an estimate of relative survival.
+The exponentiated slope parameter $exp(\beta_{1})$ represents a change in expected counts that would be expected from a 1-unit change in $log(\hat{n}_{s,t})$.
 As a diagnostic check and to understand whether there was potential temporal patterns in the response, we fit a second model as a GLMM with R package glmmTMB [@brooks2017]; this second model was identical to our Poisson GLM with random effects in $\beta_{0}$ and $\beta_{1}$, where year was used as the grouping variable.
@@
 
-When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range (@fig-spatial-parameters).
+When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range with increasing age (@fig-spatial-parameters).
 Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, meaning that spatial fields estimated for older fish are less variable.
@@
 ```{r label="fig-glm-coefficients", fig.cap="Estimated coefficients relating predicted densities of age a fish in year t to observed numbers the following year.
-The coefficients are interpreted with a multiplicate effect, such that a coefficient of 0.5 translates to an effect of exp(0.5) = 65% increase on the expected count in the future.", echo=FALSE}
+The coefficients and 95% confidence intervals (vertical lines) are estimated in log space and presented here in normal space.", echo=FALSE}
 knitr::include_graphics("plots/glm_coefficients.png")

### 2025-01-16 paper.qmd
Update paper.qmd

@@
 These results also illustrate the high temporal variability in age-1 CPUE, with peaks occuring in 2009, 2014, 2017, and 2021, @fig-bycatch-risk).
-During peak years, there is also variability among ports – forecasts of age-1 CPUE were similar for Astoria and Coos Bay in 2021 for instance, but the potential bycatch risk near Astoria was higher in 2017 (@fig-bycatch-risk).
+During peak years, there is also variability among ports -- forecasts of age-1 CPUE were similar for Astoria and Coos Bay in 2021 for instance, but the potential bycatch risk near Astoria was higher in 2017 (@fig-bycatch-risk).
 
@@
 Tracking cohorts through time using our framework proved effective in identifying strong year classes, such as the 2010 and 2014 hake cohorts and the 2013, 2016, and 2020 sablefish cohorts.
-Identifying these cohorts may be useful to stock assessment and management – for instance, understanding the spatial distribution of strong cohorts could be used to inform spatial harvest allocation.
+Identifying these cohorts may be useful to stock assessment and management -- for instance, understanding the spatial distribution of strong cohorts could be used to inform spatial harvest allocation.
 Additionally, the ability to project spatial age composition provides a tool for assessing how environmental changes, such as ocean warming or habitat shifts, may impact the distribution of key cohorts over time.
@@
 Forecasts of CPUE for age-1 sablefish revealed significant temporal and spatial variability, with higher bycatch risk near Oregon ports in certain years.
-Importantly, these forecasts could be provided to industry partners and fisheries managers 6 – 12 months before fishing occurs.
+Importantly, these forecasts could be provided to industry partners and fisheries managers 6 -- 12 months before fishing occurs.
 Such information could help fishers avoid areas associated with high bycatch risk, or be used to adjust the spatial distribution of quota in certain years.

### 2025-01-16 06_plot_spatial_compositions.R
new figures
adding figures combining distribution plots onto one map per age group or following one cohort through time. Age class = cohort. Still need to clean up color scheme.

@@ -52,7 +52,9 @@ ocean_color <- "grey80"  # grayish blue for ocean
 
 # Plot coast with custom land and ocean colors
 g <- ggplot(coast_proj) + 
-  geom_tile(data = dplyr::filter(pred_all, p <= quantile(pred_all$p,0.99)), aes(x = X * 1000, y = Y * 1000, col = p)) + 
+  geom_tile(data = dplyr::filter(pred_all, 
+                          p <= quantile(pred_all$p,0.99)), 
+                          aes(x = X * 1000, y = Y * 1000, col = p)) + 
   geom_sf(fill = land_color) +  # Set land color
   #scale_color_viridis(option="magma", begin = 0.2, end = 0.8, name = "CPUE", trans="sqrt") +  # legend title 
   scale_color_gradient2(name = "CPUE", trans = "sqrt", low = "white", high = scales::muted("red")) + 
@@ -63,6 +65,8 @@ g <- ggplot(coast_proj) +
         strip.background = element_rect(fill = "white"),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Set facet label background to white
   facet_grid(age ~ year)
+
+g
 
 ggsave(g, filename = paste0("plots/spatial_composition_",spp_name,".png"), height = 7, width = 7)
 

### 2025-01-16 07_plots_nt.R

@@ -1,0 +1,147 @@
+library(sdmTMB)
+# library(sdmTMBextra)
+library(INLA)
+library(sf)
+library(tidyverse)
+library(ggplot2)
+library(rnaturalearth)
+library(rnaturalearthdata)
+library(rasterVis)
+library(readxl)
+library(Matrix)
+library(ggpubr)
+
+home_dir = getwd()
+fig_dir = paste0(home_dir,"/plots/")
+
+spp_name <- c("Pacific hake", "sablefish")[2]
+
+if(spp_name == "Pacific hake") {
+  min_age <- 1 # not many age 0s consistently sampled
+  max_age <- 6
+  years <- 2003:2019
+  ages <- seq(min_age, min_age+4)
+} 
+if(spp_name == "sablefish") {
+  min_age <- 0
+  max_age <- 10
+  years <- 2003:2023
+  ages <- seq(min_age, min_age+4)
+}
+
+for(a in min_age:(max_age - 1)) {
+  pred <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+  pred <- dplyr::select(pred, lon, lat, X, Y, year, est) |>
+    dplyr::mutate(age = a) |>
+    dplyr::filter(year %in% years)
+  if(a == min_age) {
+    pred_all <- pred
+  } else {
+    pred_all <- rbind(pred_all, pred)
+  }
+}
+
+# now convert the 'est' to normalized probabilities across ages
+pred_all <- dplyr::mutate(pred_all, p = exp(est)/(1+exp(est))) |>
+  #dplyr::group_by(X,Y,year) |>
+  #dplyr::mutate(p_norm = p / sum(p)) |>
+  dplyr::filter(age %in% ages)
+
+map_data <- rnaturalearth::ne_countries(
+  scale = "medium",
+  returnclass = "sf", country = "united states of america")
+# Crop the polygon for plotting and efficiency:
+# st_bbox(map_data) # find the rough coordinates
+coast <- suppressWarnings(suppressMessages(
+  sf::st_crop(map_data,
+              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
+
+utm_zone10 <- 3157
+coast_proj <- sf::st_transform(coast, crs = utm_zone10)
+
+# Define the new colors
+land_color  <- "wheat3"  # beige for land
+ocean_color <- "grey80"  # grayish blue for ocean
+
+# NT Plot ######################################################################
+axis.mod = 3
+max.year = 2023
[…]

### 2025-01-17 07_plots_nt.R
more maps

@@ -42,34 +42,19 @@ for(a in min_age:(max_age - 1)) {
 }
 
 # now convert the 'est' to normalized probabilities across ages
-pred_all <- dplyr::mutate(pred_all, p = exp(est)/(1+exp(est))) |>
+pred_all <- dplyr::mutate(pred_all, p = exp(est)) |>
   #dplyr::group_by(X,Y,year) |>
   #dplyr::mutate(p_norm = p / sum(p)) |>
   dplyr::filter(age %in% ages)
 
-map_data <- rnaturalearth::ne_countries(
-  scale = "medium",
-  returnclass = "sf", country = "united states of america")
-# Crop the polygon for plotting and efficiency:
-# st_bbox(map_data) # find the rough coordinates
-coast <- suppressWarnings(suppressMessages(
-  sf::st_crop(map_data,
-              c(xmin = -132, ymin = 30, xmax = -117, ymax = 50))))
-
-utm_zone10 <- 3157
-coast_proj <- sf::st_transform(coast, crs = utm_zone10)
-
-# Define the new colors
-land_color  <- "wheat3"  # beige for land
-ocean_color <- "grey80"  # grayish blue for ocean
-
 # NT Plot ######################################################################
 axis.mod = 3
-max.year = 2023
+max_year = max.year = 2023
 pred_all$rec_year = pred_all$year - pred_all$age
 pred_all$yrs = axis.mod*(pred_all$year-max.year)
 pred_all$lon2 = pred_all$lon + pred_all$yrs
 xmin = floor( min(pred_all$lon2))
+pred_all$p <= quantile(pred_all$p,0.99)
 
 # background map info ##########################################################
 states <- map_data("state")
@@ -78,15 +63,19 @@ west_coast <- subset(states, region %in% c("california", "oregon", "washington")
 
 # plot function
 dist_maps <- function(dfile, xlim = c(-185, -117)) {
-
+  
   prob_map <- ggplot(data=west_coast, aes(x = long, y = lat), 
                    fill = grey(0.9), color = "black") +
     geom_polygon(aes(x = long, y = lat, group=group), 
                fill = grey(0.9), color = "black") +
-    geom_point(data=dfile, aes(lon2, lat, color=p), size=0.01) +
+    geom_point(data=dplyr::filter(dfile, 
[…]

### 2025-01-17 07_plots_nt.R
figure updates

@@ -62,20 +62,46 @@ west_coast <- subset(states, region %in% c("california", "oregon", "washington")
 ################################################################################
 
 # plot function
-dist_maps <- function(dfile, xlim = c(-185, -117)) {
+dist_maps <- function(dfile, 
+                      xlim = c(-185, -117), 
+                      midp = NA, 
+                      trans = 'identity',
+                      scalebiomass = T) {
+  # data preparation
+  # dfile = pred_all[pred_all$age==1,]
+  dfile = dplyr::filter(dfile, p <= quantile(dfile$p,0.99))
   
-  prob_map <- ggplot(data=west_coast, aes(x = long, y = lat), 
-                   fill = grey(0.9), color = "black") +
+  # scale data to 0-1 for each year 
+  
+  if(scalebiomass = TRUE){
+    dmean = dfile %>%
+            group_by(year) %>%
+            summarise(maxp = max(p))
+    #slow
+    dfile = left_join(dfile, dmean)
+    dfile$p = dfile$p/dfile$maxp
+  }
+  
+  
+  # set midpoing for color ramp
+  if(is.na(midp)==T){midp = 0.1*max(dfile$p)}
+  # begin plotting
+  plotx <- 
+    ggplot(data=west_coast, aes(x = long, y = lat), 
+                 fill = grey(0.9), color = "black") +
     geom_polygon(aes(x = long, y = lat, group=group), 
                fill = grey(0.9), color = "black") +
-    geom_point(data=dplyr::filter(dfile, 
-                                  p <= quantile(dfile$p,0.99)) , 
+    geom_point(data=dfile , 
                aes(lon2, lat, color=p), size=0.01) +
     # might need to adjust here based on the scale of the data, esp 'trans' term.
     # scale_color_viridis_c(option = 'turbo', direction = 1,  
     #                     name='CPUE') +
-    scale_color_gradient2(name = "CPUE", trans = "identity", 
-                          low = "white", high = scales::muted("red")) + 
+    scale_color_gradient2(name = "CPUE", 
+                          transform = trans, 
+                          low = scales::muted("black"),
+                          midpoint = midp,
+                          #high = 'red')+
+                          high = scales::muted("red")) + 
     coord_fixed(xlim=xlim,ylim=c(32,48),ratio=1.3) +
     # might need to adjust x-axes to plot all data
[…]

### 2025-01-17 07_plots_nt.R
more figure updates

@@ -14,7 +14,7 @@ library(ggpubr)
 home_dir = getwd()
 fig_dir = paste0(home_dir,"/plots/")
 
-spp_name <- c("Pacific hake", "sablefish")[2]
+spp_name <- c("Pacific hake", "sablefish")[1]
 
 if(spp_name == "Pacific hake") {
   min_age <- 1 # not many age 0s consistently sampled
@@ -73,11 +73,11 @@ dist_maps <- function(dfile,
   
   # scale data to 0-1 for each year 
   
-  if(scalebiomass = TRUE){
+  if(scalebiomass == TRUE){
     dmean = dfile %>%
             group_by(year) %>%
-            summarise(maxp = max(p))
-    #slow
+            summarise(maxp = max(p, na.rm = T))
+    # slow
     dfile = left_join(dfile, dmean)
     dfile$p = dfile$p/dfile$maxp
   }
@@ -117,18 +117,19 @@ dist_maps <- function(dfile,
     return(plotx)
 }
 
-m0 = dist_maps(pred_all[pred_all$age==0,])
-# print(m0)
 
+# print(m0)
 m1 = dist_maps(pred_all[pred_all$age==1,])
 m2 = dist_maps(pred_all[pred_all$age==2,])
 m3 = dist_maps(pred_all[pred_all$age==3,])
 m4 = dist_maps(pred_all[pred_all$age==4,])
 
-
-
-ggarrange(m0,m1,m2,m3,m4,
-          nrow = 5)
+if(spp_name=="sablefish"){
+  m0 = dist_maps(pred_all[pred_all$age==0,])
+  ggarrange(m0,m1,m2,m3,m4, nrow = 5)}
+if(spp_name=="Pacific hake"){
+  m5 = dist_maps(pred_all[pred_all$age==5,])
+  ggarrange(m1,m2,m3,m4, m5, nrow = 5)}
 
 ggsave( paste0(fig_dir ,"Map-",spp_name,"-year.png"), width = 6.2, height = 8)
 
@@ -140,15 +141,26 @@ if(spp_name == "sablefish"){
   y2008 = dist_maps(pred_all[pred_all$rec_year==2008,],xlim = c(-170, -117))
   y2010 = dist_maps(pred_all[pred_all$rec_year==2010,],xlim = c(-170, -117))
   y2016 = dist_maps(pred_all[pred_all$rec_year==2016,],xlim = c(-170, -117))
[…]

### 2025-01-17 07_plots_nt.R
plot updates

@@ -71,7 +71,7 @@ dist_maps <- function(dfile,
   # dfile = pred_all[pred_all$age==1,]
   dfile = dplyr::filter(dfile, p <= quantile(dfile$p,0.99))
   
-  # scale data to 0-1 for each year 
+  # scale data to 0-1 for each year ; if decided
   
   if(scalebiomass == TRUE){
     dmean = dfile %>%
@@ -119,16 +119,17 @@ dist_maps <- function(dfile,
 
 
 # print(m0)
-m1 = dist_maps(pred_all[pred_all$age==1,])
-m2 = dist_maps(pred_all[pred_all$age==2,])
-m3 = dist_maps(pred_all[pred_all$age==3,])
-m4 = dist_maps(pred_all[pred_all$age==4,])
+m1 = dist_maps(pred_all[pred_all$age==1,], scalebiomass = F)
+m2 = dist_maps(pred_all[pred_all$age==2,], scalebiomass = F)
+m3 = dist_maps(pred_all[pred_all$age==3,], scalebiomass = F)
+m4 = dist_maps(pred_all[pred_all$age==4,], scalebiomass = F)
 
 if(spp_name=="sablefish"){
-  m0 = dist_maps(pred_all[pred_all$age==0,])
+  m0 = dist_maps(pred_all[pred_all$age==0,], scalebiomass = F)
   ggarrange(m0,m1,m2,m3,m4, nrow = 5)}
+
 if(spp_name=="Pacific hake"){
-  m5 = dist_maps(pred_all[pred_all$age==5,])
+  m5 = dist_maps(pred_all[pred_all$age==5,], scalebiomass = F)
   ggarrange(m1,m2,m3,m4, m5, nrow = 5)}
 
 ggsave( paste0(fig_dir ,"Map-",spp_name,"-year.png"), width = 6.2, height = 8)
@@ -141,9 +142,9 @@ if(spp_name == "sablefish"){
   y2008 = dist_maps(pred_all[pred_all$rec_year==2008,],xlim = c(-170, -117))
   y2010 = dist_maps(pred_all[pred_all$rec_year==2010,],xlim = c(-170, -117))
   y2016 = dist_maps(pred_all[pred_all$rec_year==2016,],xlim = c(-170, -117))
-  #y2021 = dist_maps(pred_all[pred_all$rec_year==2021,],xlim = c(-170, -117))
[…]

### 2025-01-24 paper.qmd
Add files via upload

@@
 We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes.
-Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010 and 2014 cohorts; sablefish from 2013, 2016, and 2020 cohorts).
+Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010, 2014 and 2016 cohorts; sablefish from 2008, 2010, 2016, and 2021 cohorts).
 Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk in space, helping increase the efficiency of fisheries.
@@
 Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time.
-For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, @fig-hake-spatial-composition) and strong sablefish cohorts (2013, 2016, 2020, @fig-sablefish-spatial-composition).
+For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, 2016; @fig-hake-spatial-composition) and strong sablefish cohorts (2008, 2010, 2016, 2021, @fig-sablefish-spatial-composition).
 These predictions are also useful in revealing subtle differences between strong cohort years.
-For example, the 2009 cohort of Pacific hake (age 1 in 2010) were predominantly distributed in the northern part of the domain, while the 2013 cohort had hotspots along the entire US west coast.
-Similarly for sablefish, the 2016 cohort appeared to be more concentrated in the north, versus the more widely distributed 2013 cohort.
+For example, the 2014 cohort of Pacific hake had a larger fraction of age-1 individuals distributed in the northern part of the domain (along the coast of Washington state), while other cohorts had hotspots in the center and southern portion of the domain.
+Similarly for sablefish, the 2010 cohort also appeared to be more concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution.
 
@@
 In contrast, the lower predictive skill for younger age classes reflects their higher spatiotemporal variability, likely driven by recruitment pulses and early-life survival dynamics.
-Tracking cohorts through time using our framework proved effective in identifying strong year classes, such as the 2010 and 2014 hake cohorts and the 2013, 2016, and 2020 sablefish cohorts.
+Tracking cohorts through time using our framework proved effective in identifying strong year classes (Figs.
+3-4).
 Identifying these cohorts may be useful to stock assessment and management -- for instance, understanding the spatial distribution of strong cohorts could be used to inform spatial harvest allocation.
@@
 
-Our case study on age-1 sablefish demonstrates the practical utility of our modeling for mitigating bycatch risk.
+Our case study using age-1 sablefish demonstrates the practical utility of our modeling for mitigating bycatch risk.
 Forecasts of CPUE for age-1 sablefish revealed significant temporal and spatial variability, with higher bycatch risk near Oregon ports in certain years.
@@
 
-```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014) and columns represent predictions for ages 1 - 5.
+```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014, 2016).
+CPUE is standardized to 1.0 across cohorts for visualization purposes.
 Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
-knitr::include_graphics("plots/spatial_composition_subset_Pacific hake.png")
+knitr::include_graphics("plots/Map-Pacific hake-age-class.png")
 ```
@@
 
-```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent strong cohorts in our dataset (2013, 2016, 2019) and columns represent predictions for ages 0 - 4.
+```{r label="fig-sablefish-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; each row represents a strong cohorts in our dataset (2008, 2010, 2016, 2021).
+CPUE is standardized to 1.0 across cohorts for visualization purposes.
 Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
-knitr::include_graphics("plots/spatial_composition_subset_sablefish.png")
+knitr::include_graphics("plots/Map-sablefish-age-class.png")
 ```
@@
 
-```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent years and columns represent predictions for ages 1 - 5.", echo=FALSE}
-knitr::include_graphics("plots/spatial_composition_Pacific hake.png")
+```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5).
+CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
+knitr::include_graphics("plots/Map-Pacific hake-year.png")
 ```
@@
 
-```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent years and columns represent predictions for ages 0 - 4.", echo=FALSE}
+```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4).
+CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
 
-knitr::include_graphics("plots/spatial_composition_sablefish.png")
+knitr::include_graphics("plots/Map-sablefish-year.png")
 ```

### 2025-01-24 07_plots_nt.R

@@ -153,12 +153,12 @@ if(spp_name == "sablefish"){
 
 # hake
 if(spp_name == "Pacific hake"){
-  y2008 = dist_maps(pred_all[pred_all$rec_year==2008,],xlim = c(-170, -117))
-  y2010 = dist_maps(pred_all[pred_all$rec_year==2010,],xlim = c(-170, -117))
-  y2016 = dist_maps(pred_all[pred_all$rec_year==2016,],xlim = c(-170, -117))
+  y2008 = dist_maps(pred_all[pred_all$rec_year==2010,],xlim = c(-170, -117))
+  y2010 = dist_maps(pred_all[pred_all$rec_year==2014,],xlim = c(-170, -117))
+  y2014 = dist_maps(pred_all[pred_all$rec_year==2016,],xlim = c(-170, -117))
   #y2021 = dist_maps(pred_all[pred_all$rec_year==2021,],xlim = c(-170, -117))
   
-  ggarrange(y2008, y2010, y2016,nrow = 3)
+  ggarrange(y2008, y2010, y2014,nrow = 3)
   
   ggsave( paste0(fig_dir ,"Map-",spp_name,"-age-class.png"), width = 6.2, height = 6)
 }

### 2025-01-27 paper.qmd
minor updates renaming figures; preparing others

@@
 Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
-knitr::include_graphics("plots/Map-Pacific hake-age-class.png")
+knitr::include_graphics("plots/Pacific hake-follow-age-class.png")
 ```
@@
 Full predictions for all cohorts are in the Supplementary Information.", echo=FALSE}
-knitr::include_graphics("plots/Map-sablefish-age-class.png")
+knitr::include_graphics("plots/sablefish-follow-age-class.png")
 ```
@@
 CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
-knitr::include_graphics("plots/Map-Pacific hake-year.png")
+knitr::include_graphics("plots/Pacific hake-age-class-year.png")
 ```
@@
 
-knitr::include_graphics("plots/Map-sablefish-year.png")
+knitr::include_graphics("plots/sablefish-age-class-year.png")
 ```

### 2025-01-27 07_plots_nt.R

@@ -49,9 +49,9 @@ pred_all <- dplyr::mutate(pred_all, p = exp(est)) |>
 
 # NT Plot ######################################################################
 axis.mod = 3
-max_year = max.year = 2023
+max_year = 2023
 pred_all$rec_year = pred_all$year - pred_all$age
-pred_all$yrs = axis.mod*(pred_all$year-max.year)
+pred_all$yrs = axis.mod*(pred_all$year-max_year)
 pred_all$lon2 = pred_all$lon + pred_all$yrs
 xmin = floor( min(pred_all$lon2))
 pred_all$p <= quantile(pred_all$p,0.99)
@@ -81,8 +81,6 @@ dist_maps <- function(dfile,
     dfile = left_join(dfile, dmean)
     dfile$p = dfile$p/dfile$maxp
   }
-  
-  
   # set midpoing for color ramp
   if(is.na(midp)==T){midp = 0.1*max(dfile$p)}
   # begin plotting
@@ -117,9 +115,15 @@ dist_maps <- function(dfile,
     return(plotx)
 }
 
+# age-class each year ##########################################################
 
 # print(m0)
+# age 1 sablefish for figure 6 ####
+# combine later for fig 6
 m1 = dist_maps(pred_all[pred_all$age==1,], scalebiomass = F)
+ggsave( paste0(fig_dir ,spp_name,"-age-1-year.png"), width = 6.2, height = 2)
+
+# other age classes ####
 m2 = dist_maps(pred_all[pred_all$age==2,], scalebiomass = F)
 m3 = dist_maps(pred_all[pred_all$age==3,], scalebiomass = F)
 m4 = dist_maps(pred_all[pred_all$age==4,], scalebiomass = F)
@@ -132,10 +136,10 @@ if(spp_name=="Pacific hake"){
   m5 = dist_maps(pred_all[pred_all$age==5,], scalebiomass = F)
   ggarrange(m1,m2,m3,m4, m5, nrow = 5)}
 
-ggsave( paste0(fig_dir ,"Map-",spp_name,"-year.png"), width = 6.2, height = 8)
+ggsave( paste0(fig_dir , spp_name,"-age-class-year.png"), width = 6.2, height = 8)
 
 
-################################################################################
+# cohort abundance through time ################################################
 
 # sablefish 
[…]

### 2025-01-27 paper.qmd
updating plots; still need fine tuning

@@
 ```{r label="fig-hake-spatial-anomaly", fig.cap="Estimated spatial anomalies for Pacific hake ages 1 -- 5. These fields represent the average anomalies across all years, 2003 -- 2023.", echo=FALSE}
-knitr::include_graphics("plots/hake_spatial_anomalies.png")
+knitr::include_graphics("plots/hake_spatial_anomaly.png")
 ```
@@
 ```{r label="fig-sablefish-spatial-anomaly", fig.cap="Estimated spatial anomalies for sablefish ages 0 -- 9. These fields represent the average anomalies across all years, 2003 -- 2023.", echo=FALSE}
-knitr::include_graphics("plots/sablefish_spatial_anomalies.png")
+knitr::include_graphics("plots/sablefish-spatial-anomaly.png")
 ```

### 2025-01-27 07_plots_abundance_nt.R

@@ -54,7 +54,7 @@ pred_all$rec_year = pred_all$year - pred_all$age
 pred_all$yrs = axis.mod*(pred_all$year-max_year)
 pred_all$lon2 = pred_all$lon + pred_all$yrs
 xmin = floor( min(pred_all$lon2))
-pred_all$p <= quantile(pred_all$p,0.99)
+# pred_all$p <= quantile(pred_all$p,0.99)
 
 # background map info ##########################################################
 states <- map_data("state")

### 2025-01-27 08_plots_spatial_anomaly_nt.R

@@ -1,0 +1,127 @@
+library(tidyverse)
+library(ggplot2)
+library(DHARMa)
+library(viridis)
+library(rnaturalearth)
+library(rnaturalearthdata)
+
+for(spp in 1:2) {
+  spp_name <- c("Pacific hake", "sablefish")[spp]
+  
+  if(spp_name == "Pacific hake") {
+    min_age <- 1 # not many age 0s consistently sampled
+    max_age <- 6
+  } 
+  if(spp_name == "sablefish") {
+    min_age <- 0
+    max_age <- 10
+  }
+  
+  for(a in min_age:(max_age - 1)) {
+    d <- readRDS(paste0("predictions/",spp_name,"_",a,"_surveygrid.rds"))
+    d$species <- spp_name
+    d$age <- a
+    if(spp == 1) {
+      if(a == min_age) {
+        all_output <- d
+      } else {
+        all_output <- rbind(all_output, d)
+      }
+    } else {
+      all_output <- rbind(all_output, d)
+    }
+  }
+  
+}
+
+all_output <- dplyr::filter(all_output, year == 2019)
+
+hake <- dplyr::filter(all_output, species == "Pacific hake") |>
+  dplyr::group_by(age) |>
+  dplyr::mutate(omega_s = scale(omega_s))  # normalize for plotting
+
+sablefish <- dplyr::filter(all_output, species == "sablefish") |>
+  dplyr::group_by(age) |>
+  dplyr::mutate(omega_s = scale(omega_s))  # normalize for plotting
+
+####################
+
+
+states <- map_data("state")
+west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
+################################################################################
+
+# adjust x axis
+
+# plot function
+dist_map2 <- function(dfile=sablefish, 
+                      Z = dfile$omega_s, 
+                      xlim = c(-185, -117), 
+                      midp = 0, 
+                      trans = 'identity',
+                      axis.mod = 3,
+                      max_year = 2023,
+                      no_ages = 5){
+  # prep maps
+  library(rnaturalearth)
+  library(rnaturalearthdata)
+  states <- map_data("state")
+  west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
+  
+  # adjust axis
+  dfile$yrs = axis.mod*(no_ages - dfile$age - 1)
+  dfile$lon2 = dfile$lon - dfile$yrs
[…]

### 2025-01-27 08_plots_spatial_anomaly_nt.R
pretty age class anomaly plots

@@ -73,10 +73,11 @@ dist_map2 <- function(dfile=sablefish,
   dfile$lon2 = dfile$lon - dfile$yrs
   xmin = floor( min(dfile$lon2))
   # axis label for X
-  x1 = seq(-180,-120,axis.mod )
+  x1 = seq(-179,-119,axis.mod )
   xages = seq(min(dfile$age), no_ages-1,1)
   x2 = rep("", length(x1)-length(xages))
   labelsx = c(x2,xages)
+  xbreaks = rev(rev(x1)[1:length(xages)])
   # set midpoing for color ramp
   if(is.na(midp)==T){midp = 0.1*max(dfile$p)}
   # prep axis
@@ -95,10 +96,10 @@ dist_map2 <- function(dfile=sablefish,
                           mid = "white", 
                           high = scales::muted("red"), 
                           midpoint = midp) +  # legend title 
-     coord_fixed(xlim=xlim,ylim=c(32,48),ratio=1.3) +
+    coord_fixed(xlim=xlim,ylim=c(32,48),ratio=1.3) +
     # might need to adjust x-axes to plot all data
-    scale_x_continuous(breaks = seq(-180,-120,axis.mod ), 
-                       minor_breaks = -117:-180, labels = labelsx) +
+    scale_x_continuous(breaks = xbreaks , labels = xages) +
+    #scale_x_continuous(breaks = seq(-179, -119, axis.mod ), labels = labelsx) +
     xlab("Age-class") +
     ylab("Latitude") +
     theme_bw() + 
@@ -111,10 +112,10 @@ dist_map2 <- function(dfile=sablefish,
 
 
 dist_map2(sablefish,no_ages = 10, xlim = c(-152, -117))
-ggsave( paste0(fig_dir,"sablefish-spatial-anomaly.png"), width=6.2, height = 2)
+ggsave( paste0(fig_dir,"sablefish-spatial-anomaly.png"), width=4, height = 2)
 
 dist_map2(hake,no_ages = 6, xlim = c(-140, -117))
-ggsave( paste0(fig_dir,"hake_spatial_anomaly.png"), width=3.2, height = 2)
+ggsave( paste0(fig_dir,"hake_spatial_anomaly.png"), width=3, height = 2)
 
 
 

### 2025-01-27 paper.qmd
pretty figures

@@
 
-```{r label="fig-bycatch-risk", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for age-1 sablefish and port-level forecasts of age-1 CPUE within a radius of 232 km.", echo=FALSE}
+```{r label="fig-bycatch-risk", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for age-1 sablefish and port-level forecasts of age-1 CPUE within a radius of 232 km.
+In the upper plot CPUE is standardized to 1.0 across cohorts for visualization purposes.", echo=FALSE}
 knitr::include_graphics("plots/sablefish_spatial_risk.png")

### 2025-01-27 05_spatial_bycatch_risk_sablefish-nt.R

@@ -1,0 +1,146 @@
+library(tidyverse)
+library(ggplot2)
+library(DHARMa)
+library(viridis)
+home_dir = getwd()
+
+spp <- 2
+spp_name <- c("Pacific hake", "sablefish")[spp]
+
+if(spp_name == "Pacific hake") {
+  min_age <- 1 # not many age 0s consistently sampled
+  max_age <- 5
+} 
+if(spp_name == "sablefish") {
+  min_age <- 0
+  max_age <- 6
+}
+
+d <- readRDS(paste0("predictions/",spp_name,"_",min_age+1,"_surveygrid.rds"))
+
+sablefish <- dplyr::filter(d,year %in% seq(2015,2023,by=2)) |>
+  dplyr::mutate(p = exp(est))
+
+axis.mod = 3
+max_year = 2023
+sablefish$yrs = axis.mod*(sablefish$year-max_year)
+sablefish$lon2 = sablefish$lon + sablefish$yrs
+xmin = floor( min(sablefish$lon2))
+
+# background map info ##########################################################
+states <- map_data("state")
+west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
+################################################################################
+
+# plot function
+dist_maps <- function(dfile, 
+                      xlim = c(-185, -117), 
+                      midp = NA, 
+                      trans = 'identity',
+                      scalebiomass = T) {
+  # data preparation
+  # dfile = pred_all[pred_all$age==1,]
+  dfile = dplyr::filter(dfile, p <= quantile(dfile$p,0.99))
+  
+  # scale data to 0-1 for each year ; if decided
+  
+  if(scalebiomass == TRUE){
+    dmean = dfile %>%
+      group_by(year) %>%
+      summarise(maxp = max(p, na.rm = T))
+    # slow
+    dfile = left_join(dfile, dmean)
+    dfile$p = dfile$p/dfile$maxp
+  }
+  # set midpoing for color ramp
+  if(is.na(midp)==T){midp = 0.1*max(dfile$p)}
+  # begin plotting
+  plotx <- 
+    ggplot(data=west_coast, aes(x = long, y = lat), 
+           fill = grey(0.9), color = "black") +
+    geom_polygon(aes(x = long, y = lat, group=group), 
+                 fill = grey(0.9), color = "black") +
+    geom_point(data=dfile , 
[…]

### 2025-01-28 07_plots_abundance_nt.R
still working on fig minutia

@@ -19,7 +19,7 @@ spp_name <- c("Pacific hake", "sablefish")[1]
 if(spp_name == "Pacific hake") {
   min_age <- 1 # not many age 0s consistently sampled
   max_age <- 6
-  years <- 2003:2019
+  years <- 2003:2023
   ages <- seq(min_age, min_age+4)
 } 
 if(spp_name == "sablefish") {
@@ -66,11 +66,14 @@ dist_maps <- function(dfile,
                       xlim = c(-185, -117), 
                       midp = NA, 
                       trans = 'identity',
-                      scalebiomass = T) {
+                      scalebiomass = T,
+                      x_lab = '',
+                      scale_label = 'CPUE',
+                      xadjust = 0) {
   # data preparation
   # dfile = pred_all[pred_all$age==1,]
   dfile = dplyr::filter(dfile, p <= quantile(dfile$p,0.99))
-  
+  dfile$lon2 = dfile$lon2+xadjust
   # scale data to 0-1 for each year ; if decided
   
   if(scalebiomass == TRUE){
@@ -94,7 +97,7 @@ dist_maps <- function(dfile,
     # might need to adjust here based on the scale of the data, esp 'trans' term.
     # scale_color_viridis_c(option = 'turbo', direction = 1,  
     #                     name='CPUE') +
-    scale_color_gradient2(name = "CPUE", 
+    scale_color_gradient2(name = scale_label, 
                           transform = trans, 
                           low = scales::muted("black"),
                           midpoint = midp,
@@ -105,12 +108,14 @@ dist_maps <- function(dfile,
     scale_x_continuous(breaks = seq(-180,-120,axis.mod ), 
                      minor_breaks = -117:-180, labels = 2003:max_year) +
     #scale_y_continuous(breaks = c(35,40,45), minor_breaks = c(32:48)) +
-    xlab("Year") +
+    xlab(x_lab) +
     ylab("Latitude") +
     theme_bw() + 
-    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
-        legend.title = element_text(size = 8),
[…]

### 2025-01-28 paper.qmd
editing text

@@
 We validate our approach using the recursive nature of age data to forecast the spatial distribution of fish by age class across survey locations in subsequent years; these results highlight that forecast models have predictive skill (more so for sablefish than hake) and the predictive ability is highest for older age classes.
-Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2010, 2014 and 2016 cohorts; sablefish from 2008, 2010, 2016, and 2021 cohorts).
+Output from our models can be used to predict the spatial age composition, which may be particularly useful for identifying large cohorts (hake from 2008, 2010, 2014 and 2016 cohorts; sablefish from 2008, 2010, 2016, and 2021 cohorts).
 Using predictions of age-1 sablefish as an example, we also demonstrate how our models may be used to forecast future bycatch risk in space, helping increase the efficiency of fisheries.
@@
 These shifts can create age-specific distributions across habitats, particularly for species that undergo distinct life history phases.
-Additionally, recruitment pulses and migration patterns vary by age, influencing the spatial distribution of younger versus older fish.
+Additionally, spatiallly variable recruitment pulses and age-related migration patterns can influence the spatial distribution of younger versus older fish.
 Fishing selectivity often further accentuates these patterns by disproportionately removing larger, older fish, which alters the population structure and potentially shifts spatial distributions [@hilborn1992; @methot2013].
-By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relati
+By incorporating age, SDMs could better predict these dynamics and improve our understanding of species-environment relationships.
 
@@
 On the west coast of the USA, the West Coast Groundfish Bottom Trawl Survey (WCGBTS) has been an annual survey conduced from 2003 - present.
-The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA (Keller et al., 2017).
+The WCGBTS is designed to estimate the abundance, size, and age composition of groundfish species important to commercial and recreational fisheries found in near-bottom habitats on the west coast of the USA [@keller2017].
 The survey effort is concentrated in summer months and has been conducted annually since 2003 (here we use the data through 2018; data are publicly available at https://www.nwfsc.noaa.gov/data).
@@
 Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish.
-\[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years.
-Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada (Grandin et al.
-2024) and sablefish is frequently assessed because of its high commercial value.
+[more on biological gradients] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years.
+Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada [@Grandin2024] and sablefish is frequently assessed because of its high commercial value.
 For our analysis, we used hake collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per year).
@@
 
-```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2010, 2014, 2016).
+```{r label="fig-hake-spatial-composition", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong cohorts in our dataset (2008, 2010, 2014, 2016).
 CPUE is standardized to 1.0 across cohorts for visualization purposes.

### 2025-01-28 paper.qmd
reference and text

@@
 To illustrate potential benefits to fisheries management, we used output from our models to generate spatiotemporal predictions of age-1 sablefish (from age-0 sablefish the year prior).
-\[Need some text here on background why sablefish as bycatch is important.
-In 2017 for example, the at-sea whiting fishery had 50 mt of sablefish quota set aside in the area north of 36 deg N, but the fleet caught more than 3x as much\].
+Sablefish are part of the DTS complex, which includes Dover sole *Microstomus pacificus*, shortspine *Sebastomus alascanus,* longspine thornyheads *S.
+altivelis*, and sablefish.
+This complex forms and inportant fishery on the West Cast, and by-catch of sablefish can impact quotas for the other three species, as well as the at-sea Hake fishery.
+For exmaple, in 2017, the at-sea hake fishery had 50 mt of sablefish quota set aside in the area north of 36 °N, but the fleet caught more than 3x as much, promting in-season warnings to alter fishing behavior.
+
 Gridded predictions from the Poisson GLM represent the expected number of fish per unit of effort in each location.
 To turn these into a potentially more useful measure of risk, we calculate the predicted total age-1 sablefish catch per unit of effort (CPUE) within a given radius of major fishing ports on the US west coast.
-For consistency with previous work \[TODO: cite ESR\] we used a radius of 232km.
+For consistency with previous work [@leising2024cciea], we used a radius of **282km** **\[was listed as 232 (my fault) but should be 282 to be consistent with the ESR\]**.
 These predictions of port-level CPUE were then summarized as annual time series (1-step ahead forecasts).

### 2025-01-28 paper.qmd
update port radius to 282 km to match ESR

@@
 altivelis*, and sablefish.
-This complex forms and inportant fishery on the West Cast, and by-catch of sablefish can impact quotas for the other three species, as well as the at-sea Hake fishery.
+This complex forms and important fishery on the West Cast, and by-catch of sablefish can impact quotas for the other three species, as well as the at-sea hake fishery.
 For exmaple, in 2017, the at-sea hake fishery had 50 mt of sablefish quota set aside in the area north of 36 °N, but the fleet caught more than 3x as much, promting in-season warnings to alter fishing behavior.
@@
 To turn these into a potentially more useful measure of risk, we calculate the predicted total age-1 sablefish catch per unit of effort (CPUE) within a given radius of major fishing ports on the US west coast.
-For consistency with previous work [@leising2024cciea], we used a radius of **282km** **\[was listed as 232 (my fault) but should be 282 to be consistent with the ESR\]**.
+For consistency with previous work , we used a radius of 282 km [@leising2024cciea].
 These predictions of port-level CPUE were then summarized as annual time series (1-step ahead forecasts).

### 2025-01-28 05_spatial_bycatch_risk_sablefish.R

@@ -50,7 +50,7 @@ p1 <- ggplot(coast_proj) +
   facet_wrap(~year, nrow = 1)
 
 # read in ports
-ports_rad <- readRDS("~/Documents/Github Projects/spatial-age-models/ports_rad.rds")
+ports_rad <- readRDS( paste0(home_dir, "/ports_rad.rds"))
 d$dist_AST <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="AST")])^2 + 
                      (d$Y - ports_rad$Y[which(ports_rad$Pcid=="AST")])^2)
 d$dist_COS <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="COS")])^2 + 
@@ -66,7 +66,7 @@ d$dist_BRG <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="BRG")])^2 +
 d$dist_MRO <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="MRO")])^2 + 
                      (d$Y - ports_rad$Y[which(ports_rad$Pcid=="MRO")])^2)
 
-cutoff <- 232
+cutoff <- 282 # was 232 updated here but not run as of 2025-01-28
 distances <- dplyr::group_by(d, year) |>
   dplyr::summarise(AST = sum(exp(est[which(dist_AST < cutoff)])),
                    COS = sum(exp(est[which(dist_COS < cutoff)])),

### 2025-01-28 09_plots_spatial_bycatch_risk_sablefish_pretty.R

@@ -108,7 +108,7 @@ d$dist_BRG <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="BRG")])^2 +
 d$dist_MRO <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="MRO")])^2 + 
                      (d$Y - ports_rad$Y[which(ports_rad$Pcid=="MRO")])^2)
 
-cutoff <- 232
+cutoff <- 282
 distances <- dplyr::group_by(d, year) |>
   dplyr::summarise(AST = sum(exp(est[which(dist_AST < cutoff)])),
                    COS = sum(exp(est[which(dist_COS < cutoff)])),

### 2025-01-28 paper.qmd
nicer figures, some editing.

@@
 Though the WCGBTS survey samples hundreds of species, we concentrated our analysis on two of the well sampled species from the WCGBTS, North Pacific hake and sablefish.
-[more on biological gradients] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years.
+\[more on biological gradients\] While otoliths are sampled continuously during the WCGBTS survey for a broad range of species, otoliths are generally aged when species are being prioritized for stock assessment by the Pacific Fishery Management Council (PFMC); full assessments between species may occur irregularly and be sporadically updated every 5 -- 10 years.
 Hake and sablefish represent exceptions, as the hake stock is assessed annually by co-managers from the USA and Canada [@Grandin2024] and sablefish is frequently assessed because of its high commercial value.
@@
 
-One useful output from our spatiotemporal models is the estimated spatial field, $\omega_s$, representing the average spatial distribution across years.
+One useful output from our spatiotemporal models is the estimated field spatial f, $\omega_s$, representing the average spatial distribution across years.
 Results from our model of North Pacific hake show a concentration of age-1 hake near the coast, and more of latitudinal north - south break for 2 - 3 year old hake (with higher occurrences in the South) @fig-hake-spatial-anomaly.
@@
 Estimates of age composition by year may be useful in tracking the distribution of cohorts through time, as well as examining variability in distributions of particular age classes over time.
-For example, results from our models can be used to project the spatial distributions of strong hake cohorts (2010, 2014, 2016; @fig-hake-spatial-composition) and strong sablefish cohorts (2008, 2010, 2016, 2021, @fig-sablefish-spatial-composition).
+For example, for both hake ([Figs.
+@fig-hake-spatial-composition] & [-@fig-hake-spatial-composition-all]) and sablefish ([Figs.
+@fig-sablefish-spatial-composition] & [-@fig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct.
+Additionally, results from our models can be used to project the spatial distributions of strong hake cohorts (2008, 2010, 2014, 2016; @fig-hake-spatial-composition) and strong sablefish cohorts (2008, 2010, 2016, 2021, @fig-sablefish-spatial-composition).
 These predictions are also useful in revealing subtle differences between strong cohort years.
 For example, the 2014 cohort of Pacific hake had a larger fraction of age-1 individuals distributed in the northern part of the domain (along the coast of Washington state), while other cohorts had hotspots in the center and southern portion of the domain.
-Similarly for sablefish, the 2010 cohort also appeared to be more concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution.
+Similarly for sablefish, the 2016 cohort was concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution.
+In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 (Fig.
+@fig-sablefish-spatial-composition-all\], concurrent with 'the Blob' and other marine heatwaves.
 

### 2025-01-28 03_poisson_regression.R

@@ -71,14 +71,22 @@ for(spp in 1:2) {
 
 ggplot(all_output, aes(age, exp(mean), col = species)) + 
   geom_hline(aes(yintercept=1), col="red", alpha=0.3) + 
-  geom_pointrange(aes(ymin = exp(lower), ymax = exp(upper)), position=position_dodge(0.4)) + 
-  geom_point(position=position_dodge(0.4)) + 
+  geom_pointrange(aes(ymin = exp(lower), ymax = exp(upper)), 
+                  position=position_dodge(0.4), size=0.1) + 
+  geom_point(position=position_dodge(0.4), size = 0.1) + 
   scale_color_viridis_d(option="magma", begin=0.2, end = 0.8, name="Species") +
+  scale_x_continuous(breaks = seq(0,9,1), minor_breaks = seq(0,9,1)) + 
   xlab("Age") + 
   ylab("Estimated slope") + 
-  theme_bw()
+  theme_bw() + 
+  theme(#
+    axis.text = element_text(size=8),
+    axis.title = element_text(size=8),
+    legend.title = element_text(size = 8),
+    legend.text = element_text(size = 8),
+    legend.key.size = unit(1,'lines')) 
 
-ggsave("plots/glm_coefficients.png", width=6, height = 5)
+ggsave("plots/glm_coefficients.png", width=3.5, height = 2)
 
 
 # look at coefficients from the year model, as a diagnostic

### 2025-01-28 04_spatial_anomaly_plots.R

@@ -71,7 +71,7 @@ ggplot(coast_proj) +
         strip.background = element_rect(fill = "white"),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Set facet label background to white
   facet_wrap(~age, nrow = 1)
-ggsave("plots/hake_spatial_anomalies.png", width = 7, height = 5)  
+ggsave("plots/hake_spatial_anomalies-facet.png", width = 7, height = 5)  
   
 
 ggplot(coast_proj) + 
@@ -85,5 +85,5 @@ ggplot(coast_proj) +
         strip.background = element_rect(fill = "white"),
         axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  # Set facet label background to white
   facet_wrap(~age, nrow = 2)
-ggsave("plots/sablefish_spatial_anomalies.png", width = 7, height = 8)  
+ggsave("plots/sablefish_spatial_anomalies-facet.png", width = 7, height = 8)  
 

### 2025-01-28 05_spatial_bycatch_risk_sablefish.R

@@ -66,7 +66,7 @@ d$dist_BRG <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="BRG")])^2 +
 d$dist_MRO <- sqrt((d$X - ports_rad$X[which(ports_rad$Pcid=="MRO")])^2 + 
                      (d$Y - ports_rad$Y[which(ports_rad$Pcid=="MRO")])^2)
 
-cutoff <- 282 # was 232 updated here but not run as of 2025-01-28
+cutoff <- 282 # was 232 updated
 distances <- dplyr::group_by(d, year) |>
   dplyr::summarise(AST = sum(exp(est[which(dist_AST < cutoff)])),
                    COS = sum(exp(est[which(dist_COS < cutoff)])),
@@ -96,5 +96,5 @@ p2 <- ggplot(distances_long, aes(year, value, group = Name, col = Name)) +
 
 p3 <- gridExtra::grid.arrange(p1, p2, ncol = 1)
 
-ggsave(p3, filename = "plots/sablefish_spatial_risk.png", width = 7, height = 6)  
+ggsave(p3, filename = "plots/sablefish_spatial_risk-old.png", width = 7, height = 6)  
 

### 2025-01-28 archive/02_recursive_model.R

@@ -150,31 +150,56 @@ pars <- rbind(hake_pars, sablefish_pars)
 
 p1 <- dplyr::filter(pars, term == "range") |> 
   ggplot(aes(age, estimate, col = species)) + 
-  geom_pointrange(aes(ymin=estimate-2*std.error, ymax = estimate+2*std.error), position = position_dodge(0.4)) + 
+  geom_pointrange(aes(ymin=estimate-2*std.error, ymax = estimate+2*std.error), 
+                  position = position_dodge(0.4), size = 0.1) + 
+  scale_x_continuous(breaks = seq(0,9,1), minor_breaks = seq(0,9,1)) + 
   scale_color_viridis_d(option="magma", begin=0.2, end = 0.8) +
   xlab("Age") + 
   ylab("Estimated range (km)") + 
   theme_bw() + 
-  theme(legend.position = "none")
+  theme(legend.position = 'none',
+        axis.text = element_text(size=8),
+        axis.title = element_text(size=8),
+        legend.title = element_text(size = 8),
+        legend.text = element_text(size = 8),
+        legend.key.size = unit(1,'lines'))
 
 p2 <- dplyr::filter(pars, term == "sigma_O") |> 
   ggplot(aes(age, estimate, col = species)) + 
-  geom_pointrange(aes(ymin=estimate-2*std.error, ymax = estimate+2*std.error), position = position_dodge(0.4)) + 
+  geom_pointrange(aes(ymin=estimate-2*std.error, ymax = estimate+2*std.error), 
+                  position = position_dodge(0.4), size = 0.1) + 
   scale_color_viridis_d(option="magma", begin=0.2, end = 0.8) +
+  scale_x_continuous(breaks = seq(0,9,1), minor_breaks = seq(0,9,1)) + 
   xlab("Age") + 
   ylab(expression(paste("Spatial ", sigma))) + 
   theme_bw() + 
-  theme(legend.position = "none")
+  theme(legend.position = c(0.65, 0.85),
+        axis.text = element_text(size=8),
+        axis.title = element_text(size=8),
+        legend.title = element_blank(),
+        legend.text = element_text(size = 8),
+        legend.key.size = unit(0.4,'lines'),
+        legend.background=element_blank()
+        )
 
 p3 <- dplyr::filter(pars, term == "sigma_E") |> 
   ggplot(aes(age, estimate, col = species)) + 
[…]

### 2025-01-28 paper.qmd
Updated references; corrected titles; text edits; render

@@
 Younger age classes of hake and sablefish exhibited broader distributions generally in shallower areas, while older individuals were more concentrated offshore (sablefish in deeper waters, hake in more northern regions).
-The deeper distribution of older sablefish reflects their preference for benthic habitats, while the observed north-south gradient in older hake aligns with known migratory behaviors influenced by spawning and feeding grounds [@agostini2006].
+The deeper distribution of older sablefish reflects their preference for benthic habitats and an ontogenetic shift to deeper waters, while the observed north-south gradient in older hake aligns with known migratory behaviors influenced by spawning and feeding grounds [@agostini2006].
 
@@
 In contrast, the lower predictive skill for younger age classes reflects their higher spatiotemporal variability, likely driven by recruitment pulses and early-life survival dynamics.
-Tracking cohorts through time using our framework proved effective in identifying strong year classes (Figs.
-3-4).
+Tracking cohorts through time using our framework proved effective in identifying strong year classes ([Figs.
+@fig-hake-spatial-composition] & [-@fig-hake-spatial-composition]).
 Identifying these cohorts may be useful to stock assessment and management -- for instance, understanding the spatial distribution of strong cohorts could be used to inform spatial harvest allocation.
@@
 Such information could help fishers avoid areas associated with high bycatch risk, or be used to adjust the spatial distribution of quota in certain years.
-The observed peaks in age-1 sablefish CPUE during certain years suggest the influence of recruitment pulses, which are thought to be tied to environmental drivers [@tolimieri2018].
+The observed peaks in age-1 sablefish CPUE during certain years suggest the influence of recruitment pulses, which are thought to be tied to environmental drivers [@tolimieri2018; @tolimieri2023sea].
 Future efforts could also incorporating environmental covariates to enhance its predictive accuracy and further support climate-resilient fisheries management.
@@
 The costs of ship-based surveys, such as the WCGBTS survey, have increased around the world, and numerous efforts are being implemented to make these programs more efficient [@ices2020].
-While age samples are more expensive to collect than length information, there is also increased precision associated with age data.
+While age samples are more expensive to collect than length information, there is value from the increased precision associated with age data.
 

### 2025-01-28 07_plots_abundance_pretty.R
minor figure edits

@@ -14,7 +14,7 @@ library(ggpubr)
 home_dir = getwd()
 fig_dir = paste0(home_dir,"/plots/")
 
-spp_name <- c("Pacific hake", "sablefish")[1]
+spp_name <- c("Pacific hake", "sablefish")[2]
 
 if(spp_name == "Pacific hake") {
   min_age <- 1 # not many age 0s consistently sampled
@@ -69,7 +69,7 @@ dist_maps <- function(dfile,
                       scalebiomass = T,
                       x_lab = '',
                       scale_label = 'CPUE',
-                      xadjust = 0) {
+                      xadjust = -3) {
   # data preparation
   # dfile = pred_all[pred_all$age==1,]
   dfile = dplyr::filter(dfile, p <= quantile(dfile$p,0.99))
@@ -105,8 +105,8 @@ dist_maps <- function(dfile,
                           high = scales::muted("red")) + 
     coord_fixed(xlim=xlim,ylim=c(32,48),ratio=1.3) +
     # might need to adjust x-axes to plot all data
-    scale_x_continuous(breaks = seq(-180,-120,axis.mod ), 
-                     minor_breaks = -117:-180, labels = 2003:max_year) +
+    scale_x_continuous(breaks = seq(-179,-119,axis.mod ), 
+                     minor_breaks = -117:-179, labels = 2003:max_year) +
     #scale_y_continuous(breaks = c(35,40,45), minor_breaks = c(32:48)) +
     xlab(x_lab) +
     ylab("Latitude") +
@@ -123,11 +123,9 @@ dist_maps <- function(dfile,
 # age-class each year ##########################################################
 
 # print(m0)
-# age 1 sablefish for figure 6 ####
-# combine later for fig 6
-m1 = dist_maps(pred_all[pred_all$age==1,],scale_label = 'CPUE age-1')
-ggsave( paste0(fig_dir ,spp_name,"-age-1-year.png"), width = 6.2, height = 2)
+# might need to tinker with xadjust to get axes correct
 
+m1 = dist_maps(pred_all[pred_all$age==1,],scale_label = 'CPUE age-1')
 # other age classes ####
 m2 = dist_maps(pred_all[pred_all$age==2,],scale_label = 'CPUE age-2')
 m3 = dist_maps(pred_all[pred_all$age==3,],scale_label = 'CPUE age-3')
[…]

### 2025-01-28 07_plots_abundance_pretty.R
Update 07_plots_abundance_pretty.R

@@ -69,7 +69,7 @@ dist_maps <- function(dfile,
                       scalebiomass = T,
                       x_lab = '',
                       scale_label = 'CPUE',
-                      xadjust = -3) {
+                      xadjust = 0) {
   # data preparation
   # dfile = pred_all[pred_all$age==1,]
   dfile = dplyr::filter(dfile, p <= quantile(dfile$p,0.99))

### 2025-01-28 08_plots_spatial_anomaly_pretty.R
more minor figure edits

@@ -103,13 +103,15 @@ dist_map2 <- function(dfile=sablefish,
     xlab("Age-class") +
     ylab("Latitude") +
     theme_bw() + 
-    theme(#axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1),
+    theme(axis.text = element_text(size=8),
+          axis.title = element_text(size=8),
           legend.title = element_text(size = 8),
           legend.text = element_text(size = 8),
-          legend.key.size = unit(1,'lines')) 
+          legend.key.size = unit(0.7,'lines')) 
   return(plotx)
 }
 
+fig_dir = paste0(getwd(),"/plots/")
 
 dist_map2(sablefish,no_ages = 10, xlim = c(-152, -117))
 ggsave( paste0(fig_dir,"sablefish-spatial-anomaly.png"), width=4, height = 2)

### 2025-01-28 09_plots_spatial_bycatch_risk_sablefish_pretty.R
minor figure fixes to bycatch risk

@@ -65,7 +65,7 @@ dist_maps <- function(dfile,
     # might need to adjust here based on the scale of the data, esp 'trans' term.
     # scale_color_viridis_c(option = 'turbo', direction = 1,  
     #                     name='CPUE') +
-    scale_color_gradient2(name = "CPUE", 
+    scale_color_gradient2(name = "CPUE age-1", 
                           transform = trans, 
                           low = scales::muted("black"),
                           midpoint = midp,

### 2025-01-29 paper.qmd
playing with format and yml

@@
 format: pdf
-date: "`r Sys.Date()`"
+crossref:
+  custom:
+    - kind: float
+      key: supfig
+      latex-env: supfig
+      reference-prefix: Fig. S
+      space-before-numbering: false
+      latex-list-of-description: Supplementary Figure
 author:
-  - name: Eric & Nick
+  - name: Eric Ward
+    affiliation:
+      - id: cb
+        name: Consercation Biology Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration
+        address: 2725 Montlake Blvd E
+        city: Seattle
+        state: WA
+        zip: 98112
+  - name: Nick Tolimieri
+    affiliation:
+      - ref: cb
+
 editor: visual
@@
 
-## Abstract
+## *Abstract*
 
@@
 One useful output from our spatiotemporal models is the estimated field spatial f, $\omega_s$, representing the average spatial distribution across years.
-Results from our model of North Pacific hake show a concentration of age-1 hake near the coast, and more of latitudinal north - south break for 2 - 3 year old hake (with higher occurrences in the South) @fig-hake-spatial-anomaly.
+Results from our model of North Pacific hake show a concentration of age-1 hake near the coast, and more of latitudinal north - south break for 2 - 3 year old hake (with higher occurrences in the South, [Fig.
+@fig-hake-spatial-anomaly]).
 As expected, age 4 hake appear to have a northern distribution, though the gradient is less clear than for ages 2 - 3.
-For sablefish, our models estimate a concentration of age 0 - 2 sablefish in shallow water, with a deeper distribution by age 3 (@fig-sablefish-spatial-anomaly).
-Latitudinally, there are several areas in central Oregon that appear to have consistently lower occurrences than average for older age classes (@fig-sablefish-spatial-anomaly).
+For sablefish, our models estimate a concentration of age 0 - 2 sablefish in shallow water, with a deeper distribution by age 3 ([Fig.
+@fig-sablefish-spatial-anomaly]).
+Latitudinally, there are several areas in central Oregon that appear to have consistently lower occurrences than average for older age classes ([Fig.
+@fig-sablefish-spatial-anomaly]).
 
@@
 For example, for both hake ([Figs.
-@fig-hake-spatial-composition] & [-@fig-hake-spatial-composition-all]) and sablefish ([Figs.
-@fig-sablefish-spatial-composition] & [-@fig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct.
-Additionally, results from our models can be used to project the spatial distributions of strong hake cohorts (2008, 2010, 2014, 2016; @fig-hake-spatial-composition) and strong sablefish cohorts (2008, 2010, 2016, 2021, @fig-sablefish-spatial-composition).
+@fig-hake-spatial-composition] & [@supfig-hake-spatial-composition-all]) and sablefish ([Figs.
+@fig-sablefish-spatial-composition] & [@supfig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct.
+Additionally, results from our models can be used to project the spatial distributions of strong hake cohorts (2008, 2010, 2014, 2016; [Fig.
+@fig-hake-spatial-composition]) and strong sablefish cohorts (2008, 2010, 2016, 2021, [Fig.
+@fig-sablefish-spatial-composition]).
 These predictions are also useful in revealing subtle differences between strong cohort years.
@@
 Similarly for sablefish, the 2016 cohort was concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution.
-In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 (Fig.
-@fig-sablefish-spatial-composition-all\], concurrent with 'the Blob' and other marine heatwaves.
+In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 ([@supfig-sablefish-spatial-composition-all]), concurrent with 'the Blob' and other marine heatwaves.
 
-When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range with increasing age (@fig-spatial-parameters).
+When comparing the spatial and spatiotemporal parameters across ages and species, we find that for most ages, there is a decrease in the estimated spatial and spatiotemporal variances as well as a decrease in the spatial range with increasing age ([Fig.
+@fig-spatial-parameters]).
 Spatial and spatiotemporal variances control the magnitude of the peaks and valleys in estimated latent fields, meaning that spatial fields estimated for older fish are less variable.
 Similarly, the spatial range determines the distance at which locations are functionally independent (for the Matern, $\rho$ \~ 0.13).
-The decline in range with age implies that spatial distributions also become more concentrated as a function of age (@fig-spatial-parameters).
+The decline in range with age implies that spatial distributions also become more concentrated as a function of age ([Fig.
+@fig-spatial-parameters]).
 
-Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish (@fig-glm-coefficients).
+Our 1-step ahead validation analyses indicate that there is a stronger association between older ages compared to younger ages, and this effect is strongest for sablefish ([Fig.
+@fig-glm-coefficients]).
 When comparing our initial Poisson GLM to a more complicated GLMM with random year effects in the intercept and slope, we found little support for shared correlations across age classes of a given species.
 Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another).
-With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude (@fig-glm-coefficients-time).
+With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude ([@supfig-glm-coefficients-time]).
 
 Our forecasts of age-1 sablefish CPUE illustrates the variability of potential bycatch risk in space and time.
-In most years for example, age-1 CPUE is lowest in ports in the state of California (Eureka, Fort Bragg, Morro Bay) and highest in several ports in the state of Oregon (Astoria, Coos Bay, @fig-bycatch-risk).
+In most years for example, age-1 CPUE is lowest in ports in the state of California (Eureka, Fort Bragg, Morro Bay) and highest in several ports in the state of Oregon (Astoria, Coos Bay, [Fig.
+@fig-bycatch-risk]).
 These results also illustrate the high temporal variability in age-1 CPUE, with peaks occuring in 2009, 2014, 2017, and 2021, @fig-bycatch-risk).
-During peak years, there is also variability among ports -- forecasts of age-1 CPUE were similar for Astoria and Coos Bay in 2021 for instance, but the potential bycatch risk near Astoria was higher in 2017 (@fig-bycatch-risk).
+During peak years, there is also variability among ports -- forecasts of age-1 CPUE were similar for Astoria and Coos Bay in 2021 for instance, but the potential bycatch risk near Astoria was higher in 2017 ([Fig.
+@fig-bycatch-risk]).
 
@@
 
-\newpage
+### Acknowledgments
+
+We thank Owen Liu and XX reviewers for constructive comments on the manuscript.
+
+```{=tex}
+\break
+\clearpage
+```
+## References
+
+::: {#refs}
+:::
+
+```{=tex}
+\break
+\clearpage
+```
+## Figures
 
@@
 
-\newpage
+```{=tex}
+\break
+\clearpage
+```
+## Supplementary material {.appendix}
 
-```{r label="fig-spatial-parameters", fig.cap="Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations).
+```{r label="supfig-spatial-parameters", fig.cap="Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations).
 Points represent mean estimates; lines represent 95% confidence intervals", echo=FALSE}
@@
 
-```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5).
+```{r label="supfig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5).
 CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
@@
 
-```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4).
+```{r label="supfig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4).
 CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
@@
 
-```{r label="fig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year.
+```{r label="supfig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year.
 ", echo=FALSE}
@@
 ```
-
-```{=tex}
-\break
-\clearpage
-```
-## References

### 2025-01-29 paper.qmd
tinkering with YAML

@@
 title: "Advancing spatially explicit fisheries management with age-structured species distribution models"
-format: pdf
-crossref:
-  custom:
-    - kind: float
-      key: supfig
-      latex-env: supfig
-      reference-prefix: Fig. S
-      space-before-numbering: false
-      latex-list-of-description: Supplementary Figure
+
+format: 
+  pdf:
+    keep-tex: true
+    template-partials: 
+      - title.tex
+    include-in-header:
+      text: |
+        \usepackage[noblocks]{authblk}
+        \renewcommand*{\Authsep}{, }
+        \renewcommand*{\Authand}{, }
+        \renewcommand*{\Authands}{, }
+        \renewcommand\Affilfont{\small}
 author:
   - name: Eric Ward
-    affiliation:
-      - id: cb
-        name: Consercation Biology Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration
-        address: 2725 Montlake Blvd E
-        city: Seattle
-        state: WA
-        zip: 98112
   - name: Nick Tolimieri
-    affiliation:
-      - ref: cb
+    affiliations:
+    - Conservation Biology Division, Northwest Fisheries Science Center, National Marine Fisheries Service, National Oceanographic and Atmospheric Administration, 2725 Montlake Blvd E, Seattle WA 98112
 
@@
 For example, for both hake ([Figs.
-@fig-hake-spatial-composition] & [@supfig-hake-spatial-composition-all]) and sablefish ([Figs.
-@fig-sablefish-spatial-composition] & [@supfig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct.
+@fig-hake-spatial-composition] & [-@fig-hake-spatial-composition-all]) and sablefish ([Figs.
+@fig-sablefish-spatial-composition] & [-@fig-sablefish-spatial-composition-all]), distinct patches of high age-0 abundance tended to disperse as the fish aged such that spatial distributions for older age-classes were less distinct.
 Additionally, results from our models can be used to project the spatial distributions of strong hake cohorts (2008, 2010, 2014, 2016; [Fig.
@@
 Similarly for sablefish, the 2016 cohort was concentrated in the north (age-1 individuals in 2011), versus other cohorts that had a more coastwide distribution.
-In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 ([@supfig-sablefish-spatial-composition-all]), concurrent with 'the Blob' and other marine heatwaves.
+In fact, age-1 fish were uncommon south of approximately Cape Mendocino (40 °N) from 2015 - 2020 ([@fig-sablefish-spatial-composition-all]), concurrent with 'the Blob' and other marine heatwaves.
 
@@
 Shared correlations would indicate for example that age-2 and age-3 sablefish may be more or less predictable in certain years (indicating that information about one age class may be useful in predicting another).
-With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude ([@supfig-glm-coefficients-time]).
+With the exception of a strong positive correlation between age-0 and age-1 sablefish, most correlations between age classes were smaller than 0.3 in magnitude ([@fig-glm-coefficients-time]).
 
@@
 By providing high-resolution predictions of spatial age composition, these models enable managers to align fishing effort with ecological and economic goals, ensuring the long-term sustainability of marine resources.
-Future research could extend this framework by incorporating environmental covariates, such as temperature or oxygen, to account for dynamic habitat preferences, or fisheries removals to account for local depletion [@ono2016].
+Future research could extend this framework by incorporating environmental covariates, such as temperature or oxygen, to account for dynamic habitat preferences, or fisheries removals to account for local depletion [@ono2016space].
 Additionally, exploring the potential for multi-species models could provide insights into ecosystem-level interactions and their influence on spatial distributions.
@@
 ```
-## Supplementary material {.appendix}
+# Supplementary material {.appendix}
 
-```{r label="supfig-spatial-parameters", fig.cap="Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations).
-Points represent mean estimates; lines represent 95% confidence intervals", echo=FALSE}
+```{r }
+#| label: fig-spatial-parameters
+#| echo: false
+#| fig-cap: "Estimated spatial parameters (range, and spatial and spatiotemporal standard deviations). Points represent mean estimates; lines represent 95% confidence intervals"
+
 knitr::include_graphics("plots/spatial_parameters.png")
@@
 
-```{r label="supfig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5).
+```{r label="fig-hake-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages (1-5).
 CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
@@
 
-```{r label="supfig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4).
+```{r label="fig-sablefish-spatial-composition-all", fig.cap="Estimated spatial catch per unit effort (CPUE, kg per km2) for sablefish; rows represent ages (0-4).
 CPUE has been scaled for each cohort for visualization purposes.", echo=FALSE}
@@
 
-```{r label="supfig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year.
+```{r label="fig-glm-coefficients-time", fig.cap="Time-varying coefficients (with random effects in year) relating predicted densities of age a fish in year t to observed numbers the following year.
 ", echo=FALSE}

### 2025-01-29 paper.tex
tinkering with YAML

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}

\usepackage{amsmath,amssymb}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
\usepackage{lmodern}
\ifPDFTeX\else  
    % xetex/luatex font selection
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\setlength{\emergencystretch}{3em} % prevent overfull lines
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
% Make \paragraph and \subparagraph free-standing
\ifx\paragraph\undefined\else
  \let\oldparagraph\paragraph
  \renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
  \let\oldsubparagraph\subparagraph
  \renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}\usepackage{longtable,booktabs,array}
\usepackage{calc} % for calculating minipage widths
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

\usepackage[noblocks]{authblk}
\renewcommand*{\Authsep}{, }
\renewcommand*{\Authand}{, }
\renewcommand*{\Authands}{, }
\renewcommand\Affilfont{\small}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\makeatother
\makeatletter
\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Advancing spatially explicit fisheries management with age-structured species distribution models},
  pdfauthor={Eric Ward; Nick Tolimieri},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

  \title{Advancing spatially explicit fisheries management with
age-structured species distribution models}

    \author{Eric Ward}
    \author{Nick Tolimieri}
    \affil{%
          Conservation Biology Division, Northwest Fisheries Science
          Center, National Marine Fisheries Service, National
          Oceanographic and Atmospheric Administration, 2725 Montlake
          Blvd E, Seattle WA 98112
        }

  \date{}
\begin{document}
\maketitle
\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi

\hypertarget{abstract}{%
\subsection{\texorpdfstring{\emph{Abstract}}{Abstract}}\label{abstract}}

Species distribution models (SDMs) are widely used in fisheries and
ecology, providing insights into species' spatial distributions. While
many SDMs incorporate external information, such as environmental or
habitat features, the majority overlook age-specific information,
despite the known influence of age structure on fish distribution and
abundance. Including age information may be particularly valuable for
species exhibiting age-related spatial patterns driven by ontogenetic
shifts, recruitment dynamics, and selective fishing pressures. Here, we
develop an age-structured SDM framework using data from the west coast
of the USA, and concentrate on two species with distinct life histories:
North Pacific hake (\emph{Merluccius productus}) and sablefish
(\emph{Anoplopoma fimbria}). Our framework uses spatiotemporal models to
predict age-specific distributions. We validate our approach using the
recursive nature of age data to forecast the spatial distribution of
fish by age class across survey locations in subsequent years; these
results highlight that forecast models have predictive skill (more so
for sablefish than hake) and the predictive ability is highest for older
age classes. Output from our models can be used to predict the spatial
age composition, which may be particularly useful for identifying large
cohorts (hake from 2008, 2010, 2014 and 2016 cohorts; sablefish from
2008, 2010, 2016, and 2021 cohorts). Using predictions of age-1
sablefish as an example, we also demonstrate how our models may be used
to forecast future bycatch risk in space, helping increase the
efficiency of fisheries. Our framework is applicable to a wide range of
survey types and platforms around the world, and supports the
development of spatially explicit management strategies and optimal
allocation of fishing effort.

\hypertarget{introduction}{%
\subsection{Introduction}\label{introduction}}

Species distribution models (SDMs) are widely used in marine ecology to
understand and predict the spatial distribution of species based on
environmental and demographic variables (Melo-Merino et al., 2020).
These models have informed management decisions for conservation and
resource allocation, especially in the face of changing environmental
conditions and human impacts, such as fishing pressure (Elith \&
Leathwick, 2009; Guisan \& Thuiller, 2005). Recent advances in SDM
applications in the marine environment have incorporated external
drivers such as temperature (Fredston et al., 2021) and habitat features
(Laman et al., 2018; Phillips et al., 2017). A second use of these
models has been to predict temporal trends in species biomass from
spatiotemporal fisheries independent survey data (Thorson et al., 2015).
Trends estimated from SDMs may be included in integrated population
models (stock assessments) for commercially fished species, or used to
identify risks to populations of conservation concern. Recent
developments in these models have included depth (Johnson et al., 2019)
and habitat variables (Cao et al., 2017) to improve trend estimation.

While many SDMs around the world have included variables such as
temperature, depth, or habitat, they have generally overlooked age
structure, and instead modeled either species occurrences or densities
across all age or size classes. Omitting information on age or size may
be partially dependent on data availability (ageing fish generally
requires sampling otoliths, which increases monetary costs of sampling
programs), but may also be driven by the modeling choices of analysts.
Much of the previous research around the world has demonstrated that age
structure is a fundamental characteristic of fish populations,
influencing spatial distribution and abundance due to mechanisms like
ontogenetic shifts, recruitment variability, and selective fishing
pressures (Rijnsdorp et al., 2018). Recent work has also highlighted
that age-based SDMs can improve understanding of environmental drivers
on commercially valuable populations (Ono et al., 2024). Ignoring
age-specific information in SDMs may lead to incomplete or inaccurate
predictions, especially for species with strong age-related spatial
gradients.

Incorporating age structure into SDMs may provide more accurate
predictions by capturing spatial and temporal variability related to
life history traits, age-specific habitat preferences, and population
dynamics. Fish often exhibit ontogenetic shifts, where younger
individuals occupy different habitats than adults due to factors like
predator avoidance, habitat preferences, and diet changes (Ciannelli et
al., 2004; Werner \& Gilliam, 1984). These shifts can create
age-specific distributions across habitats, particularly for species
that undergo distinct life history phases. Additionally, spatiallly
variable recruitment pulses and age-related migration patterns can
influence the spatial distribution of younger versus older fish. Fishing
selectivity often further accentuates these patterns by
disproportionately removing larger, older fish, which alters the
population structure and potentially shifts spatial distributions
(Hilborn \& Walters, 1992; Methot \& Wetzel, 2013). By incorporating
age, SDMs could better predict these dynamics and improve our
understanding of species-environment relationships.

Because of the recursive nature of population dynamics, e.g.~numbers at
age \emph{a} at time \emph{t} can be used to predict numbers at age
\emph{a}+1 at time \emph{t}+1, age-structured SDMs with predictive skill
also provide a tool for forward prediction, with considerable
implications for marine resource management. Reliable predictions of
species distributions can help managers anticipate and mitigate issues
such as bycatch risk by identifying regions with high densities of
specific age cohorts (Lewison et al., 2004). Forecasting the spatial
distribution of certain age classes can inform spatially explicit
management strategies and help optimize fishing effort to target
sustainable cohorts. For example, knowing the expected future locations
of specific age classes could aid in creating time-area closures that
protect vulnerable life stages or ensure harvests focus on more
resilient portions of the population. However, building robust
age-structured SDMs requires high-quality, long-term datasets with
consistent sampling methods across both age and space to ensure the
validity of these predictions.

Around the world, age and life history information is routinely
collected by a number of sampling platforms, including from fisheries
and fisheries independent surveys. While fisheries dependent data may be
used to inform age structured models, sampling from fisheries
independent surveys is typically more consistent in space and time. The
primary objective of our paper is to explore the utility of developing
age - structured SDMs from fisheries independent data. Focusing on two
species with contrasting life history characteristics in the North
Pacific, the semi-pelagic North Pacific hake (\emph{Merluccius
productus}) and longer lived benthic sablefish (\emph{Anoplopoma
fimbria}), we first develop a flexible framework for modeling numbers at
age. Second, we evaluate the ability of this framework to offer
predictive skill in future forecasts, and quantify how the predictive
relationship changes as a function of age. Finally, we demonstrate how
model output may be used to examine changing species compositions
through time, make maps of potential bycatch hotspots, and identify
areas with higher concentrations of mature fish as potential targets of
fishing effort.

\hypertarget{methods}{%
\subsection{Methods}\label{methods}}

\hypertarget{data}{%
\subsubsection{Data}\label{data}}

On the west coast of the USA, the West Coast Groundfish Bottom Trawl
Survey (WCGBTS) has been an annual survey conduced from 2003 - present.
The WCGBTS is designed to estimate the abundance, size, and age
composition of groundfish species important to commercial and
recreational fisheries found in near-bottom habitats on the west coast
of the USA (Keller et al., 2017). The survey effort is concentrated in
summer months and has been conducted annually since 2003 (here we use
the data through 2018; data are publicly available at
https://www.nwfsc.noaa.gov/data). Importantly, the random stratified
sampling design, spatial and seasonal coverage, effort, and gears have
remained relatively constant within the period we analyze. Like many
other surveys around the world, no WCGBTS survey occurred in 2020
because of the Covid-19 pandemic. Though the WCGBTS survey samples
hundreds of species, we concentrated our analysis on two of the well
sampled species from the WCGBTS, North Pacific hake and sablefish.
{[}more on biological gradients{]} While otoliths are sampled
continuously during the WCGBTS survey for a broad range of species,
otoliths are generally aged when species are being prioritized for stock
assessment by the Pacific Fishery Management Council (PFMC); full
assessments between species may occur irregularly and be sporadically
updated every 5 -- 10 years. Hake and sablefish represent exceptions, as
the hake stock is assessed annually by co-managers from the USA and
Canada (Grandin et al., 2024) and sablefish is frequently assessed
because of its high commercial value. For our analysis, we used hake
collected 2007 -- 2019 (mean 650.7 individuals sampled per year) and
sablefish collected 2003 -- 2023 (mean 1314.4 individuals sampled per
year).

\hypertarget{spatial-age-models}{%
\subsubsection{Spatial age models}\label{spatial-age-models}}

For each species, we filtered by sex to focus on females and truncated
ages to focus on those with the highest data availability; for hake this
included ages 1 -- 5, and for sablefish ages 0 -- 9 (data from these
ages represented 54\% and 75\% of the total aged fish, respectively).
For each species - age combination, we constructed a unique
spatiotemporal model fit to all years except the last available year for
each species. We adopted this approach instead of modeling ages
simultaneously (Ono et al., 2024), because this framework allows greater
flexibility (e.g.~unique autocorrelation and spatial variance parameters
by age). Individual fish were aggregated at the haul level, allowing us
to model counts at age with a Poisson distribution. We constructed a
spatiotemporal model as an extension of Generalized Linear Mixed Models
(GLMMs) such that the prediction in location \(s\) and time \(t\) can be
written as

\[
\begin{aligned}
log \left( u_{\boldsymbol{s},t} \right) &= \boldsymbol{\beta_{t}} + \omega_{\boldsymbol{s}} + \delta_{\boldsymbol{s},t}
\end{aligned}
\]

where \(\boldsymbol{\beta_{t}}\) represent time - varying intercepts
modeled as a random walk
\(\boldsymbol{\beta_{t}} \sim N \left( \boldsymbol{\beta_{t-1}}, \sigma_\beta \right)\),
the spatial field
\(\boldsymbol{\omega_{s}} \sim \operatorname{MVNormal} \left( \boldsymbol{0}, \boldsymbol{\Sigma}_\omega \right)\)
and the spatiotemporal fields \(\delta_{\boldsymbol{s},t}\) are modeled
as an AR(1) process
\(\boldsymbol{\delta}_{t} = \rho \boldsymbol{\delta}_{t-1} + \sqrt{1 - \rho^2} \boldsymbol{\epsilon_{t}}\),
where
\(\boldsymbol{\epsilon_{t}} \sim \operatorname{MVNormal} \left(\boldsymbol{0}, \boldsymbol{\Sigma}_{\epsilon} \right)\)

Spatial and spatiotemporal random fields were constructed as Gaussian
Markov random fields (GMRFs) using the stochastic partial differential
equation approach (SPDE) (Lindgren et al., 2011; Lindgren \& Rue, 2015).
The SPDE method models the spatial correlation between points as a
Matérn covariance function with smoothness parameter \(\nu = 1\).
Spatial meshes for all ages and species were constructed with a cutoff
distance of 50km (this distance controls the spacing of mesh vertices).
Parameter estimation was done using the sdmTMB software package
(Anderson et al., 2024) with R 4.3.1 (R Core Team, 2024). The sdmTMB
package relies on Template Model Builder (TMB)(Kristensen et al., 2016)
to quickly and efficiently maximize the marginal log likelihood using
auto-differentiation and the Laplace approximation to integrate out
random effects. Models were evaluated for convergence (positive-definite
Hessian matrix, and a maximum absolute log likelihood gradient
\textless{} 0.001) and residuals diagnostics were evaluated with the
DHARMa package (Hartig, 2022). Code and data to replicate our analysis
is publicly available on Github,
\url{https://github.com/ericward-noaa/spatial-age-models}.

\hypertarget{validating-future-predictive-ability}{%
\subsubsection{Validating future predictive
ability}\label{validating-future-predictive-ability}}

As a first validation, we leveraged the natural recursive element of our
data to quantify the ability of our models to predict the future
distribution by age. For each of the species - age models constructed
above, we made predictions to the survey locations in the following year
(e.g.~age 3 hake in years 2007 -- 2018 was used to predict the
distribution of age 4 hake in 2008 -- 2019). We related the predictions
\(\hat{u}_{\boldsymbol{s},t}\) to observations by multiplying
predictions by the total number of fish sampled for ageing,
\(\hat{n}_{s,t} = N_{s,t} \cdot \hat{u}_{\boldsymbol{s},t}\). We
quantified the relationship between predictions and observations by
fitting a simple Poisson GLM, where counts of fish of age \(a+1\) in
year \(t+1\) were treated as the response and
\(log(\lambda_{s,t}) = \beta_{0} + \beta_{1} \cdot log \left( \hat{n}_{s,t} \right)\).
The exponentiated slope parameter \(exp(\beta_{1})\) represents a change
in expected counts that would be expected from a 1-unit change in
\(log(\hat{n}_{s,t})\). As a diagnostic check and to understand whether
there was potential temporal patterns in the response, we fit a second
model as a GLMM with R package glmmTMB (Brooks et al., 2017); this
second model was identical to our Poisson GLM with random effects in
\(\beta_{0}\) and \(\beta_{1}\), where year was used as the grouping
variable.

\hypertarget{using-predictions-to-quantify-bycatch-risk}{%
\subsubsection{Using predictions to quantify bycatch
risk}\label{using-predictions-to-quantify-bycatch-risk}}

To illustrate potential benefits to fisheries management, we used output
from our models to generate spatiotemporal predictions of age-1
sablefish (from age-0 sablefish the year prior). Sablefish are part of
the DTS complex, which includes Dover sole \emph{Microstomus pacificus},
shortspine \emph{Sebastomus alascanus,} longspine thornyheads \emph{S.
altivelis}, and sablefish. This complex forms and important fishery on
the West Cast, and by-catch of sablefish can impact quotas for the other
three species, as well as the at-sea hake fishery. For exmaple, in 2017,
the at-sea hake fishery had 50 mt of sablefish quota set aside in the
area north of 36 °N, but the fleet caught more than 3x as much, promting
in-season warnings to alter fishing behavior.

Gridded predictions from the Poisson GLM represent the expected number
of fish per unit of effort in each location. To turn these into a
potentially more useful measure of risk, we calculate the predicted
total age-1 sablefish catch per unit of effort (CPUE) within a given
radius of major fishing ports on the US west coast. For consistency with
previous work , we used a radius of 282 km (Leising et al., 2024). These
predictions of port-level CPUE were then summarized as annual time
series (1-step ahead forecasts).

\hypertarget{results}{%
\subsection{Results}\label{results}}

One useful output from our spatiotemporal models is the estimated field
spatial f, \(\omega_s\), representing the average spatial distribution
across years. Results from our model of North Pacific hake show a
concentration of age-1 hake near the coast, and more of latitudinal
north - south break for 2 - 3 year old hake (with higher occurrences in
the South, Fig.~\ref{fig-hake-spatial-anomaly}). As expected, age 4 hake
appear to have a northern distribution, though the gradient is less
clear than for ages 2 - 3. For sablefish, our models estimate a
concentration of age 0 - 2 sablefish in shallow water, with a deeper
distribution by age 3 (Fig.~\ref{fig-sablefish-spatial-anomaly}).
Latitudinally, there are several areas in central Oregon that appear to
have consistently lower occurrences than average for older age classes
(Fig.~\ref{fig-sablefish-spatial-anomaly}).

Estimates of age composition by year may be useful in tracking the
distribution of cohorts through time, as well as examining variability
in distributions of particular age classes over time. For example, for
both hake (Figs.~\ref{fig-hake-spatial-composition} \&
\ref{fig-hake-spatial-composition-all}) and sablefish
(Figs.~\ref{fig-sablefish-spatial-composition} \&
\ref{fig-sablefish-spatial-composition-all}), distinct patches of high
age-0 abundance tended to disperse as the fish aged such that spatial
distributions for older age-classes were less distinct. Additionally,
results from our models can be used to project the spatial distributions
of strong hake cohorts (2008, 2010, 2014, 2016;
Fig.~\ref{fig-hake-spatial-composition}) and strong sablefish cohorts
(2008, 2010, 2016, 2021, Fig.~\ref{fig-sablefish-spatial-composition}).
These predictions are also useful in revealing subtle differences
between strong cohort years. For example, the 2014 cohort of Pacific
hake had a larger fraction of age-1 individuals distributed in the
northern part of the domain (along the coast of Washington state), while
other cohorts had hotspots in the center and southern portion of the
domain. Similarly for sablefish, the 2016 cohort was concentrated in the
north (age-1 individuals in 2011), versus other cohorts that had a more
coastwide distribution. In fact, age-1 fish were uncommon south of
approximately Cape Mendocino (40 °N) from 2015 - 2020
(Figure~\ref{fig-sablefish-spatial-composition-all}), concurrent with
`the Blob' and other marine heatwaves.

When comparing the spatial and spatiotemporal parameters across ages and
species, we find that for most ages, there is a decrease in the
estimated spatial and spatiotemporal variances as well as a decrease in
the spatial range with increasing age
(Fig.~\ref{fig-spatial-parameters}). Spatial and spatiotemporal
variances control the magnitude of the peaks and valleys in estimated
latent fields, meaning that spatial fields estimated for older fish are
less variable. Similarly, the spatial range determines the distance at
which locations are functionally independent (for the Matern, \(\rho\)
\textasciitilde{} 0.13). The decline in range with age implies that
spatial distributions also become more concentrated as a function of age
(Fig.~\ref{fig-spatial-parameters}).

Our 1-step ahead validation analyses indicate that there is a stronger
association between older ages compared to younger ages, and this effect
is strongest for sablefish (Fig.~\ref{fig-glm-coefficients}). When
comparing our initial Poisson GLM to a more complicated GLMM with random
year effects in the intercept and slope, we found little support for
shared correlations across age classes of a given species. Shared
correlations would indicate for example that age-2 and age-3 sablefish
may be more or less predictable in certain years (indicating that
information about one age class may be useful in predicting another).
With the exception of a strong positive correlation between age-0 and
age-1 sablefish, most correlations between age classes were smaller than
0.3 in magnitude (Figure~\ref{fig-glm-coefficients-time}).

Our forecasts of age-1 sablefish CPUE illustrates the variability of
potential bycatch risk in space and time. In most years for example,
age-1 CPUE is lowest in ports in the state of California (Eureka, Fort
Bragg, Morro Bay) and highest in several ports in the state of Oregon
(Astoria, Coos Bay, Fig.~\ref{fig-bycatch-risk}). These results also
illustrate the high temporal variability in age-1 CPUE, with peaks
occuring in 2009, 2014, 2017, and 2021, Figure~\ref{fig-bycatch-risk}).
During peak years, there is also variability among ports -- forecasts of
age-1 CPUE were similar for Astoria and Coos Bay in 2021 for instance,
but the potential bycatch risk near Astoria was higher in 2017
(Fig.~\ref{fig-bycatch-risk}).

\hypertarget{discussion}{%
\subsection{Discussion}\label{discussion}}

In this study, we developed an age-structured species distribution
modeling (SDM) framework to explore spatial age composition and predict
age-specific distributions of two commercially important fish species
with different life history characteristics: North Pacific hake and
sablefish. Our approach integrates spatiotemporal dynamics, age
structure, and forward-predictive capabilities, providing a useful tool
for fisheries management. By examining age-specific patterns in spatial
distributions and validating predictive skill, we highlight the
potential utility of incorporating age structure into SDMs for advancing
spatially explicit fisheries management.

\hypertarget{age-structured-spatial-patterns}{%
\paragraph{\texorpdfstring{\emph{Age-Structured Spatial
Patterns}}{Age-Structured Spatial Patterns}}\label{age-structured-spatial-patterns}}

Results from our age-structured spatial models underscore the importance
of accounting for age in SDMs. The spatial patterns observed for both
hake and sablefish were consistent with their distinct life histories,
supporting the idea that ontogenetic shifts significantly influence fish
distributions. Younger age classes of hake and sablefish exhibited
broader distributions generally in shallower areas, while older
individuals were more concentrated offshore (sablefish in deeper waters,
hake in more northern regions). The deeper distribution of older
sablefish reflects their preference for benthic habitats and an
ontogenetic shift to deeper waters, while the observed north-south
gradient in older hake aligns with known migratory behaviors influenced
by spawning and feeding grounds (Agostini et al., 2006).

\hypertarget{predictive-skill-and-cohort-tracking}{%
\paragraph{\texorpdfstring{\emph{Predictive skill and cohort
tracking}}{Predictive skill and cohort tracking}}\label{predictive-skill-and-cohort-tracking}}

Our validation analyses revealed that the predictive skill of our
framework increases with age, particularly for sablefish. This result is
intuitive given that older age classes generally exhibit more stable
spatial patterns due to reduced recruitment variability and more
predictable habitat preferences. In contrast, the lower predictive skill
for younger age classes reflects their higher spatiotemporal
variability, likely driven by recruitment pulses and early-life survival
dynamics. Tracking cohorts through time using our framework proved
effective in identifying strong year classes
(Figs.~\ref{fig-hake-spatial-composition} \&
\ref{fig-hake-spatial-composition}). Identifying these cohorts may be
useful to stock assessment and management -- for instance, understanding
the spatial distribution of strong cohorts could be used to inform
spatial harvest allocation. Additionally, the ability to project spatial
age composition provides a tool for assessing how environmental changes,
such as ocean warming or habitat shifts, may impact the distribution of
key cohorts over time.

\hypertarget{managing-bycatch-risk}{%
\paragraph{\texorpdfstring{\emph{Managing bycatch
risk}}{Managing bycatch risk}}\label{managing-bycatch-risk}}

Our case study using age-1 sablefish demonstrates the practical utility
of our modeling for mitigating bycatch risk. Forecasts of CPUE for age-1
sablefish revealed significant temporal and spatial variability, with
higher bycatch risk near Oregon ports in certain years. Importantly,
these forecasts could be provided to industry partners and fisheries
managers 6 -- 12 months before fishing occurs. Such information could
help fishers avoid areas associated with high bycatch risk, or be used
to adjust the spatial distribution of quota in certain years. The
observed peaks in age-1 sablefish CPUE during certain years suggest the
influence of recruitment pulses, which are thought to be tied to
environmental drivers (Tolimieri et al., 2018; Tolimieri \& Haltuch,
2023). Future efforts could also incorporating environmental covariates
to enhance its predictive accuracy and further support climate-resilient
fisheries management.

\hypertarget{need-for-reliable-age-data}{%
\paragraph{Need for reliable age
data}\label{need-for-reliable-age-data}}

Perhaps the biggest challenge in implementing our modeling framework in
the future is that these models depend on reliable age data being
collected. While the WCGBTS provides a robust dataset, gaps in sampling
(e.g., the absence of a 2020 survey due to the COVID-19 pandemic) and
variability in otolith aging efforts over time could introduce
uncertainty. The costs of ship-based surveys, such as the WCGBTS survey,
have increased around the world, and numerous efforts are being
implemented to make these programs more efficient (ICES, 2020). While
age samples are more expensive to collect than length information, there
is value from the increased precision associated with age data.

\hypertarget{future-directions}{%
\paragraph{Future directions}\label{future-directions}}

Our age-structured SDM framework represents a significant advance in
spatial modeling, offering novel insights into the dynamics of fish
populations and their age-specific distributions. From a management
perspective, age-structured SDMs offer a powerful tool for addressing
critical challenges, including bycatch mitigation, spatial planning, and
sustainable harvesting. By providing high-resolution predictions of
spatial age composition, these models enable managers to align fishing
effort with ecological and economic goals, ensuring the long-term
sustainability of marine resources. Future research could extend this
framework by incorporating environmental covariates, such as temperature
or oxygen, to account for dynamic habitat preferences, or fisheries
removals to account for local depletion (Ono et al., 2016).
Additionally, exploring the potential for multi-species models could
provide insights into ecosystem-level interactions and their influence
on spatial distributions.

\hypertarget{acknowledgments}{%
\subsubsection{Acknowledgments}\label{acknowledgments}}

We thank Owen Liu and XX reviewers for constructive comments on the
manuscript.

\break
\clearpage

\hypertarget{references}{%
\subsection{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-agostini2006}{}}%
Agostini, V. N., Francis, R. C., Hollowed, A. B., Pierce, S. D., Wilson,
C., \& Hendrix, A. N. (2006). {The relationship between Pacific hake
({\emph{Merluccius productus}}) distribution and poleward subsurface
flow in the California Current System}. \emph{Canadian Journal of
Fisheries and Aquatic Sciences}, \emph{63}(12), 2648--2659.
\url{https://doi.org/10.1139/f06-139}

\leavevmode\vadjust pre{\hypertarget{ref-Anderson2024}{}}%
Anderson, S. C., Ward, E. J., English, P. A., Barnett, L. A. K., \&
Thorson, J. T. (2024). {sdmTMB: An R package for fast, flexible, and
user-friendly generalized linear mixed effects models with spatial and
spatiotemporal random fields}. \emph{bioRxiv}.
\url{https://doi.org/10.1101/2022.03.24.485545}

\leavevmode\vadjust pre{\hypertarget{ref-brooks2017}{}}%
Brooks, E., M., Kristensen, K., Benthem, A. van, K. J.and Magnusson,
Berg, C. W., Nielsen, A., Skaug, H. J., Mächler, M., \& Bolker, M., B.
(2017). {glmmTMB balances speed and flexibility among packages for
zero-inflated generalized linear mixed modeling}. \emph{The R Journal},
\emph{9}(2), 378. \url{https://doi.org/10.32614/rj-2017-066}

\leavevmode\vadjust pre{\hypertarget{ref-cao2017}{}}%
Cao, J., Thorson, J. T., Richards, R. A., \& Chen, Y. (2017).
{Spatiotemporal index standardization improves the stock assessment of
northern shrimp in the Gulf of Maine}. \emph{Canadian Journal of
Fisheries and Aquatic Sciences}, \emph{74}(11), 1781--1793.
\url{https://doi.org/10.1139/cjfas-2016-0137}

\leavevmode\vadjust pre{\hypertarget{ref-ciannelli2004}{}}%
Ciannelli, L., Chan, K.-S., Bailey, K. M., \& Stenseth, N. Chr. (2004).
{Nonadditive effects of the environment on the survival of a large
marine fish population}. \emph{Ecology}, \emph{85}(12), 3418--3427.
\url{https://doi.org/10.1890/03-0755}

\leavevmode\vadjust pre{\hypertarget{ref-elith2009}{}}%
Elith, J., \& Leathwick, J. R. (2009). {Species distribution models:
Ecological explanation and prediction across space and time}.
\emph{Annual Review of Ecology, Evolution, and Systematics},
\emph{40}(1), 677--697.
\url{https://doi.org/10.1146/annurev.ecolsys.110308.120159}

\leavevmode\vadjust pre{\hypertarget{ref-fredston2021}{}}%
Fredston, A., Pinsky, M., Selden, R. L., Szuwalski, C., Thorson, J. T.,
Gaines, S. D., \& Halpern, B. S. (2021). {Range edges of North American
marine species are tracking temperature over decades}. \emph{Global
Change Biology}, \emph{27}(13), 3145--3156.
\url{https://doi.org/10.1111/gcb.15614}

\leavevmode\vadjust pre{\hypertarget{ref-Grandin2024}{}}%
Grandin, C. J., Johnson, K. F., M., E. A., \& M., B. A. (2024).
\emph{{Status of the Pacific Hake (whiting) stock in U.S. and Canadian
waters in 2024. Prepared by the Joint Technical Committee of the U.S.
and Canada Pacific Hake/Whiting Agreement, National Marine Fisheries
Service and Fisheries and Oceans Canada. 246 p.}}

\leavevmode\vadjust pre{\hypertarget{ref-guisan2005}{}}%
Guisan, A., \& Thuiller, W. (2005). {Predicting species distribution:
offering more than simple habitat models}. \emph{Ecology Letters},
\emph{8}(9), 993--1009.
\url{https://doi.org/10.1111/j.1461-0248.2005.00792.x}

\leavevmode\vadjust pre{\hypertarget{ref-dharma2022}{}}%
Hartig, F. (2022). \emph{{DHARMa: Residual diagnostics for hierarchical
(multi-level / mixed) regression models}}.
\url{https://CRAN.R-project.org/package=DHARMa}

\leavevmode\vadjust pre{\hypertarget{ref-hilborn1992}{}}%
Hilborn, R., \& Walters, C. J. (1992). \emph{{Quantitative fisheries
stock assessment}}. Springer US.
\url{https://doi.org/10.1007/978-1-4615-3598-0}

\leavevmode\vadjust pre{\hypertarget{ref-ices2020}{}}%
ICES. (2020). {Workshop on unavoidable survey effort reduction
(WKUSER)}. \emph{ICES}. \url{https://doi.org/10.17895/ICES.PUB.7453}

\leavevmode\vadjust pre{\hypertarget{ref-johnson2019}{}}%
Johnson, K. F., Thorson, J. T., \& Punt, A. E. (2019). {Investigating
the value of including depth during spatiotemporal index
standardization}. \emph{Fisheries Research}, \emph{216}, 126--137.
\url{https://doi.org/10.1016/j.fishres.2019.04.004}

\leavevmode\vadjust pre{\hypertarget{ref-keller2017}{}}%
Keller, A. A., Wallace, J. R., \& Methot, R. D. (2017). \emph{{The
Northwest Fisheries Science Center's West coast groundfish bottom trawl
survey: History design, and description}}. NOAA, U.S. Department of
Commerce. \url{https://doi.org/10.7289/V5/TM-NWFSC-136}

\leavevmode\vadjust pre{\hypertarget{ref-kristensen2016}{}}%
Kristensen, K., Nielsen, A., Berg, C. W., Skaug, H., \& Bell, B. M.
(2016). {TMB: Automatic differentiation and laplace approximation}.
\emph{Journal of Statistical Software}, \emph{70}(5).
\url{https://doi.org/10.18637/jss.v070.i05}

\leavevmode\vadjust pre{\hypertarget{ref-laman2018}{}}%
Laman, E. A., Rooper, C. N., Turner, K., Rooney, S., Cooper, D. W., \&
Zimmermann, M. (2018). {Using species distribution models to describe
essential fish habitat in Alaska}. \emph{Canadian Journal of Fisheries
and Aquatic Sciences}, \emph{75}(8), 1230--1255.
\url{https://doi.org/10.1139/cjfas-2017-0181}

\leavevmode\vadjust pre{\hypertarget{ref-leising2024cciea}{}}%
Leising, A., M., H., N., T., G., W., \& A., H. (2024).

\leavevmode\vadjust pre{\hypertarget{ref-lewison2004}{}}%
Lewison, R. L., Freeman, S. A., \& Crowder, L. B. (2004). {Quantifying
the effects of fisheries on threatened species: the impact of pelagic
longlines on loggerhead and leatherback sea turtles}. \emph{Ecology
Letters}, \emph{7}(3), 221--231.
\url{https://doi.org/10.1111/j.1461-0248.2004.00573.x}

\leavevmode\vadjust pre{\hypertarget{ref-lindgren2015}{}}%
Lindgren, F., \& Rue, H. (2015). {Bayesian spatial modelling
with{\emph{R}}-{\textbf{INLA}}}. \emph{Journal of Statistical Software},
\emph{63}(19). \url{https://doi.org/10.18637/jss.v063.i19}

\leavevmode\vadjust pre{\hypertarget{ref-lindgren2011}{}}%
Lindgren, F., Rue, H., \& Lindström, J. (2011). {An explicit link
between Gaussian fields and Gaussian Markov Random Fields: The
dtochastic partial fifferential rquation spproach}. \emph{Journal of the
Royal Statistical Society Series B: Statistical Methodology},
\emph{73}(4), 423--498.
\url{https://doi.org/10.1111/j.1467-9868.2011.00777.x}

\leavevmode\vadjust pre{\hypertarget{ref-melo-merino2020}{}}%
Melo-Merino, S. M., Reyes-Bonilla, H., \& Lira-Noriega, A. (2020).
{Ecological niche models and species distribution models in marine
environments: A literature review and spatial analysis of evidence}.
\emph{Ecological Modelling}, \emph{415}, 108837.
\url{https://doi.org/10.1016/j.ecolmodel.2019.108837}

\leavevmode\vadjust pre{\hypertarget{ref-methot2013}{}}%
Methot, R. D., \& Wetzel, C. R. (2013). {Stock synthesis: A biological
and statistical framework for fish stock assessment and fishery
management}. \emph{Fisheries Research}, \emph{142}, 86--99.
\url{https://doi.org/10.1016/j.fishres.2012.10.012}

\leavevmode\vadjust pre{\hypertarget{ref-ono2024}{}}%
Ono, K., Katara, I., Eliasen, S. K., Broms, C., Campbell, A., dos Santos
Schmidt, T. C., Egan, A., Hølleland, S. N., Jacobsen, J. A., Jansen, T.,
Mackinson, S., Mousing, E. A., Nash, R. D. M., Nikolioudakis, N.,
Nnanatu, C., Nøttestad, L., Singh, W., Slotte, A., Wieland, K., \&
Olafsdottir, A. H. (2024). {Effect of environmental drivers on the
spatiotemporal distribution of mackerel at age in the Nordic Seas during
2010{-}20}. \emph{ICES Journal of Marine Science}, \emph{81}(7),
1282--1294. \url{https://doi.org/10.1093/icesjms/fsae087}

\leavevmode\vadjust pre{\hypertarget{ref-ono2016space}{}}%
Ono, K., Shelton, A. O., Ward, E. J., Thorson, J. T., Feist, B. E., \&
Hilborn, R. (2016). {Space-time investigation of the effects of fishing
on fish populations}. \emph{Ecological Applications}, \emph{26}(2),
392--406.

\leavevmode\vadjust pre{\hypertarget{ref-phillips2017}{}}%
Phillips, N. D., Reid, N., Thys, T., Harrod, C., Payne, N. L., Morgan,
C. A., White, H. J., Porter, S., \& Houghton, J. D. R. (2017). {Applying
species distribution modelling to a data poor, pelagic fish complex: the
ocean sunfishes}. \emph{Journal of Biogeography}, \emph{44}(10),
2176--2187. \url{https://doi.org/10.1111/jbi.13033}

\leavevmode\vadjust pre{\hypertarget{ref-rcore2024}{}}%
R Core Team. (2024). \emph{{R: A language and environment for
statistical computing}}. R Foundation for Statistical Computing.
\url{https://www.R-project.org/}

\leavevmode\vadjust pre{\hypertarget{ref-rijnsdorp2018}{}}%
Rijnsdorp, A. D., Bolam, S. G., Garcia, C., Hiddink, J. G., Hintzen, N.
T., Denderen, P. D. van, \& Kooten, T. van. (2018). {Estimating
sensitivity of seabed habitats to disturbance by bottom trawling based
on the longevity of benthic fauna}. \emph{Ecological Applications},
\emph{28}(5), 1302--1312. \url{https://doi.org/10.1002/eap.1731}

\leavevmode\vadjust pre{\hypertarget{ref-thorson2015}{}}%
Thorson, J. T., Shelton, A. O., Ward, E. J., \& Skaug, H. J. (2015).
{Geostatistical delta-generalized linear mixed models improve precision
for estimated abundance indices for West Coast groundfishes}. \emph{ICES
Journal of Marine Science}, \emph{72}(5), 1297--1310.
\url{https://doi.org/10.1093/icesjms/fsu243}

\leavevmode\vadjust pre{\hypertarget{ref-tolimieri2023sea}{}}%
Tolimieri, N., \& Haltuch, M. (2023). {Sea-level index of recruitment
variability improves assessment model performance for sablefish
/textit{Anoplopoma fimbria}}. \emph{Canadian Journal of Fisheries and
Aquatic Sciences}, \emph{80}(6), 1006--1016.

\leavevmode\vadjust pre{\hypertarget{ref-tolimieri2018}{}}%
Tolimieri, N., Haltuch, M. A., Lee, Q., Jacox, M. G., \& Bograd, S. J.
(2018). {Oceanographic drivers of sablefish recruitment in the
California Current}. \emph{Fisheries Oceanography}, \emph{27}(5),
458--474. \url{https://doi.org/10.1111/fog.12266}

\leavevmode\vadjust pre{\hypertarget{ref-werner1984}{}}%
Werner, E. E., \& Gilliam, J. F. (1984). {The ontogenetic niche and
species interactions in size-structured populations}. \emph{Annual
Review of Ecology and Systematics}, \emph{15}(1), 393--425.
\url{https://doi.org/10.1146/annurev.es.15.110184.002141}

\end{CSLReferences}

\break
\clearpage

\hypertarget{figures}{%
\subsection{Figures}\label{figures}}

\begin{figure}

{\centering \includegraphics[width=3in,height=\textheight]{plots/hake_spatial_anomaly.png}

}

\caption{\label{fig-hake-spatial-anomaly}Estimated spatial anomalies for
Pacific hake ages 1 -- 5. These fields represent the average anomalies
across all years, 2003 -- 2023.}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=4in,height=\textheight]{plots/sablefish-spatial-anomaly.png}

}

\caption{\label{fig-sablefish-spatial-anomaly}Estimated spatial
anomalies for sablefish ages 0 -- 9. These fields represent the average
anomalies across all years, 2003 -- 2023.}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=6.2in,height=\textheight]{plots/Pacific hake-follow-age-class.png}

}

\caption{\label{fig-hake-spatial-composition}Estimated spatial catch per
unit effort (CPUE, kg per km2) for Pacific hake; rows represent strong
cohorts in our dataset (2008, 2010, 2014, 2016). CPUE is standardized to
1.0 across cohorts for visualization purposes. Full predictions for all
cohorts are in the Supplementary Information.}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=6.2in,height=\textheight]{plots/sablefish-follow-age-class.png}

}

\caption{\label{fig-sablefish-spatial-composition}Estimated spatial
catch per unit effort (CPUE, kg per km2) for sablefish; each row
represents a strong cohorts in our dataset (2008, 2010, 2016, 2021).
CPUE is standardized to 1.0 across cohorts for visualization purposes.
Full predictions for all cohorts are in the Supplementary Information.}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=3.5in,height=\textheight]{plots/glm_coefficients.png}

}

\caption{\label{fig-glm-coefficients}Estimated coefficients relating
predicted densities of age a fish in year t to observed numbers the
following year. The coefficients and 95\% confidence intervals (vertical
lines) are estimated in log space and presented here in normal space.}

\end{figure}

\clearpage

\newpage

\begin{figure}

{\centering \includegraphics[width=5in,height=\textheight]{plots/sablefish_spatial_risk.png}

}

\caption{\label{fig-bycatch-risk}Estimated spatial catch per unit effort
(CPUE, kg per km2) for age-1 sablefish and port-level forecasts of age-1
CPUE within a radius of 232 km. In the upper plot CPUE is standardized
to 1.0 across cohorts for visualization purposes.}

\end{figure}

\break
\clearpage

\hypertarget{supplementary-material}{%
\section{Supplementary material}\label{supplementary-material}}

\begin{figure}

{\centering \includegraphics[width=5in,height=\textheight]{plots/spatial_parameters.png}

}

\caption{\label{fig-spatial-parameters}Estimated spatial parameters
(range, and spatial and spatiotemporal standard deviations). Points
represent mean estimates; lines represent 95\% confidence intervals}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=6.2in,height=\textheight]{plots/Pacific hake-age-class-year.png}

}

\caption{\label{fig-hake-spatial-composition-all}Estimated spatial catch
per unit effort (CPUE, kg per km2) for Pacific hake; rows represent ages
(1-5). CPUE has been scaled for each cohort for visualization purposes.}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=6.2in,height=\textheight]{plots/sablefish-age-class-year.png}

}

\caption{\label{fig-sablefish-spatial-composition-all}Estimated spatial
catch per unit effort (CPUE, kg per km2) for sablefish; rows represent
ages (0-4). CPUE has been scaled for each cohort for visualization
purposes.}

\end{figure}

\newpage

\begin{figure}

{\centering \includegraphics[width=7in,height=\textheight]{plots/glm_coefficients_time.png}

}

\caption{\label{fig-glm-coefficients-time}Time-varying coefficients
(with random effects in year) relating predicted densities of age a fish
in year t to observed numbers the following year.}

\end{figure}

\end{document}

### 2025-01-30 03_poisson_regression.R
cleaning

@@ -72,7 +72,7 @@ for(spp in 1:2) {
 ggplot(all_output, aes(age, exp(mean), col = species)) + 
   geom_hline(aes(yintercept=1), col="red", alpha=0.3) + 
   geom_pointrange(aes(ymin = exp(lower), ymax = exp(upper)), 
-                  position=position_dodge(0.4), size=0.1) + 
+                  position=position_dodge(0.4), size=0.3) + 
   geom_point(position=position_dodge(0.4), size = 0.1) + 
   scale_color_viridis_d(option="magma", begin=0.2, end = 0.8, name="Species") +
   scale_x_continuous(breaks = seq(0,9,1), minor_breaks = seq(0,9,1)) + 
@@ -85,7 +85,7 @@ ggplot(all_output, aes(age, exp(mean), col = species)) +
     legend.title = element_text(size = 8),
     legend.text = element_text(size = 8),
     legend.key.size = unit(1,'lines')) 
-
+ggsave("plots/final_plots/05_glm_coefficients.jpg", width=6, height = 5)
 ggsave("plots/glm_coefficients.png", width=3.5, height = 2)
 
 
@@ -99,6 +99,7 @@ dplyr::filter(all_coefs) |>
   theme(strip.background = element_rect(fill="white")) + 
   facet_wrap(~species, ncol = 1, scale="free_y")
 ggsave("plots/glm_coefficients_time.png", width=7, height = 5)
+ggsave("plots/final_plots/S4_glm_coefficients_time.jpg", width=6, height = 5)
 
 library(tidyr)
 library(ggplot2)

### 2025-01-30 05_spatial_bycatch_risk_sablefish.R

@@ -97,4 +97,5 @@ p2 <- ggplot(distances_long, aes(year, value, group = Name, col = Name)) +
 p3 <- gridExtra::grid.arrange(p1, p2, ncol = 1)
 
 ggsave(p3, filename = "plots/sablefish_spatial_risk-old.png", width = 7, height = 6)  
+ggsave(p3, filename = "plots/final_plots/06_sablefish_spatial_risk-old.jpg", width = 7, height = 6)  
 

### 2025-01-30 08_plots_spatial_anomaly_pretty.R

@@ -66,7 +66,7 @@ dist_map2 <- function(dfile=sablefish,
   library(rnaturalearth)
   library(rnaturalearthdata)
   states <- map_data("state")
-  west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
+  west_coast <- subset(states, region %in% c("california", "oregon", "washington","idaho","nevada","montana"))
   
   # adjust axis
   dfile$yrs = axis.mod*(no_ages - dfile$age - 1)
@@ -87,7 +87,7 @@ dist_map2 <- function(dfile=sablefish,
     ggplot(data=west_coast, aes(x = long, y = lat), 
            fill = grey(0.9), color = "black") +
     geom_polygon(aes(x = long, y = lat, group=group), 
-                 fill = grey(0.9), color = "black") +
+                 fill = "wheat", color = "grey70") +
     geom_point(data=dfile , 
                aes(lon2, lat, color=Z), size=0.01) +
     # might need to adjust here based on the scale of the data, esp 'trans' term.
@@ -115,9 +115,11 @@ fig_dir = paste0(getwd(),"/plots/")
 
 dist_map2(sablefish,no_ages = 10, xlim = c(-152, -117))
 ggsave( paste0(fig_dir,"sablefish-spatial-anomaly.png"), width=4, height = 2)
+ggsave( paste0(fig_dir,"final_plots/02_sablefish-spatial-anomaly.jpg"), width=7, height = 6)
 
 dist_map2(hake,no_ages = 6, xlim = c(-140, -117))
 ggsave( paste0(fig_dir,"hake_spatial_anomaly.png"), width=3, height = 2)
+ggsave( paste0(fig_dir,"final_plots/01_hake_spatial_anomaly.jpg"), width=7, height = 6)
 
 
 

### 2025-01-30 09_plots_spatial_bycatch_risk_sablefish_pretty.R

@@ -29,7 +29,7 @@ xmin = floor( min(sablefish$lon2))
 
 # background map info ##########################################################
 states <- map_data("state")
-west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
+west_coast <- subset(states, region %in% c("california", "oregon", "washington","idaho","nevada","montana"))
 ################################################################################
 
 # plot function
@@ -59,7 +59,7 @@ dist_maps <- function(dfile,
     ggplot(data=west_coast, aes(x = long, y = lat), 
            fill = grey(0.9), color = "black") +
     geom_polygon(aes(x = long, y = lat, group=group), 
-                 fill = grey(0.9), color = "black") +
+                 fill = "wheat", color = "grey70") +
     geom_point(data=dfile , 
                aes(lon2, lat, color=p), size=0.01) +
     # might need to adjust here based on the scale of the data, esp 'trans' term.
@@ -143,4 +143,7 @@ png("plots/sablefish_spatial_risk.png", width=5, heigh=5, res=300, units = 'in')
 patchwork::plot_layout(p1 / p2)
 dev.off()
 
+jpeg("plots/final_plots/06_sablefish_spatial_risk.jpg", width=7, height=7, res=300, units = 'in')
+patchwork::plot_layout(p1 / p2)
+dev.off()
 

### 2025-01-30 paper.tex
paper

@@
 \makeatother
+% definitions for citeproc citations
+\NewDocumentCommand\citeproctext{}{}
+\NewDocumentCommand\citeproc{mm}{%
+  \begingroup\def\citeproctext{#2}\cite{#1}\endgroup}
+\makeatletter
+ % allow citations to break across lines
+ \let\@cite@ofmt\@firstofone
+ % avoid brackets around text for \cite:
+ \def\@biblabel#1{}
+ \def\@cite#1#2{{#1\if@tempswa , #2\fi}}
+\makeatother
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
+\newenvironment{CSLReferences}[2] % #1 hanging-indent, #2 entry-spacing
+ {\begin{list}{}{%
+  \setlength{\itemindent}{0pt}
+  \setlength{\leftmargin}{0pt}
+  \setlength{\parsep}{0pt}
   % turn on hanging indent if param 1 is 1
   \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+   \setlength{\leftmargin}{\cslhangindent}
+   \setlength{\itemindent}{-1\cslhangindent}
   \fi
   % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
+  \setlength{\itemsep}{#2\baselineskip}}}
+ {\end{list}}
 \usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLBlock}[1]{\hfill\break\parbox[t]{\linewidth}{\strut\ignorespaces#1\strut}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{\strut#1\strut}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{\strut#1\strut}}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 \renewcommand\Affilfont{\small}
-\makeatletter
-\makeatother
-\makeatletter
-\makeatother
 \makeatletter
@@
 \makeatletter
+\makeatother
+\makeatletter
 \@ifpackageloaded{caption}{}{\usepackage{caption}}
 \@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
-\makeatother
-\makeatletter
-\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
-\makeatother
-\makeatletter
-\@ifundefined{shadecolor}{\definecolor{shadecolor}{rgb}{.97, .97, .97}}
-\makeatother
-\makeatletter
-\makeatother
-\makeatletter
 \makeatother
@@
 \fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\usepackage{bookmark}
+
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
@@
 \maketitle
-\ifdefined\Shaded\renewenvironment{Shaded}{\begin{tcolorbox}[frame hidden, interior hidden, borderline west={3pt}{0pt}{shadecolor}, sharp corners, breakable, boxrule=0pt, enhanced]}{\end{tcolorbox}}\fi
-
-\hypertarget{abstract}{%
-\subsection{\texorpdfstring{\emph{Abstract}}{Abstract}}\label{abstract}}
+
+\subsection{\texorpdfstring{\emph{Abstract}}{Abstract}}\label{abstract}
 
@@
 
-\hypertarget{introduction}{%
-\subsection{Introduction}\label{introduction}}
+\subsection{Introduction}\label{introduction}
 
@@
 
-\hypertarget{methods}{%
-\subsection{Methods}\label{methods}}
-
-\hypertarget{data}{%
-\subsubsection{Data}\label{data}}
+\subsection{Methods}\label{methods}
+
+\subsubsection{Data}\label{data}
 
@@
 
-\hypertarget{spatial-age-models}{%
-\subsubsection{Spatial age models}\label{spatial-age-models}}
+\subsubsection{Spatial age models}\label{spatial-age-models}
 
@@
 
-\hypertarget{validating-future-predictive-ability}{%
 \subsubsection{Validating future predictive
-ability}\label{validating-future-predictive-ability}}
+ability}\label{validating-future-predictive-ability}
 
@@
 
-\hypertarget{using-predictions-to-quantify-bycatch-risk}{%
 \subsubsection{Using predictions to quantify bycatch
-risk}\label{using-predictions-to-quantify-bycatch-risk}}
+risk}\label{using-predictions-to-quantify-bycatch-risk}
 
@@
 
-\hypertarget{results}{%
-\subsection{Results}\label{results}}
+\subsection{Results}\label{results}
 
@@
 
-\hypertarget{discussion}{%
-\subsection{Discussion}\label{discussion}}
+\subsection{Discussion}\label{discussion}
 
@@
 
-\hypertarget{age-structured-spatial-patterns}{%
 \paragraph{\texorpdfstring{\emph{Age-Structured Spatial
-Patterns}}{Age-Structured Spatial Patterns}}\label{age-structured-spatial-patterns}}
+Patterns}}{Age-Structured Spatial Patterns}}\label{age-structured-spatial-patterns}
 
@@
 
-\hypertarget{predictive-skill-and-cohort-tracking}{%
 \paragraph{\texorpdfstring{\emph{Predictive skill and cohort
-tracking}}{Predictive skill and cohort tracking}}\label{predictive-skill-and-cohort-tracking}}
+tracking}}{Predictive skill and cohort tracking}}\label{predictive-skill-and-cohort-tracking}
 
@@
 
-\hypertarget{managing-bycatch-risk}{%
 \paragraph{\texorpdfstring{\emph{Managing bycatch
-risk}}{Managing bycatch risk}}\label{managing-bycatch-risk}}
+risk}}{Managing bycatch risk}}\label{managing-bycatch-risk}
 
@@
 
-\hypertarget{need-for-reliable-age-data}{%
-\paragraph{Need for reliable age
-data}\label{need-for-reliable-age-data}}
+\paragraph{Need for reliable age data}\label{need-for-reliable-age-data}
 
@@
 
-\hypertarget{future-directions}{%
-\paragraph{Future directions}\label{future-directions}}
+\paragraph{Future directions}\label{future-directions}
 
@@
 
-\hypertarget{acknowledgments}{%
-\subsubsection{Acknowledgments}\label{acknowledgments}}
-
-We thank Owen Liu and XX reviewers for constructive comments on the
+\subsubsection{Acknowledgments}\label{acknowledgments}
+
+We thank Owen Liu and Chris Jordan for constructive comments on the
 manuscript.
@@
 
-\hypertarget{references}{%
-\subsection{References}\label{references}}
-
-\hypertarget{refs}{}
+\subsection{References}\label{references}
+
+\phantomsection\label{refs}
 \begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-agostini2006}{}}%
+\bibitem[\citeproctext]{ref-agostini2006}
 Agostini, V. N., Francis, R. C., Hollowed, A. B., Pierce, S. D., Wilson,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Anderson2024}{}}%
+\bibitem[\citeproctext]{ref-Anderson2024}
 Anderson, S. C., Ward, E. J., English, P. A., Barnett, L. A. K., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-brooks2017}{}}%
+\bibitem[\citeproctext]{ref-brooks2017}
 Brooks, E., M., Kristensen, K., Benthem, A. van, K. J.and Magnusson,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-cao2017}{}}%
+\bibitem[\citeproctext]{ref-cao2017}
 Cao, J., Thorson, J. T., Richards, R. A., \& Chen, Y. (2017).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ciannelli2004}{}}%
+\bibitem[\citeproctext]{ref-ciannelli2004}
 Ciannelli, L., Chan, K.-S., Bailey, K. M., \& Stenseth, N. Chr. (2004).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-elith2009}{}}%
+\bibitem[\citeproctext]{ref-elith2009}
 Elith, J., \& Leathwick, J. R. (2009). {Species distribution models:
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-fredston2021}{}}%
+\bibitem[\citeproctext]{ref-fredston2021}
 Fredston, A., Pinsky, M., Selden, R. L., Szuwalski, C., Thorson, J. T.,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Grandin2024}{}}%
+\bibitem[\citeproctext]{ref-Grandin2024}
 Grandin, C. J., Johnson, K. F., M., E. A., \& M., B. A. (2024).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-guisan2005}{}}%
+\bibitem[\citeproctext]{ref-guisan2005}
 Guisan, A., \& Thuiller, W. (2005). {Predicting species distribution:
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-dharma2022}{}}%
+\bibitem[\citeproctext]{ref-dharma2022}
 Hartig, F. (2022). \emph{{DHARMa: Residual diagnostics for hierarchical
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-hilborn1992}{}}%
+\bibitem[\citeproctext]{ref-hilborn1992}
 Hilborn, R., \& Walters, C. J. (1992). \emph{{Quantitative fisheries
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ices2020}{}}%
+\bibitem[\citeproctext]{ref-ices2020}
 ICES. (2020). {Workshop on unavoidable survey effort reduction
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-johnson2019}{}}%
+\bibitem[\citeproctext]{ref-johnson2019}
 Johnson, K. F., Thorson, J. T., \& Punt, A. E. (2019). {Investigating
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-keller2017}{}}%
+\bibitem[\citeproctext]{ref-keller2017}
 Keller, A. A., Wallace, J. R., \& Methot, R. D. (2017). \emph{{The
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-kristensen2016}{}}%
+\bibitem[\citeproctext]{ref-kristensen2016}
 Kristensen, K., Nielsen, A., Berg, C. W., Skaug, H., \& Bell, B. M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-laman2018}{}}%
+\bibitem[\citeproctext]{ref-laman2018}
 Laman, E. A., Rooper, C. N., Turner, K., Rooney, S., Cooper, D. W., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-leising2024cciea}{}}%
+\bibitem[\citeproctext]{ref-leising2024cciea}
 Leising, A., M., H., N., T., G., W., \& A., H. (2024).
 
-\leavevmode\vadjust pre{\hypertarget{ref-lewison2004}{}}%
+\bibitem[\citeproctext]{ref-lewison2004}
 Lewison, R. L., Freeman, S. A., \& Crowder, L. B. (2004). {Quantifying
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lindgren2015}{}}%
+\bibitem[\citeproctext]{ref-lindgren2015}
 Lindgren, F., \& Rue, H. (2015). {Bayesian spatial modelling
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-lindgren2011}{}}%
+\bibitem[\citeproctext]{ref-lindgren2011}
 Lindgren, F., Rue, H., \& Lindström, J. (2011). {An explicit link
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-melo-merino2020}{}}%
+\bibitem[\citeproctext]{ref-melo-merino2020}
 Melo-Merino, S. M., Reyes-Bonilla, H., \& Lira-Noriega, A. (2020).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-methot2013}{}}%
+\bibitem[\citeproctext]{ref-methot2013}
 Methot, R. D., \& Wetzel, C. R. (2013). {Stock synthesis: A biological
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ono2024}{}}%
+\bibitem[\citeproctext]{ref-ono2024}
 Ono, K., Katara, I., Eliasen, S. K., Broms, C., Campbell, A., dos Santos
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-ono2016space}{}}%
+\bibitem[\citeproctext]{ref-ono2016space}
 Ono, K., Shelton, A. O., Ward, E. J., Thorson, J. T., Feist, B. E., \&
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-phillips2017}{}}%
+\bibitem[\citeproctext]{ref-phillips2017}
 Phillips, N. D., Reid, N., Thys, T., Harrod, C., Payne, N. L., Morgan,
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rcore2024}{}}%
+\bibitem[\citeproctext]{ref-rcore2024}
 R Core Team. (2024). \emph{{R: A language and environment for
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rijnsdorp2018}{}}%
+\bibitem[\citeproctext]{ref-rijnsdorp2018}
 Rijnsdorp, A. D., Bolam, S. G., Garcia, C., Hiddink, J. G., Hintzen, N.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-thorson2015}{}}%
+\bibitem[\citeproctext]{ref-thorson2015}
 Thorson, J. T., Shelton, A. O., Ward, E. J., \& Skaug, H. J. (2015).
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tolimieri2023sea}{}}%
+\bibitem[\citeproctext]{ref-tolimieri2023sea}
 Tolimieri, N., \& Haltuch, M. (2023). {Sea-level index of recruitment
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-tolimieri2018}{}}%
+\bibitem[\citeproctext]{ref-tolimieri2018}
 Tolimieri, N., Haltuch, M. A., Lee, Q., Jacox, M. G., \& Bograd, S. J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-werner1984}{}}%
+\bibitem[\citeproctext]{ref-werner1984}
 Werner, E. E., \& Gilliam, J. F. (1984). {The ontogenetic niche and
@@
 
-\hypertarget{figures}{%
-\subsection{Figures}\label{figures}}
+\subsection{Figures}\label{figures}
 
@@
 
-{\centering \includegraphics[width=3in,height=\textheight]{plots/hake_spatial_anomaly.png}
+\centering{
+
+\includegraphics[width=3in,height=\textheight]{plots/hake_spatial_anomaly.png}
 
@@
 
-\end{figure}
+\end{figure}%
 
@@
 
-{\centering \includegraphics[width=4in,height=\textheight]{plots/sablefish-spatial-anomaly.png}
+\centering{
+
+\includegraphics[width=4in,height=\textheight]{plots/sablefish-spatial-anomaly.png}
 
@@
 
-{\centering \includegraphics[width=6.2in,height=\textheight]{plots/Pacific hake-follow-age-class.png}
+\centering{
+
+\includegraphics[width=6.2in,height=\textheight]{plots/Pacific hake-follow-age-class.png}
 
@@
 
-{\centering \includegraphics[width=6.2in,height=\textheight]{plots/sablefish-follow-age-class.png}
+\centering{
+
+\includegraphics[width=6.2in,height=\textheight]{plots/sablefish-follow-age-class.png}
 
@@
 
-{\centering \includegraphics[width=3.5in,height=\textheight]{plots/glm_coefficients.png}
+\centering{
+
+\includegraphics[width=3.5in,height=\textheight]{plots/glm_coefficients.png}
 
@@
 
-{\centering \includegraphics[width=5in,height=\textheight]{plots/sablefish_spatial_risk.png}
+\centering{
+
+\includegraphics[width=5in,height=\textheight]{plots/sablefish_spatial_risk.png}
 
@@
 
-\hypertarget{supplementary-material}{%
-\section{Supplementary material}\label{supplementary-material}}
+\section{Supplementary material}\label{supplementary-material}
 
@@
 
-{\centering \includegraphics[width=5in,height=\textheight]{plots/spatial_parameters.png}
+\centering{
+
+\includegraphics[width=5in,height=\textheight]{plots/spatial_parameters.png}
 
@@
 
-{\centering \includegraphics[width=6.2in,height=\textheight]{plots/Pacific hake-age-class-year.png}
+\centering{
+
+\includegraphics[width=6.2in,height=\textheight]{plots/Pacific hake-age-class-year.png}
 
@@
 
-{\centering \includegraphics[width=6.2in,height=\textheight]{plots/sablefish-age-class-year.png}
+\centering{
+
+\includegraphics[width=6.2in,height=\textheight]{plots/sablefish-age-class-year.png}
 
@@
 
-{\centering \includegraphics[width=7in,height=\textheight]{plots/glm_coefficients_time.png}
+\centering{
+
+\includegraphics[width=7in,height=\textheight]{plots/glm_coefficients_time.png}
 

### 2025-01-30 01_recursive_model.R
more cleaning up

@@ -203,3 +203,4 @@ combo <- ggpubr::ggarrange( p2, p3, p1,nrow = 1,
                             labels = c('a)','b)','c)'),
                             font.label = list(size=8))
 ggsave(combo, filename="plots/spatial_parameters.png", height = 2, width=5)
+ggsave(combo, filename="plots/final_plots/S1_spatial_parameters.jpg", height = 4, width=5)

### 2025-01-30 06_plots_abundance_pretty.R

@@ -14,7 +14,7 @@ library(ggpubr)
 home_dir = getwd()
 fig_dir = paste0(home_dir,"/plots/")
 
-spp_name <- c("Pacific hake", "sablefish")[2]
+spp_name <- c("Pacific hake", "sablefish")[1]
 
 if(spp_name == "Pacific hake") {
   min_age <- 1 # not many age 0s consistently sampled
@@ -58,7 +58,7 @@ xmin = floor( min(pred_all$lon2))
 
 # background map info ##########################################################
 states <- map_data("state")
-west_coast <- subset(states, region %in% c("california", "oregon", "washington"))
+west_coast <- subset(states, region %in% c("california", "oregon", "washington","nevada","idaho","montana"))
 ################################################################################
 
 # plot function
@@ -91,7 +91,7 @@ dist_maps <- function(dfile,
     ggplot(data=west_coast, aes(x = long, y = lat), 
                  fill = grey(0.9), color = "black") +
     geom_polygon(aes(x = long, y = lat, group=group), 
-               fill = grey(0.9), color = "black") +
+               fill = "wheat", color = "grey70") +
     geom_point(data=dfile , 
                aes(lon2, lat, color=p), size=0.01) +
     # might need to adjust here based on the scale of the data, esp 'trans' term.
@@ -141,6 +141,7 @@ if(spp_name=="Pacific hake"){
   ggarrange(m1,m2,m3,m4, m5, nrow = 5)}
 
 ggsave( paste0(fig_dir , spp_name,"-age-class-year.png"), width = 6.2, height = 8)
+ggsave( paste0(fig_dir ,"final_plots/S_", spp_name,"-age-class-year.jpg"), width = 6.2, height = 8)
 # cohort abundance through time ################################################
 
 # sablefish 
@@ -157,6 +158,7 @@ if(spp_name == "sablefish"){
   ggarrange(y2008, y2010, y2016, y2021,nrow = 4)
 
   ggsave( paste0(fig_dir ,spp_name,"-follow-age-class.png"), width = 6.2, height = 6.5)
+  ggsave( paste0(fig_dir ,"final_plots/04_",spp_name,"-follow-age-class.jpg"), width = 6.2, height = 6.5)
 }
 
 
@@ -175,6 +177,7 @@ if(spp_name == "Pacific hake"){
   ggarrange(y2008, y2010, y2014,y2016, nrow = 4)
   
[…]

### 2025-02-27 paper/paper-word.qmd
reorg

--- 2025-01-29 paper.qmd
@@
 format: 
-  pdf:
-    keep-tex: true
-    template-partials: 
-      - title.tex
-    include-in-header:
-      text: |
-        \usepackage[noblocks]{authblk}
-        \renewcommand*{\Authsep}{, }
-        \renewcommand*{\Authand}{, }
-        \renewcommand*{\Authands}{, }
-        \renewcommand\Affilfont{\small}
+  docx: default
+
 author:
@@
 
-We thank Owen Liu and XX reviewers for constructive comments on the manuscript.
+We thank Owen Liu and Chris Jordan for constructive comments on the manuscript.
 
